---
title: "Ballast water bacteria 16S rRNA analysis"
author: "Sarah Brown"
date: "July 19, 2023"
output: 
  github_document:
    toc: true
    number_sections: true
  theme: bright
  highlight: default
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Packages and Import Data

Here, we load the required packages and then import the .qza files exported from QIIME2 using the qiime2R package. The final object is a phyloseq object called Ballast_physeq.

```{r packages, warning = FALSE, messages = FALSE}

library(ggplot2)
library(vegan)
library(plyr)
library(dplyr)
library(scales)
library(grid)
library(reshape2)
library(phyloseq)
library(picante)
library(tidyr)
library(viridis)
library(qiime2R)
library(DESeq2)
library(patchwork)
library(RColorBrewer)
library(RColorBrewer)
library(microViz)
library(speedyseq)
library(ComplexHeatmap)
library(plotly)
library(seecolor)

set.seed(13745)

#Create a phyloseq object from the .qza files exported from qiime2 using 
#the qiime2R package
Ballast_physeq <- qza_to_phyloseq(
  features="Qiime output/ballast-dada2-table.qza",
  tree="Qiime output/ballast-rooted-tree.qza",
  taxonomy="Qiime output/ballast-taxonomy.qza",
  metadata = "Qiime output/Metadata.tsv"
)
```

# Preparing the Data

Before performing any analyses, we need to examine the phyloseq object and remove chloroplast and mitochondrial sequences:

```{r prep}

#Check the rank names to make sure they are accurate
rank_names(Ballast_physeq)

#Correct output:[1] "Kingdom" "Phylum"  "Class"   "Order"   "Family"  
#                   "Genus"   "Species"

#Check sample variables
sample_variables(Ballast_physeq)

#Remove chloroplast sequences and any contaminant sequences
Ballast_physeq <- subset_taxa(Ballast_physeq, Kingdom != "d__Archaea")
Ballast_physeq <- subset_taxa(Ballast_physeq, Order != "Chloroplast")
Ballast_physeq <- subset_taxa(Ballast_physeq, Family != "Mitochondria")

#Check that contaminant sequences are removed (easiest to save as data frame and search)
taxtabl <- as.data.frame(tax_table(Ballast_physeq))

#At this point, blanks and positive controls should also be removed
Ballast_physeq = subset_samples(Ballast_physeq, Sample_type != "control")
Ballast_physeq = subset_samples(Ballast_physeq, Sample_type != "negative")
```

## Examine the total number of reads and ASV's

This part of the code examines the number of times each ASV occurs in the data set and the number of reads per sample; this will help determine which ASV's and samples need to be filtered out of the data set.

```{r asv, warning = FALSE}

#Create graph showing number of reads per ASV and number of reads per sample
readsumsdf = data.frame(nreads = sort(taxa_sums(Ballast_physeq), TRUE), 
                        sorted = 1:ntaxa(Ballast_physeq), type = "ASVs")
readsumsdf = rbind(readsumsdf, data.frame(nreads = sort(sample_sums(Ballast_physeq), TRUE), sorted = 1:nsamples(Ballast_physeq), type = "Samples"))
title = "Total number of reads"
p = ggplot(readsumsdf, aes(x = sorted, y = nreads)) + geom_bar(stat = "identity")
p + ggtitle(title) + scale_y_log10() + facet_wrap(~type, 1, scales = "free")
```

```{r ASV abundance}

#Create a dataframe that includes only the total number of times that each ASV occurs
readsumsdf_ASVs <- readsumsdf %>% 
    filter(type == "ASVs")  
  
#How many singletons are present?
length(which(readsumsdf_ASVs$nreads <= 0))
#Answer: 11
length(which(readsumsdf_ASVs$nreads == 1))
#Answer: 0

#How many doubletons are present?
length(which(readsumsdf_ASVs$nreads == 2))
#Answer: 29

#Create a bar plot showing the number of times each ASV occurs
ASV_occurance <- ggplot(data = readsumsdf_ASVs) +
    geom_bar(mapping = aes(x=nreads)) + 
    ylim(0, 20) +
    labs(x = "Number of reads", y = "Number of ASV's") +
    theme(text = element_text(size = 18), 
        axis.title = element_text(size = 15),
        panel.spacing = unit(1, "lines"), 
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5), 
        panel.background = element_blank())

print(ASV_occurance)
```

## Create table showing the number of reads per sample

```{r sample reads}

#Sum reads and sort by samples with least to greatest number of reads
samples_reads <- sort(sample_sums(Ballast_physeq))

#Create table 
knitr::kable(samples_reads) %>% 
  kableExtra::kable_styling("striped", 
                            latex_options="scale_down") %>% 
  kableExtra::scroll_box(width = "100%")
```

## Create a rarefaction curve

```{r rarefaction, warning = FALSE, messages = FALSE}

#Rarefaction curve using vegan
#From: https://micca.readthedocs.io/en/latest/phyloseq.html
taxa_are_rows(Ballast_physeq)

mat <- t(otu_table(Ballast_physeq))
class(mat) <- "matrix"
class(mat)

mat <- as(t(otu_table(Ballast_physeq)), "matrix")
class(mat)

raremax <- min(rowSums(mat))

system.time(rarecurve(mat, step = 100, sample = raremax, col = "blue", label = FALSE))
```

## Filter samples

Based on the table above, there are some samples with very few to no reads, so we'll filter these out here. In this case, we keep all samples with more than 1,000 reads.

Removing these samples with low sequencing depths is important because samples with low reads typically are lower quality and have a greater probability of containing contaminant sequences ([Weiss et al., 2017](https://microbiomejournal.biomedcentral.com/articles/10.1186/s40168-017-0237-y)).

```{r filter}

#Remove samples with few reads
Ballast_physeq = subset_samples(Ballast_physeq, 
                            Sample_number != "104")
Ballast_physeq = subset_samples(Ballast_physeq, 
                            Sample_number != "105")
Ballast_physeq = subset_samples(Ballast_physeq, 
                            Sample_number != "107")
Ballast_physeq = subset_samples(Ballast_physeq, 
                            Sample_number != "108")
Ballast_physeq = subset_samples(Ballast_physeq, 
                            Sample_number != "109")
Ballast_physeq = subset_samples(Ballast_physeq, 
                            Sample_number != "106")

#Filter out low abundance ASVs
Ballast_physeq = prune_taxa(taxa_sums(Ballast_physeq) > 2, Ballast_physeq)

```

## Label and order variables

Now we can relabel the variables; if you type in Ballast_physeq you will see that sample_data is the matrix that holds the information that we want to change, so we need to include sample_data in our code here.

```{r label}

#Order factors
sample_data(Ballast_physeq)$Sample_type <- factor(sample_data(Ballast_physeq)$Sample_type, 
               levels = c("port_uptake", "ocean_uptake", "BWT", "BWT_BWE"),
               labels = c("Port uptake", "Ocean uptake", "BWT", "BWT+BWE"))

sample_data(Ballast_physeq)$Sample_number <- factor(sample_data(Ballast_physeq)$Sample_number, 
               levels = c("80", "81", "82", "83", "84", "85", "86", "87", "88", "89", "90", "91", "92", "93", "94", "95", "96", "97", "98", "100", "101", "102", "103", "110", "111", "112", "113", "114", "115"),
               labels = c("80", "81", "82", "83", "84", "85", "86", "87", "88", "89", "90", "91", "92", "93", "94", "95", "96", "97", "98", "100", "101", "102", "103", "110", "111", "112", "113", "114", "115"))

sample_data(Ballast_physeq)$Voyage <- factor(sample_data(Ballast_physeq)$Voyage, 
               levels = c("1", "2"),
               labels = c("1", "2"))
```

## Abundance transformation

We have to transform the count data to account for differences in library size between samples. This is often done using rarefaction, but rarefaction results in the elimination of valid data and can also increase the rate of false positives when testing for ASV's that are differentially abundant in different sample categories.

Additionally, here are some helpful references regarding library normalization strategies:

[Waste Not, Want Not: Why Rarefying Microbiome Data is Inadmissible](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1003531)

[Normalization and microbial differential abundance strategies depend on data characteristics](https://microbiomejournal.biomedcentral.com/articles/10.1186/s40168-017-0237-y)

Below, we'll use relative abundance as our variance-stabilizing method.

```{r abundance}

#Transform to relative abundance
Ballast_physeqRA = transform_sample_counts(Ballast_physeq, function(x){x / sum(x)})
```

Alternatively, you could also choose to log-transform the data:

```{r log transform}

Ballast_physeqlog = transform_sample_counts(Ballast_physeq, function(x) log(1 + x))
```

# Alpha Diversity

Note that while these data are not rarefied, they are normalized. We'll use the Shannon diversity index (based on richness AND evenness; examines how many different taxa are present and how evenly they're distributed within a sample) to analyze alpha diversity between variable types. This means it considers both the number of species and the inequality between species abundances.

## Calculate Shannon Diversity per Sample

Using the estimate_richness function in the phyloseq package to calculate Shannon diversity. The estimate_richness function can also take measures "Chao1" "ACE" "Simpson" and "Fisher".

Note that running this code results in an error message stating that the provided data does not have any singletons. See here for more information:

<https://github.com/benjjneb/dada2/issues/214>

<https://forum.qiime2.org/t/singletons-and-diversity-richness-indices/2971/9>

```{r shannon, warning = FALSE, messages = FALSE}

#Calculating richness - Shannon diversity in a new dataframe
richness <- data.frame(estimate_richness(Ballast_physeqRA, measures = c("Shannon")))
richness <- setNames(cbind(rownames(richness), richness, row.names = NULL), 
                     c("sample-id", "Shannon"))

#Add the sample metadata to the dataframe
s <- data.frame(sample_data(Ballast_physeqRA))
s <- setNames(cbind(rownames(s), s, row.names = NULL), 
              c("sample-id", "Sample_type", "Voyage", "Tank", 
                "Sample_number", "Sample_site", "Sample_date"))

alphadiv <- merge(s, richness, by = "sample-id")

#Order factors
alphadiv$Sample_type <- factor(alphadiv$Sample_type, 
                      levels = c("Port uptake", "Ocean uptake", "BWT", "BWT+BWE"),
                      labels = c("Port uptake", "Ocean uptake", "BWT", "BWT+BWE")) 

#Shows the calculated indices
knitr::kable(head(alphadiv)) %>% 
  kableExtra::kable_styling("striped", latex_options="scale_down") %>% 
  kableExtra::scroll_box(width = "100%")
```

## Create Alpha Diversity Boxplots

### Comparison: sample type

```{r boxplot}

#Plot
ggplot(data=alphadiv, aes(x=Sample_type, y=Shannon), alpha=0.1) + 
  geom_boxplot(aes(fill=Sample_type)) +
  geom_point(position=position_dodge(width=0.75),aes(group=Sample_type)) +
  theme(legend.position="right",
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5), 
        panel.background = element_blank())
```

### Comparison: sample type and voyage

```{r boxplot2}

#Plot
ggplot(data=alphadiv, aes(x=Voyage, y=Shannon), alpha=0.1) + 
  geom_boxplot(aes(fill=Sample_type)) +
  geom_point(position=position_dodge(width=0.75),aes(group=Sample_type)) +
  facet_wrap(~Sample_type, scale="free") +
  theme(legend.position="right",
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5), 
        panel.background = element_blank())
```

## Alpha Diversity statistics

Perform Wilcoxon test.

```{r alpha stats}



```

# Ordination plots

## Find plot color hex codes

```{r colors}

#For points
brewer.pal(4, "Dark2")

#For lines
brewer.pal(3, "Set1")

```

## Create NMDS with the Bray-Curtis dissimilarity matrix

Here, we use the relative abundance transformed data to create an NMDS plot.

Also, pay attention to the printout of the stress here - less than 0.05 means that the two dimensional plot is an excellent representation of the data (i.e., the NMDS ordination is a 'good fit' for the data), \<0.1 and 0.2 are also okay, but \> or = to 0.3 means that the NMDS ordination does not fit the data well.

```{r NMDS Bray}

all.nmds.source.ord <- ordinate(
  physeq = Ballast_physeqRA, 
  method = "NMDS", 
  distance = "bray"
)

#Plot, color coding by Sample_type
all.nmds.sample.type <- plot_ordination(
  physeq = Ballast_physeqRA,
  ordination = all.nmds.source.ord) + 
  scale_colour_manual(values = c("Port uptake" = "#E7298A", "Ocean uptake" = "#7570B3", "BWT" = "#1B9E77", "BWT+BWE" = "#D95F02"), "Sample Type") +
  scale_shape_manual(values = c("1" = 16, "2" = 17), name = "Voyage") +
  geom_point(mapping = aes(colour = factor(Sample_type), shape = factor(Voyage), size = 5)) +
  guides(size=FALSE) +
  guides(shape = guide_legend(override.aes = list(size = 3))) +
  theme(plot.title = element_text(size = 18),
        text = element_text(size = 18), 
        axis.title = element_text(size = 15),
        panel.spacing = unit(1, "lines"), 
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5), 
        panel.background = element_blank(), 
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15),
        legend.justification = c("right", "top")) 

print(all.nmds.sample.type)

```

## Create PCoA with the Bray-Curtis dissimilarity matrix

```{r PCoA Bray}

all.pcoa.source.ord <- ordinate(
  physeq = Ballast_physeqRA, 
  method = "PCoA", 
  distance = "bray"
)

#Plot, color coding by Sample_type
all.pcoa.sample.type <- plot_ordination(
  physeq = Ballast_physeqRA,
  ordination = all.pcoa.source.ord) + 
  scale_colour_manual(values = c("Port uptake" = "#E7298A", "Ocean uptake" = "#7570B3", "BWT" = "#1B9E77", "BWT+BWE" = "#D95F02"), "Sample Type") +
  scale_shape_manual(values = c("1" = 16, "2" = 17), name = "Voyage") +
  geom_point(mapping = aes(colour = factor(Sample_type), shape = factor(Voyage), size = 5)) +
  geom_text(aes(label = Sample_number, colour = factor(Sample_type)), nudge_x=0.05, nudge_y=0.05, check_overlap = TRUE) +
  guides(size=FALSE) +
  guides(shape = guide_legend(override.aes = list(size = 3))) +
  theme(plot.title = element_text(size = 18),
        text = element_text(size = 18), 
        axis.title = element_text(size = 15),
        panel.spacing = unit(1, "lines"), 
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5), 
        panel.background = element_blank(), 
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15),
        legend.justification = c("right", "top")) +
  geom_segment(
  aes(x = -0.45, y = -0.30, xend = 0.30, yend = -0.30),
  arrow = arrow(length = unit(0.03, "npc"))) +
   geom_segment(
  aes(x = -0.08, y = 0.03, xend = 0.29, yend = -0.08),
  arrow = arrow(length = unit(0.03, "npc")), colour = "#E41A1C") +
   geom_segment(
  aes(x = -0.14, y = 0.35, xend = 0.04, yend = 0.34),
  arrow = arrow(length = unit(0.03, "npc"))) +
  geom_curve(
  aes(x = -0.14, y = 0.43, xend = 0.25, yend = 0.21),
  arrow = arrow(length = unit(0.03, "npc")), curvature = -0.3) +
  geom_segment(
  aes(x = 0.06, y = 0.34, xend = 0.29, yend = -0.08),
  arrow = arrow(length = unit(0.03, "npc")), colour = "#E41A1C") 
  

print(all.pcoa.sample.type)

```

### 

# Preliminary taxonomy fix

Some of the taxonomic levels in the phyloseq object have NA's or short values such as "g\_" which are uninformative when graphing bacterial taxonomy at those levels. The tax_fix function from the microViz package fills in these unknown values with useful information from higher taxonomic levels.

This should be run prior to running the following code for generating community composition barcharts, heatmaps, or any other analysis involving the visualization of bacterial community composition.

```{r tax fix}

#First, view the tax_table
knitr::kable(head(tax_table(Ballast_physeq))) %>% 
  kableExtra::kable_styling("striped") %>% 
  kableExtra::scroll_box(width = "100%")

#Now fix the labels
Ballast_physeq <- tax_fix(Ballast_physeq)

#View the relabeled tax_table
knitr::kable(head(tax_table(Ballast_physeq))) %>% 
  kableExtra::kable_styling("striped") %>% 
  kableExtra::scroll_box(width = "100%")
```

# Bacterial community composition bar charts

## Find color hex codes

I've included here a list of hex codes for colors that can be used to identify taxa in plots of bacterial community composition; these hex codes are visualized with the seecolor package.

```{r bcc hex, messages = FALSE}

bcc_hex <- print_color(c("black", "#440154FF", "#450659FF","#460B5EFF","#472D7AFF","#3B518BFF", "#3A548CFF", "#38598CFF", "#365C8DFF", "#34608DFF", "#33638DFF", "#31678EFF", "#306A8EFF","#2E6E8EFF", "#2D718EFF", "#2B748EFF","#2A778EFF", "#297B8EFF","#1F958BFF", "#1F988BFF", "#1E9C89FF", "#1F9F88FF","#1FA287FF", "#21A585FF", "#23A983FF","#25AC82FF", "#29AF7FFF", "#2DB27DFF","#32B67AFF", "#37B878FF", "#3CBC74FF","#57C766FF", "#5EC962FF","#A2DA37FF","#DAE319FF", "#E4E419FF",  "#ECE51BFF", "#F5E61FFF","darkorchid1", "darkorchid2", "darkorchid3","#A21C9AFF", "#A62098FF", "#AB2394FF", "#AE2892FF", "#B22B8FFF", "#B6308BFF", "#BA3388FF", "#BE3885FF", "#C13B82FF", "#C53F7EFF","#C8437BFF", "#CC4678FF", "#CE4B75FF", "#D14E72FF", "#D5536FFF", "#D7566CFF", "#DA5B69FF", "#DD5E66FF", "#E06363FF","#FF6699","#F68D45FF", "#FCA537FF","#F6E726FF", "#F4ED27FF", "#00489C", "#CCCCCC", "#999999", "#A1C299","#300018"), type = "r")

print(bcc_hex)
```

## Family level, individual samples

```{r bcc sample id}

#Examining BCC in each individual sample first
BallastSeqR_family <- Ballast_physeq %>%
  tax_glom(taxrank = "Family") %>%                    #Agglomerate at family level
  transform_sample_counts(function(x) {x/sum(x)}) %>% #Transform to rel. abundance
  psmelt() %>%                                        # Melt to long format
  group_by(Sample, 
           Kingdom, Phylum, Class, Order, Family) %>%
  filter(Abundance > 0.01) %>%                        #Filter 
  arrange(Class)

#Creating a new column that combines both the Class and Family level information
#so that Family level identifiers that are not unique to one Class (such as Ambiguous_taxa) aren't merged into a single large category when graphed
BallastSeqR_family$Class_family <- paste(BallastSeqR_family$Class, BallastSeqR_family$Family, sep="_")

#Creating plot  
BallastSeqR_family_bar <- ggplot(BallastSeqR_family, aes(x = Sample, y = Abundance, fill = Class_family)) + 
  geom_bar(stat="identity", colour = "black", size=0.3, position = "fill") +
  scale_fill_manual(values = c("black", 
                               
                               "#440154FF", "#450659FF","#460B5EFF",
                               
                               "#472D7AFF", 
                               
                               "#3B518BFF", "#3A548CFF", "#38598CFF", "#365C8DFF", 
                               "#34608DFF", "#33638DFF", "#31678EFF", "#306A8EFF",
                               "#2E6E8EFF", "#2D718EFF", "#2B748EFF",
                               "#2A778EFF", "#297B8EFF",
                               
                               "#1F958BFF", "#1F988BFF", "#1E9C89FF", "#1F9F88FF",
                               "#1FA287FF", "#21A585FF", "#23A983FF",
                               "#25AC82FF", "#29AF7FFF", "#2DB27DFF",
                               "#32B67AFF", "#37B878FF", "#3CBC74FF",
                               
                               "#57C766FF", "#5EC962FF",
                               
                               "#A2DA37FF",
                               
                               "#DAE319FF", "#E4E419FF",  "#ECE51BFF", "#F5E61FFF",
                               
                               "darkorchid1", "darkorchid2", "darkorchid3", 
                               
                               "#A21C9AFF", "#A62098FF", "#AB2394FF", "#AE2892FF", "#B22B8FFF", "#B6308BFF", "#BA3388FF", "#BE3885FF", "#C13B82FF", "#C53F7EFF",
                               "#C8437BFF", "#CC4678FF", "#CE4B75FF", "#D14E72FF", "#D5536FFF", "#D7566CFF", "#DA5B69FF", "#DD5E66FF", "#E06363FF",
                               
                               "#FF6699",
                               
                               "#F68D45FF", "#FCA537FF",
                               
                               "#F6E726FF", "#F4ED27FF", 
                               
                               
                               "#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018",
                               
                               "#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018",
                               "#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018",
                               "#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018")) +
  
  guides(fill = guide_legend(reverse = FALSE, keywidth = 1, keyheight = 1)) +
  ylab("Relative Abundance (Family > 1%) \n") + xlab("Sample ID") +
  theme_minimal()+theme(panel.border = element_rect(fill=NA, colour = "black"), panel.grid.minor = element_blank(), 
                        panel.grid.major = element_blank(), axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, size=8, colour="black"), 
                        axis.text.y = element_text(size=8, colour="black"), plot.title = element_text(hjust = 0.5), 
                        axis.ticks.x = element_line(colour="#000000", size=0.1), axis.ticks.y = element_line(colour="#000000", size=0.1),
                        legend.text = element_text(size = 7))  

print(BallastSeqR_family_bar)

#Make interactive plot
ggplotly(BallastSeqR_family_bar)

```

## Family level, grouped by sample type

```{r bcc sample type}

#Examining BCC by sample type
BallastSeqR_family_sample <- Ballast_physeq %>%
  tax_glom(taxrank = "Family") %>%                    #Agglomerate at family level
  transform_sample_counts(function(x) {x/sum(x)}) %>% #Transform to rel. abundance
  psmelt() %>%                                        # Melt to long format
  group_by(Sample_type, 
           Kingdom, Phylum, Class, Order, Family) %>%
  dplyr::summarize(Mean = 
                     mean(Abundance, na.rm=TRUE)) %>% #Calculate average
  filter(Mean > 0.01) %>%                             #Filter 
  arrange(Class)

#Creating a new column that combines both the Class and Family level information
#so that Family level identifiers that are not unique to one Class (such as Ambiguous_taxa) aren't merged into a single large category when graphed
BallastSeqR_family_sample$Class_family <- paste(BallastSeqR_family_sample$Class, BallastSeqR_family_sample$Family, sep="_")

#Creating plot  
BallastSeqR_sample_family_bar <- ggplot(BallastSeqR_family_sample, aes(x = Sample_type, y = Mean, fill = Class_family)) + 
  geom_bar(stat="identity", colour = "black", size=0.3, position = "fill") +
  scale_fill_manual(values = c("black", "#440154FF", "#450659FF","#460B5EFF","#472D7AFF", "#3B518BFF", "#3A548CFF", "#38598CFF", "#365C8DFF","#34608DFF", "#33638DFF", "#31678EFF", "#306A8EFF","#2E6E8EFF", "#2D718EFF", "#2B748EFF","#2A778EFF", "#297B8EFF","#1F958BFF", "#1F988BFF", "#1E9C89FF", "#1F9F88FF",
"#1FA287FF", "#21A585FF", "#23A983FF","#25AC82FF", "#29AF7FFF", "#2DB27DFF","#32B67AFF", "#37B878FF", "#3CBC74FF","#57C766FF", "#5EC962FF","#A2DA37FF",
"#DAE319FF", "#E4E419FF",  "#ECE51BFF", "#F5E61FFF","darkorchid1", "darkorchid2", "darkorchid3", "#A21C9AFF", "#A62098FF", "#AB2394FF", "#AE2892FF", "#B22B8FFF", "#B6308BFF", "#BA3388FF", "#BE3885FF", "#C13B82FF", "#C53F7EFF","#C8437BFF", "#CC4678FF", "#CE4B75FF", "#D14E72FF", "#D5536FFF", "#D7566CFF", "#DA5B69FF", "#DD5E66FF", "#E06363FF","#FF6699","#F68D45FF", "#FCA537FF","#F6E726FF", "#F4ED27FF", "#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018","#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018","#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018","#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018")) +
  
  guides(fill = guide_legend(reverse = FALSE, keywidth = 1, keyheight = 1)) +
  ylab("Relative Abundance (Family > 1%) \n") + xlab("Sample ID") +
  theme_minimal()+theme(panel.border = element_rect(fill=NA, colour = "black"), panel.grid.minor = element_blank(), 
                        panel.grid.major = element_blank(), axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, size=8, colour="black"), 
                        axis.text.y = element_text(size=8, colour="black"), plot.title = element_text(hjust = 0.5), 
                        axis.ticks.x = element_line(colour="#000000", size=0.1), axis.ticks.y = element_line(colour="#000000", size=0.1),
                        legend.text = element_text(size = 7))  

print(BallastSeqR_sample_family_bar)

#Make interactive plot
ggplotly(BallastSeqR_sample_family_bar)

```

## Family level, grouped by sample type and voyage

Creating the same bar plot as above, but with subsets within sample type showing the different voyages, since similar sample types that were collected during different voyages cluster seperately on the NMDS and PCoA plots.

```{r}

#Examining BCC by sample type & voyage
BallastSeqR_family_voyage <- Ballast_physeq %>%
  tax_glom(taxrank = "Family") %>%                    #Agglomerate at family level
  transform_sample_counts(function(x) {x/sum(x)}) %>% #Transform to rel. abundance
  psmelt() %>%                                        # Melt to long format
  group_by(Sample_type, Voyage, 
           Kingdom, Phylum, Class, Order, Family) %>%
  dplyr::summarize(Mean = 
                     mean(Abundance, na.rm=TRUE)) %>% #Calculate average
  filter(Mean > 0.01) %>%                             #Filter 
  arrange(Class)

#Creating a new column that combines both the Class and Family level information
#so that Family level identifiers that are not unique to one Class (such as Ambiguous_taxa) aren't merged into a single large category when graphed
BallastSeqR_family_voyage$Class_family <- paste(BallastSeqR_family_voyage$Class, BallastSeqR_family_voyage$Family, sep="_")

#Creating plot  
BallastSeqR_voyage_family_bar <- ggplot(BallastSeqR_family_voyage, aes(x = Voyage, y = Mean, fill = Class_family)) + 
  geom_bar(stat="identity", colour = "black", size=0.3, position = "fill") +
  scale_fill_manual(values = c("black", "#440154FF", "#450659FF","#460B5EFF","#472D7AFF", "#3B518BFF", "#3A548CFF", "#38598CFF", "#365C8DFF","#34608DFF", "#33638DFF", "#31678EFF", "#306A8EFF","#2E6E8EFF", "#2D718EFF", "#2B748EFF","#2A778EFF", "#297B8EFF","#1F958BFF", "#1F988BFF", "#1E9C89FF", "#1F9F88FF",
"#1FA287FF", "#21A585FF", "#23A983FF","#25AC82FF", "#29AF7FFF", "#2DB27DFF","#32B67AFF", "#37B878FF", "#3CBC74FF","#57C766FF", "#5EC962FF","#A2DA37FF",
"#DAE319FF", "#E4E419FF",  "#ECE51BFF", "#F5E61FFF","darkorchid1", "darkorchid2", "darkorchid3", "#A21C9AFF", "#A62098FF", "#AB2394FF", "#AE2892FF", "#B22B8FFF", "#B6308BFF", "#BA3388FF", "#BE3885FF", "#C13B82FF", "#C53F7EFF","#C8437BFF", "#CC4678FF", "#CE4B75FF", "#D14E72FF", "#D5536FFF", "#D7566CFF", "#DA5B69FF", "#DD5E66FF", "#E06363FF","#FF6699","#F68D45FF", "#FCA537FF","#F6E726FF", "#F4ED27FF", "#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018","#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018","#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018","#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018")) +
  
  guides(fill = guide_legend(reverse = FALSE, keywidth = 1, keyheight = 1)) +
  ylab("Relative Abundance (Family > 1%) \n") + xlab("Voyage") +
  theme_minimal()+theme(panel.border = element_rect(fill=NA, colour = "black"), 
  panel.grid.minor = element_blank(),
  panel.grid.major = element_blank(), 
  axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, size=8, 
                              colour="black"), 
  axis.text.y = element_text(size=8, colour="black"), 
  plot.title = element_text(hjust = 0.5), 
  axis.ticks.x = element_line(colour="#000000", size=0.1), 
  axis.ticks.y = element_line(colour="#000000", size=0.1),
  legend.text = element_text(size = 7)) +
  facet_grid(. ~ Sample_type, margins = TRUE, scale="free")

print(BallastSeqR_voyage_family_bar)

#Make interactive plot
ggplotly(BallastSeqR_voyage_family_bar)
```

# PICRUSt2 analysis

## Load PICRUSt2 and dependent packages

```{r picrust packages, warning=FALSE}

library("ggpicrust2")
library("phyloseq")
library("ALDEx2")
library("SummarizedExperiment")
library("Biobase")
library("devtools")
library("ComplexHeatmap")
library("BiocGenerics")
library("BiocManager")
library("metagenomeSeq")
library("Maaslin2")
library("edgeR")
library("lefser")
library("limma")
library("KEGGREST")
library("DESeq2")
library("aplot")
library("dplyr")
library("ggplot2")
library("grid")
library("MicrobiomeStat")
library("readr")
library("stats")
library("tibble")
library("tidyr")
library("ggprism")
library("cowplot")
library("ggforce")
library("ggplotify")
library("magrittr")
library("utils")
 
         
```

## Load data (KO's)

```{r picrust data}



```

## Load data (MetaCyc)

```{r picrust metacyc data}



```

# Session Info

```{r info}

devtools::session_info()
```
