---
title: "Ballast water bacteria 16S rRNA analysis"
author: "Sarah Brown"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
    number_sections: true
  theme: bright
  highlight: default
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# renv is used to manage package versions for reproducibility.
# Run renv::restore() to install the exact package versions used in this analysis.
```

# Load Packages and Import Data

Here, we load the required packages and then import the .qza files exported from QIIME2 using the qiime2R package. The final object is a phyloseq object called Ballast_physeq.

```{r packages, warning = FALSE, messages = FALSE, error = FALSE}

library(ggplot2)
library(vegan)
library(plyr)
library(dplyr)
library(scales)
library(grid)
library(reshape2)
library(phyloseq)
library(picante)
library(tidyr)
library(viridis)
library(qiime2R)
library(DESeq2)
library(patchwork)
library(RColorBrewer)
library(microViz)
library(speedyseq)
library(ComplexHeatmap)
library(plotly)
library(seecolor)
library(tibble)
library(tidyverse)
library(ggh4x)
library(kableExtra)
library(colorspace)

set.seed(13745)

# Reusable color palette for sample types
sample_type_colors <- c(
  "Port uptake" = "#E7298A",
  "Ocean uptake" = "#7570B3",
  "BWT" = "#1B9E77",
  "BWT+BWE" = "#D95F02"
)

# Reusable theme for clean plots
theme_ballast <- function() {
  theme(
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
    panel.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10, colour = "black"),
    axis.text.y = element_text(size = 10, colour = "black"),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.ticks.x = element_line(colour = "#000000", linewidth = 0.1),
    axis.ticks.y = element_line(colour = "#000000", linewidth = 0.1),
    strip.text = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12)
  )
}

# Voyage shape mapping
voyage_shapes <- c("Voyage 1" = 16, "Voyage 2" = 17)

# Samples to exclude (low read counts)
samples_to_exclude <- c("104", "105", "106", "107", "108", "109")

# Factor levels and labels for sample types
sample_type_levels <- c("port_uptake", "BWT", "ocean_uptake", "BWT_BWE")
sample_type_labels <- c("Port uptake", "BWT", "Ocean uptake", "BWT+BWE")

#Create a phyloseq object from the .qza files exported from qiime2 using 
#the qiime2R package
Ballast_physeq <- qza_to_phyloseq(
  features="Qiime output/ballast-dada2-table.qza",
  tree="Qiime output/ballast-rooted-tree.qza",
  taxonomy="Qiime output/ballast-taxonomy.qza",
  metadata = "Qiime output/Metadata.tsv"
)
```

# Preparing the Data

Before performing any analyses, we need to examine the phyloseq object and remove chloroplast and mitochondrial sequences:

```{r prep}

#Check the rank names to make sure they are accurate
rank_names(Ballast_physeq)

#Correct output:[1] "Kingdom" "Phylum"  "Class"   "Order"   "Family"  
#                   "Genus"   "Species"

#Check sample variables
sample_variables(Ballast_physeq)

#Remove chloroplast sequences and any contaminant sequences
Ballast_physeq <- subset_taxa(Ballast_physeq, Kingdom != "d__Archaea")
Ballast_physeq <- subset_taxa(Ballast_physeq, Order != "Chloroplast")
Ballast_physeq <- subset_taxa(Ballast_physeq, Family != "Mitochondria")

#Check that contaminant sequences are removed (easiest to save as data frame and search)
taxtabl <- as.data.frame(tax_table(Ballast_physeq))
```

## Examine the control samples

Before removing the blanks and controls, we examine how many reads they contain, the identity (taxonomy) of the ASVs detected in the negative controls (`Sample_type == "negative"`), and whether any of those ASVs also occur in the true samples. ASVs that appear in both the negative controls and the samples are potential contaminants.

```{r controls}

#Examine the number of reads in the control samples before removing them
control_reads <- sort(sample_sums(subset_samples(Ballast_physeq, Sample_type %in% c("control", "negative"))))
knitr::kable(control_reads, col.names = "Number of reads") %>%
    kableExtra::kable_styling("striped", latex_options = "scale_down") %>%
    kableExtra::scroll_box(width = "100%")

#Identify the ASVs found in the negative controls
#Subset to the negative controls and keep only ASVs that actually occur in them
neg_controls <- subset_samples(Ballast_physeq, Sample_type == "negative")
neg_controls <- prune_taxa(taxa_sums(neg_controls) > 0, neg_controls)

#Reads for each ASV summed across the true (non-control) samples
true_samples <- subset_samples(Ballast_physeq, !(Sample_type %in% c("control", "negative")))
sample_taxa_sums <- taxa_sums(true_samples)

#Build a table of the negative-control ASV identities, their read counts in the
#negative controls, and whether/how much they also occur in the true samples
neg_control_taxa <- as.data.frame(tax_table(neg_controls))
neg_control_taxa$Reads_in_negatives <- as.integer(taxa_sums(neg_controls))
neg_control_taxa$Reads_in_samples <- as.integer(sample_taxa_sums[rownames(neg_control_taxa)])
neg_control_taxa$Present_in_samples <- neg_control_taxa$Reads_in_samples > 0

#Order by abundance in the negative controls
neg_control_taxa <- neg_control_taxa[order(-neg_control_taxa$Reads_in_negatives), ]

knitr::kable(neg_control_taxa, row.names = FALSE) %>%
    kableExtra::kable_styling("striped", latex_options = "scale_down") %>%
    kableExtra::scroll_box(width = "100%")

#How many of the negative-control ASVs were also detected in the samples?
cat(sprintf("%d of %d ASVs detected in the negative controls were also present in the samples.\n",
            sum(neg_control_taxa$Present_in_samples), nrow(neg_control_taxa)))

#At this point, blanks and positive controls should also be removed
Ballast_physeq = subset_samples(Ballast_physeq, Sample_type != "control")
Ballast_physeq = subset_samples(Ballast_physeq, Sample_type != "negative")
```

## Examine the total number of reads and ASV's

This part of the code examines the number of times each ASV occurs in the data set and the number of reads per sample; this will help determine which ASV's and samples need to be filtered out of the data set.

```{r asv, warning = FALSE}

#Create graph showing number of reads per ASV and number of reads per sample
readsumsdf = data.frame(nreads = sort(taxa_sums(Ballast_physeq), TRUE), 
                        sorted = 1:ntaxa(Ballast_physeq), type = "ASVs")
readsumsdf = rbind(readsumsdf, data.frame(nreads = sort(sample_sums(Ballast_physeq), TRUE), sorted = 1:nsamples(Ballast_physeq), type = "Samples"))
title = "Total number of reads"
p = ggplot(readsumsdf, aes(x = sorted, y = nreads)) + geom_bar(stat = "identity")
p + ggtitle(title) + scale_y_log10() + facet_wrap(~type, 1, scales = "free")
```

```{r ASV abundance}

#Create a dataframe that includes only the total number of times that each ASV occurs
readsumsdf_ASVs <- readsumsdf %>% 
    filter(type == "ASVs")  
  
#How many singletons are present?
length(which(readsumsdf_ASVs$nreads <= 0))
#Answer: 11
length(which(readsumsdf_ASVs$nreads == 1))
#Answer: 0

#How many doubletons are present?
length(which(readsumsdf_ASVs$nreads == 2))
#Answer: 29

#Create a bar plot showing the number of times each ASV occurs
ASV_occurance <- ggplot(data = readsumsdf_ASVs) +
    geom_bar(mapping = aes(x=nreads)) + 
    ylim(0, 20) +
    labs(x = "Number of reads", y = "Number of ASV's") +
    theme(text = element_text(size = 18), 
        axis.title = element_text(size = 15),
        panel.spacing = unit(1, "lines"), 
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5), 
        panel.background = element_blank())

print(ASV_occurance)
```

## Create table showing the number of reads per sample

```{r sample reads}

#Sum reads and sort by samples with least to greatest number of reads
samples_reads <- sort(sample_sums(Ballast_physeq))

#Create table 
knitr::kable(samples_reads) %>% 
    kableExtra::kable_styling("striped", 
                            latex_options="scale_down") %>% 
    kableExtra::scroll_box(width = "100%") 
 
 
```

## Create a rarefaction curve

```{r rarefaction, warning = FALSE, messages = FALSE}

#Rarefaction curve using vegan
#From: https://micca.readthedocs.io/en/latest/phyloseq.html
taxa_are_rows(Ballast_physeq)

mat <- t(otu_table(Ballast_physeq))
class(mat) <- "matrix"
class(mat)

mat <- as(t(otu_table(Ballast_physeq)), "matrix")
class(mat)

raremax <- min(rowSums(mat))

system.time(rarecurve(mat, step = 100, sample = raremax, col = "blue", label = FALSE))
```

## Filter samples

Based on the table above, there are some samples with very few to no reads, so we'll filter these out here. In this case, we keep all samples with more than 1,000 reads.

Removing these samples with low sequencing depths is important because samples with low reads typically are lower quality and have a greater probability of containing contaminant sequences ([Weiss et al., 2017](https://microbiomejournal.biomedcentral.com/articles/10.1186/s40168-017-0237-y)).

```{r filter}

#Remove samples with few reads
Ballast_physeq <- subset_samples(Ballast_physeq,
                                 !(Sample_number %in% samples_to_exclude))

#Filter out low abundance ASVs
Ballast_physeq = prune_taxa(taxa_sums(Ballast_physeq) > 2, Ballast_physeq)

```

## Re-check negative-control ASVs after filtering

Now that the low-read samples and low-abundance ASVs have been removed, we check whether the ASVs that were detected in the negative controls are still present in the retained samples. ASVs that dropped out here were only supported by the samples/reads that were filtered out.

```{r neg control after filter}

#taxa_sums over the filtered object; ASVs that were pruned out are treated as 0 reads
filtered_taxa_sums <- taxa_sums(Ballast_physeq)

#Add post-filtering read counts to the negative-control ASV table built earlier
neg_control_after_filter <- neg_control_taxa
neg_control_after_filter$Reads_after_filtering <-
  as.integer(filtered_taxa_sums[rownames(neg_control_after_filter)])
neg_control_after_filter$Reads_after_filtering[is.na(neg_control_after_filter$Reads_after_filtering)] <- 0L
neg_control_after_filter$Present_after_filtering <- neg_control_after_filter$Reads_after_filtering > 0

knitr::kable(neg_control_after_filter, row.names = FALSE) %>%
    kableExtra::kable_styling("striped", latex_options = "scale_down") %>%
    kableExtra::scroll_box(width = "100%")

#Of the negative-control ASVs that overlapped with the samples, how many remain?
n_overlap <- sum(neg_control_after_filter$Present_in_samples)
n_overlap_retained <- sum(neg_control_after_filter$Present_in_samples &
                          neg_control_after_filter$Present_after_filtering)
cat(sprintf("Of the %d negative-control ASVs that were present in the samples, %d were still present after low-read samples and low-abundance ASVs were removed.\n",
            n_overlap, n_overlap_retained))
```

## Label and order variables

Now we can relabel the variables; if you type in Ballast_physeq you will see that sample_data is the matrix that holds the information that we want to change, so we need to include sample_data in our code here.

```{r label}

#Order factors
sample_data(Ballast_physeq)$Sample_type <- factor(
  sample_data(Ballast_physeq)$Sample_type,
  levels = sample_type_levels,
  labels = sample_type_labels)

sample_numbers <- c("80", "81", "82", "83", "84", "85", "86", "87", "88",
                    "89", "90", "91", "92", "93", "94", "95", "96", "97",
                    "98", "100", "101", "102", "103", "110", "111", "112",
                    "113", "114", "115")
sample_data(Ballast_physeq)$Sample_number <- factor(
  sample_data(Ballast_physeq)$Sample_number,
  levels = sample_numbers)

sample_data(Ballast_physeq)$Voyage <- factor(
  sample_data(Ballast_physeq)$Voyage,
  levels = c("1", "2"),
  labels = c("Voyage 1", "Voyage 2"))
```

## Group rarefaction curves by Sample_type

Comparing the sequencing depth and species richness between different sample types. Using the ggrare function in the phyloseq.extended package which wraps vegan's rarecurve function so that it can be easily used with phyloseq objects.

Note: I load the phyloseq-extended package here instead of above in the 'Load packages' section because it seems to create conflicts with the phyloseq package and results in several warning messages.

```{r  rarefaction 2, warning = FALSE, messages = FALSE}

library(phyloseq.extended)

p <- phyloseq.extended:::ggrare(Ballast_physeq, step = 1000, color = "Sample_type", label = "Sample_number", se = FALSE)
 
  
p <- p + facet_wrap(~Sample_type, scale="free") +
  scale_colour_manual(values = sample_type_colors, "Sample Type") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none") +
  labs(x = "Number of Reads")


print(p)

#Save as svg
ggsave(filename="R output/rarecurve_sampletype.svg", plot=p, width=8, height=6, device=svg)
  
```

## Abundance transformation

We have to transform the count data to account for differences in library size between samples. This is often done using rarefaction, but rarefaction results in the elimination of valid data and can also increase the rate of false positives when testing for ASV's that are differentially abundant in different sample categories.

Additionally, here are some helpful references regarding library normalization strategies:

[Waste Not, Want Not: Why Rarefying Microbiome Data is Inadmissible](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1003531)

[Normalization and microbial differential abundance strategies depend on data characteristics](https://microbiomejournal.biomedcentral.com/articles/10.1186/s40168-017-0237-y)

Below, we'll use relative abundance as our variance-stabilizing method.

```{r abundance}

#Transform to relative abundance
Ballast_physeqRA = transform_sample_counts(Ballast_physeq, function(x){x / sum(x)})
```

Alternatively, you could also choose to log-transform the data:

```{r log transform}

Ballast_physeqlog = transform_sample_counts(Ballast_physeq, function(x) log(1 + x))
```

# Alpha Diversity

Note that while these data are not rarefied, they are normalized. We'll use the Shannon diversity index (based on richness AND evenness; examines how many different taxa are present and how evenly they're distributed within a sample) to analyze alpha diversity between variable types. This means it considers both the number of species and the inequality between species abundances.

## Calculate Shannon Diversity per Sample

Using the estimate_richness function in the phyloseq package to calculate Shannon diversity. The estimate_richness function can also take measures "Chao1" "ACE" "Simpson" and "Fisher".

Note that running this code results in an error message stating that the provided data does not have any singletons. See here for more information:

<https://github.com/benjjneb/dada2/issues/214>

<https://forum.qiime2.org/t/singletons-and-diversity-richness-indices/2971/9>

```{r shannon, warning = FALSE, messages = FALSE}

#Change the - in the sample names to _ to prevent the estimate_richness function from changing the - to a . during the calculation of Observed richness
sample_names(Ballast_physeq) <- paste(gsub("-", "_", sample_names(Ballast_physeq)), sep = "_")

#Calculating richness - Shannon diversity and number of observations (Observed) in a new dataframe
richness <- data.frame(estimate_richness(Ballast_physeq, measures = c("Shannon", "Observed")))

richness <- setNames(cbind(rownames(richness), richness, row.names = NULL), 
                     c("sample-id", "Observed", "Shannon"))

#Add the sample metadata to the dataframe
s <- data.frame(sample_data(Ballast_physeq))
s <- setNames(cbind(rownames(s), s, row.names = NULL), 
              c("sample-id", "Sample_type", "Voyage", "Tank", 
                "Sample_number", "Sample_site", "Sample_date"))

alphadiv <- merge(s, richness, by = "sample-id")

#Order factors
alphadiv$Sample_type <- factor(alphadiv$Sample_type, 
                      levels = c("Port uptake", "BWT", "Ocean uptake", "BWT+BWE"),
                      labels = c("Port uptake", "BWT", "Ocean uptake", "BWT+BWE")) 

#Shows the calculated indices
knitr::kable(head(alphadiv)) %>% 
  kableExtra::kable_styling("striped", latex_options="scale_down") %>% 
  kableExtra::scroll_box(width = "100%")
```

## Create Alpha Diversity Boxplots

### Shannon diversity: sample type

```{r boxplot}

#Plot
ggplot(data=alphadiv, aes(x=Sample_type, y=Shannon), alpha=0.1) + 
  geom_boxplot(aes(fill=Sample_type)) +
  scale_fill_manual(values = sample_type_colors, "Sample Type") +
  geom_point(position=position_dodge(width=0.75),aes(group=Sample_type)) +
  scale_y_continuous(limits = c(0, 4.5)) +
  theme(legend.position="right",
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5), 
        panel.background = element_blank(), 
        axis.title.x = element_blank())
```

### Shannon diversity: sample type and voyage

```{r boxplot2}

#Plot
ggplot(data=alphadiv, aes(x=Voyage, y=Shannon), alpha=0.1) + 
  geom_boxplot(aes(fill=Sample_type)) +
  scale_fill_manual(values = sample_type_colors, "Sample Type") +
  geom_point(position=position_dodge(width=0.75),aes(group=Sample_type)) +
  facet_wrap(~Sample_type, scale="free") +
  theme(legend.position="right",
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5), 
        panel.background = element_blank(),
        axis.title.x = element_blank())
```

### Shannon diversity: voyage and sample type

```{r boxplot3}

#Plot
ggplot(data=alphadiv, aes(x=Sample_type, y=Shannon), alpha=0.1) + 
  geom_boxplot(aes(fill=Sample_type)) +
  scale_fill_manual(values = sample_type_colors, "Sample Type") +
  geom_point(position=position_dodge(width=0.75),aes(group=Voyage)) +
  facet_wrap(~Voyage, scale="free") +
  scale_y_continuous(limits = c(0, 4.5)) +
  theme(legend.position="right",
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5), 
        panel.background = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1))
```

### Richness: sample type

```{r boxplot richness 1}

#Plot
ggplot(data=alphadiv, aes(x=Sample_type, y=Observed), alpha=0.1) + 
  geom_boxplot(aes(fill=Sample_type)) +
  scale_fill_manual(values = sample_type_colors, "Sample Type") +
  geom_point(position=position_dodge(width=0.75),aes(group=Sample_type)) +
  scale_y_continuous(limits = c(0, 100)) +
  theme(legend.position="right",
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5), 
        panel.background = element_blank(), 
        axis.title.x = element_blank())
```

### Richness: voyage and sample type

```{r boxplot richness 2}

#Plot
ggplot(data=alphadiv, aes(x=Sample_type, y=Observed), alpha=0.1) + 
  geom_boxplot(aes(fill=Sample_type)) +
  scale_fill_manual(values = sample_type_colors, "Sample Type") +
  geom_point(position=position_dodge(width=0.75),aes(group=Voyage)) +
  facet_wrap(~Voyage, scale="free") +
  scale_y_continuous(limits = c(0, 100)) +
  theme(legend.position="right",
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5), 
        panel.background = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1))
```

### Reads: voyage and sample type

```{r boxplot reads}

#Sum reads and sort by samples with least to greatest number of reads
samples_reads <- sort(sample_sums(Ballast_physeq))

samples_reads_df <- data.frame(sample_sums(Ballast_physeq))

#Make column out of the row names
samples_reads_df <- tibble::rownames_to_column(samples_reads_df, "sample-id")

#Add the sample metadata to the dataframe
s <- data.frame(sample_data(Ballast_physeq))
s <- setNames(cbind(rownames(s), s, row.names = NULL), 
              c("sample-id", "Sample_type", "Voyage", "Tank", 
                "Sample_number", "Sample_site", "Sample_date"))

reads_df <- merge(s, samples_reads_df, by = "sample-id")

#Order factors
reads_df$Sample_type <- factor(reads_df$Sample_type, 
                      levels = c("Port uptake", "BWT", "Ocean uptake", "BWT+BWE"),
                      labels = c("Port uptake", "BWT", "Ocean uptake", "BWT+BWE"))


#Plot
reads_box <- ggplot(data=reads_df, aes(x=Sample_type, y=sample_sums.Ballast_physeq.), 
       alpha=0.1) + 
  labs(y = "Reads") +
  geom_boxplot(aes(fill=Sample_type)) +
  scale_fill_manual(values = sample_type_colors, "Sample Type") +
  geom_point(position=position_dodge(width=0.75),aes(group=Voyage)) +
  facet_wrap(~Voyage, scale="free") +
  #scale_y_continuous(limits = c(0, 100)) +
  theme(legend.position="right",
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5), 
        panel.background = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1))
```

### Combine plots for Fig. 2

```{r fig 1 combo}

#Observed without x-axis labels
richness_box <- ggplot(data=alphadiv, aes(x=Sample_type, y=Observed), alpha=0.1) + 
  geom_boxplot(aes(fill=Sample_type)) +
  scale_fill_manual(values = sample_type_colors, "Sample Type") +
  geom_point(position=position_dodge(width=0.75),aes(group=Voyage)) +
  facet_wrap(~Voyage, scale="free") +
  scale_y_continuous(limits = c(0, 100)) +
  theme(legend.position="right",
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5), 
        panel.background = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

#Shannon without x-axis labels
shannon_box <- ggplot(data=alphadiv, aes(x=Sample_type, y=Shannon), alpha=0.1) + 
  geom_boxplot(aes(fill=Sample_type)) +
  scale_fill_manual(values = sample_type_colors, "Sample Type") +
  geom_point(position=position_dodge(width=0.75),aes(group=Voyage)) +
  facet_wrap(~Voyage, scale="free") +
  scale_y_continuous(limits = c(0, 4.5)) +
  theme(legend.position="right",
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5), 
        panel.background = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks = element_blank())

#Combine with the sample coverage plot using patchwork
boxplot_combo <- richness_box / shannon_box / reads_box

#Save as an png file
ggsave(filename="R output/Fig2_combined_boxplot.png", 
       plot=boxplot_combo, width=5, height=8, device=png)
```

## Alpha Diversity statistics

Perform Wilcoxon test.

Use section 7.2 <https://microbiome.github.io/course_2021_radboud/alpha-diversity.html>

Some information from the tutorial linked above: "To further investigate if patient status could explain the variation of Shannon index, let's do a Wilcoxon test. This is a non-parametric test that doesn't make specific assumptions about the distribution, unlike popular parametric tests, such as the t test, which assumes normally distributed observations. Wilcoxon test can be used to estimate whether the differences between two groups is statistically significant."

### Subset alpha diversity dataframe

```{r alpha stats}

#Subset the alphadiversity dataframe by voyage
v1.alpha <- subset(alphadiv, Voyage == "Voyage 1")
v2.alpha <- subset(alphadiv, Voyage == "Voyage 2")

```

### Shannon diversity: sample type

Voyages are grouped together and differences in sample type are tested using a pairwise Wilcoxon test.

```{r wilcoxon shannon sample type}

sample.shan.wil <- pairwise.wilcox.test(alphadiv$Shannon, alphadiv$Sample_type, p.adjust.method="fdr", paired=FALSE) 
sample.shan.wil
```

```{r wilcoxon shannon sample type table}

#Convert output to dataframe
sample.shan.wil <- as.data.frame(sample.shan.wil$p.value)

#Create table of p-values
knitr::kable(sample.shan.wil) %>% 
  kableExtra::kable_styling("striped", 
                            latex_options="scale_down") %>% 
  kableExtra::scroll_box(width = "100%")
```

### Shannon diversity: voyage and sample type

Testing for differences in sample types in the two voyages separately using a pairwise Wilcoxon test.

```{r wilcoxon shannon voyage sample type}

#Voyage 1
v1.sample.shan.wil <- pairwise.wilcox.test(v1.alpha$Shannon, v1.alpha$Sample_type, p.adjust.method="fdr", paired=FALSE) 
v1.sample.shan.wil

#Voyage 2
v2.sample.shan.wil <- pairwise.wilcox.test(v2.alpha$Shannon, v2.alpha$Sample_type, p.adjust.method="fdr", paired=FALSE) 
v2.sample.shan.wil
```

```{r wilcoxon shannon voyage sample type table v1}

#Convert output to dataframe
v1.sample.shan.wil.df <- as.data.frame(v1.sample.shan.wil$p.value)

#Create table of p-values
knitr::kable(v1.sample.shan.wil.df) %>% 
  kableExtra::kable_styling("striped", 
                            latex_options="scale_down") %>% 
  kableExtra::scroll_box(width = "100%")

```

```{r wilcoxon shannon voyage sample type table v2}

#Convert output to dataframe
v2.sample.shan.wil.df <- as.data.frame(v2.sample.shan.wil$p.value)

#Create table of p-values
knitr::kable(v2.sample.shan.wil.df) %>% 
  kableExtra::kable_styling("striped", 
                            latex_options="scale_down") %>% 
  kableExtra::scroll_box(width = "100%")
```

### Richness: sample type

Voyages are grouped together and differences in sample type are tested using a pairwise Wilcoxon test.

```{r wilcoxon richness sample type}

sample.richness.wil <- pairwise.wilcox.test(alphadiv$Observed, alphadiv$Sample_type, p.adjust.method="fdr", paired=FALSE) 
sample.richness.wil
```

```{r wilcoxon richness sample type table}

#Convert output to dataframe
sample.richness.wil <- as.data.frame(sample.richness.wil$p.value)

#Create table of p-values
knitr::kable(sample.richness.wil) %>% 
  kableExtra::kable_styling("striped", 
                            latex_options="scale_down") %>% 
  kableExtra::scroll_box(width = "100%")
```

### Richness: voyage and sample type

Testing for differences in sample types in the two voyages separately using a pairwise Wilcoxon test.

```{r wilcoxon richness voyage sample type}

#Voyage 1
v1.sample.richness.wil <- pairwise.wilcox.test(v1.alpha$Observed, v1.alpha$Sample_type, p.adjust.method="fdr", paired=FALSE) 
v1.sample.richness.wil

#Voyage 2
v2.sample.richness.wil <- pairwise.wilcox.test(v2.alpha$Observed, v2.alpha$Sample_type, p.adjust.method="fdr", paired=FALSE) 
v2.sample.richness.wil
```

```{r wilcoxon richness voyage sample type table v1}

#Convert output to dataframe
v1.sample.richness.wil.df <- as.data.frame(v1.sample.richness.wil$p.value)

#Create table of p-values
knitr::kable(v1.sample.richness.wil.df) %>% 
  kableExtra::kable_styling("striped", 
                            latex_options="scale_down") %>% 
  kableExtra::scroll_box(width = "100%")

```

```{r wilcoxon richness voyage sample type table v2}

#Convert output to dataframe
v2.sample.richness.wil.df <- as.data.frame(v2.sample.richness.wil$p.value)

#Create table of p-values
knitr::kable(v2.sample.richness.wil.df) %>% 
  kableExtra::kable_styling("striped", 
                            latex_options="scale_down") %>% 
  kableExtra::scroll_box(width = "100%")

```

# Ordination plots

## Find plot color hex codes

```{r colors}

#For points
brewer.pal(4, "Dark2")

#For lines
brewer.pal(3, "Set1")

```

## Create NMDS with the Bray-Curtis dissimilarity matrix

Here, we use the relative abundance transformed data to create an NMDS plot.

Also, pay attention to the printout of the stress here - less than 0.05 means that the two dimensional plot is an excellent representation of the data (i.e., the NMDS ordination is a 'good fit' for the data), \<0.1 and 0.2 are also okay, but \> or = to 0.3 means that the NMDS ordination does not fit the data well.

```{r NMDS Bray}

all.nmds.source.ord <- ordinate(
  physeq = Ballast_physeqRA, 
  method = "NMDS", 
  distance = "bray"
)

#Plot, color coding by Sample_type
all.nmds.sample.type <- plot_ordination(
  physeq = Ballast_physeqRA,
  ordination = all.nmds.source.ord) + 
  scale_colour_manual(values = sample_type_colors, "Sample Type") +
  scale_shape_manual(values = voyage_shapes, name = "Voyage") +
  geom_point(mapping = aes(colour = factor(Sample_type), shape = factor(Voyage), size = 5)) +
  guides(size = "none") +
  guides(shape = guide_legend(override.aes = list(size = 3))) +
  theme(plot.title = element_text(size = 18),
        text = element_text(size = 18), 
        axis.title = element_text(size = 15),
        panel.spacing = unit(1, "lines"), 
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5), 
        panel.background = element_blank(), 
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15),
        legend.justification = c("right", "top")) 

print(all.nmds.sample.type)

```

## Create PCoA with the Bray-Curtis dissimilarity matrix

### PCoA with labels and arrows

```{r PCoA Bray labels}

all.pcoa.source.ord <- ordinate(
  physeq = Ballast_physeqRA, 
  method = "PCoA", 
  distance = "bray"
)

#Plot, color coding by Sample_type
all.pcoa.sample.type <- plot_ordination(
  physeq = Ballast_physeqRA,
  ordination = all.pcoa.source.ord) + 
  scale_colour_manual(values = sample_type_colors, "Sample Type") +
  scale_shape_manual(values = voyage_shapes, name = "Voyage") +
  geom_point(mapping = aes(colour = factor(Sample_type), shape = factor(Voyage), size = 5)) +
  geom_text(aes(label = Sample_number, colour = factor(Sample_type)), nudge_x=0.05, nudge_y=0.05, check_overlap = TRUE) +
  guides(size = "none") +
  guides(shape = guide_legend(override.aes = list(size = 3))) +
  theme(plot.title = element_text(size = 18),
        text = element_text(size = 18), 
        axis.title = element_text(size = 15),
        panel.spacing = unit(1, "lines"), 
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5), 
        panel.background = element_blank(), 
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15),
        legend.justification = c("right", "top")) +
  geom_segment(
  aes(x = -0.45, y = -0.30, xend = 0.30, yend = -0.30),
  arrow = arrow(length = unit(0.03, "npc"))) +
   geom_segment(
  aes(x = -0.08, y = 0.03, xend = 0.29, yend = -0.08),
  arrow = arrow(length = unit(0.03, "npc")), colour = "#E41A1C") +
   geom_segment(
  aes(x = -0.14, y = 0.35, xend = 0.04, yend = 0.34),
  arrow = arrow(length = unit(0.03, "npc"))) +
  geom_curve(
  aes(x = -0.14, y = 0.43, xend = 0.25, yend = 0.21),
  arrow = arrow(length = unit(0.03, "npc")), curvature = -0.3) +
  geom_segment(
  aes(x = 0.06, y = 0.34, xend = 0.29, yend = -0.08),
  arrow = arrow(length = unit(0.03, "npc")), colour = "#E41A1C") 
  

print(all.pcoa.sample.type)

```

### PCoA without labels and arrows

```{r PCoA Bray no labels}

all.pcoa.source.ord <- ordinate(
  physeq = Ballast_physeqRA, 
  method = "PCoA", 
  distance = "bray"
)

#Plot, color coding by Sample_type
all.pcoa.sample.type.no.label <- plot_ordination(
  physeq = Ballast_physeqRA,
  ordination = all.pcoa.source.ord) + 
  scale_colour_manual(values = sample_type_colors, "Sample Type") +
  scale_shape_manual(values = voyage_shapes, name = "Voyage") +
  geom_point(mapping = aes(colour = factor(Sample_type), shape = factor(Voyage), size = 5)) +
  guides(size = "none") +
  guides(shape = guide_legend(override.aes = list(size = 3))) +
  theme(plot.title = element_text(size = 18),
        text = element_text(size = 18), 
        axis.title = element_text(size = 15),
        panel.spacing = unit(1, "lines"), 
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5), 
        panel.background = element_blank(), 
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15),
        legend.justification = c("right", "top"))  
  

print(all.pcoa.sample.type.no.label)
```

# PERMANOVA Analysis

Using the relative abundance-transformed phyloseq object here (Ballast_physeqRA). The adonis PERMANOVA function requires the input to be an abundance table; the output of this function will tell you if environmental factors are influencing differences in bacterial community composition. Useful sources with information on how to interpret PERMANOVA results:

<https://forum.qiime2.org/t/adonis-vs-anosim-vs-permanova/9744>

<https://forum.qiime2.org/t/adonis-correct-interpretation-of-interaction-among-factors/10327>

<https://sites.google.com/site/mb3gustame/hypothesis-tests/manova/npmanova>

## PERMANOVA analysis

```{r Bray PERMANOVA}

#Run PERMANOVA with Bray-Curtis dissimilarity
permanova_bray <- phyloseq::distance(Ballast_physeqRA, method = "bray")
permanova_bray_df <- data.frame(sample_data(Ballast_physeqRA))
all <- adonis2(permanova_bray ~ Voyage*Sample_type, data = permanova_bray_df)
all


```

```{r PERMANOVA table}

#Convert output to dataframe
all.pvalue <- as.data.frame(all)

#Create table of p-values
knitr::kable(all.pvalue) %>% 
  kableExtra::kable_styling("striped", 
                            latex_options="scale_down") %>% 
  kableExtra::scroll_box(width = "100%")

```

## Pairwise PERMANOVA

Examine whether BWT from different tank was statistically the same

```{r pairwise PERMANOVA}

library(pairwiseAdonis)

#Run pairwise PERMANOVA with bray-curtis dissimilarity
Ballast_physeq_df <- data.frame(sample_data(Ballast_physeq))

Ballast_physeq_pairwise <- phyloseq::distance(Ballast_physeq, method = "bray")

Ballast.physeq.pairwise <- pairwise.adonis(Ballast_physeq_pairwise, Ballast_physeq_df$Sample_type, p.adjust.m='BH')

Ballast.physeq.pairwise
```

# Preliminary taxonomy fix

Some of the taxonomic levels in the phyloseq object have NA's or short values such as "g\_" which are uninformative when graphing bacterial taxonomy at those levels. The tax_fix function from the microViz package fills in these unknown values with useful information from higher taxonomic levels.

This should be run prior to running the following code for generating community composition barcharts, heatmaps, or any other analysis involving the visualization of bacterial community composition.

```{r tax fix}

#First, view the tax_table
knitr::kable(head(tax_table(Ballast_physeq))) %>% 
  kableExtra::kable_styling("striped") %>% 
  kableExtra::scroll_box(width = "100%")

#Now fix the labels
Ballast_physeq <- tax_fix(Ballast_physeq)

#View the relabeled tax_table
knitr::kable(head(tax_table(Ballast_physeq))) %>% 
  kableExtra::kable_styling("striped") %>% 
  kableExtra::scroll_box(width = "100%")
```

# Bacterial community composition bar charts

## Find color hex codes

I've included here a list of hex codes for colors that can be used to identify taxa in plots of bacterial community composition; these hex codes are visualized with the seecolor package.

```{r bcc hex, messages = FALSE}

bcc_hex <- c("black", "#440154FF", "#450659FF","#460B5EFF","#472D7AFF","#3B518BFF", "#3A548CFF", "#38598CFF", "#365C8DFF", "#34608DFF", "#33638DFF", "#31678EFF", "#306A8EFF","#2E6E8EFF", "#2D718EFF", "#2B748EFF","#2A778EFF", "#297B8EFF","#1F958BFF", "#1F988BFF", "#1E9C89FF", "#1F9F88FF","#1FA287FF", "#21A585FF", "#23A983FF","#25AC82FF", "#29AF7FFF", "#2DB27DFF","#32B67AFF", "#37B878FF", "#3CBC74FF","#57C766FF", "#5EC962FF","#A2DA37FF","#DAE319FF", "#E4E419FF",  "#ECE51BFF", "#F5E61FFF","darkorchid1", "darkorchid2", "darkorchid3","#A21C9AFF", "#A62098FF", "#AB2394FF", "#AE2892FF", "#B22B8FFF", "#B6308BFF", "#BA3388FF", "#BE3885FF", "#C13B82FF", "#C53F7EFF","#C8437BFF", "#CC4678FF", "#CE4B75FF", "#D14E72FF", "#D5536FFF", "#D7566CFF", "#DA5B69FF", "#DD5E66FF", "#E06363FF","#FF6699","#F68D45FF", "#FCA537FF","#F6E726FF", "#F4ED27FF", "#00489C", "#CCCCCC", "#999999", "#A1C299","#300018")

print_color(bcc_hex, type = "r")
```

## Family level, individual samples

```{r bcc family sample id}

BallastSeqR_family <- Ballast_physeq %>%
  tax_glom(taxrank = "Family") %>%
  transform_sample_counts(function(x) {x/sum(x)}) %>%
  psmelt() %>%
  group_by(Sample, Kingdom, Phylum, Class, Order, Family) %>%
  filter(Abundance > 0.01) %>%
  arrange(Class)

BallastSeqR_family$Class_family <- paste(BallastSeqR_family$Class,
                                         BallastSeqR_family$Family, sep = "_")

n_family_taxa <- length(unique(BallastSeqR_family$Class_family))
family_colors <- colorRampPalette(c(bcc_hex))(n_family_taxa)

BallastSeqR_family_bar <- ggplot(BallastSeqR_family,
                                 aes(x = Sample, y = Abundance, fill = Class_family)) +
  geom_bar(stat = "identity", colour = "black", linewidth = 0.3, position = "fill") +
  scale_fill_manual(values = family_colors) +
  guides(fill = guide_legend(reverse = FALSE, keywidth = 1, keyheight = 1)) +
  ylab("Relative Abundance (Family > 1%) \n") + xlab("Sample ID") +
  theme_minimal() + theme_ballast()

print(BallastSeqR_family_bar)

BallastSeqR_family_bar_interactive <- ggplotly(BallastSeqR_family_bar)
BallastSeqR_family_bar_interactive
```

## Family level, grouped by sample type

```{r bcc family sample type}

BallastSeqR_family_sample <- Ballast_physeq %>%
  tax_glom(taxrank = "Family") %>%
  transform_sample_counts(function(x) {x/sum(x)}) %>%
  psmelt() %>%
  group_by(Sample_type, Kingdom, Phylum, Class, Order, Family) %>%
  dplyr::summarize(Mean = mean(Abundance, na.rm = TRUE), .groups = "keep") %>%
  filter(Mean > 0.01) %>%
  arrange(Class)

BallastSeqR_family_sample$Class_family <- paste(BallastSeqR_family_sample$Class,
                                                BallastSeqR_family_sample$Family, sep = "_")

n_family_sample_taxa <- length(unique(BallastSeqR_family_sample$Class_family))
family_sample_colors <- colorRampPalette(c(bcc_hex))(n_family_sample_taxa)

BallastSeqR_sample_family_bar <- ggplot(BallastSeqR_family_sample,
                                        aes(x = Sample_type, y = Mean, fill = Class_family)) +
  geom_bar(stat = "identity", colour = "black", linewidth = 0.3, position = "fill") +
  scale_fill_manual(values = family_sample_colors) +
  guides(fill = guide_legend(reverse = FALSE, keywidth = 1, keyheight = 1)) +
  ylab("Relative Abundance (Family > 1%) \n") + xlab("Sample Type") +
  theme_minimal() + theme_ballast()

print(BallastSeqR_sample_family_bar)

BallastSeqR_sample_family_bar_interactive <- ggplotly(BallastSeqR_sample_family_bar)
BallastSeqR_sample_family_bar_interactive
```

## Family level, grouped by sample type and voyage

Creating the same bar plot as above, but with subsets within sample type showing the different voyages, since similar sample types that were collected during different voyages cluster separately on the NMDS and PCoA plots.

```{r bcc family sample type and voyage}

BallastSeqR_family_voyage <- Ballast_physeq %>%
  tax_glom(taxrank = "Family") %>%
  transform_sample_counts(function(x) {x/sum(x)}) %>%
  psmelt() %>%
  group_by(Sample_type, Voyage, Kingdom, Phylum, Class, Order, Family) %>%
  dplyr::summarize(Mean = mean(Abundance, na.rm = TRUE), .groups = "keep") %>%
  filter(Mean > 0.01) %>%
  arrange(Class)

BallastSeqR_family_voyage$Class_family <- paste(BallastSeqR_family_voyage$Class,
                                                BallastSeqR_family_voyage$Family, sep = "_")

n_family_voyage_taxa <- length(unique(BallastSeqR_family_voyage$Class_family))
family_voyage_colors <- colorRampPalette(c(bcc_hex))(n_family_voyage_taxa)

BallastSeqR_voyage_family_bar <- ggplot(BallastSeqR_family_voyage,
                                        aes(x = Voyage, y = Mean, fill = Class_family)) +
  geom_bar(stat = "identity", colour = "black", linewidth = 0.3, position = "fill") +
  scale_fill_manual(values = family_voyage_colors) +
  guides(fill = guide_legend(reverse = FALSE, keywidth = 1, keyheight = 1)) +
  ylab("Relative Abundance (Family > 1%) \n") + xlab("Voyage") +
  theme_minimal() + theme_ballast() +
  facet_grid(. ~ Sample_type, margins = TRUE, scale = "free")

print(BallastSeqR_voyage_family_bar)

BallastSeqR_voyage_family_bar_interactive <- ggplotly(BallastSeqR_voyage_family_bar)
BallastSeqR_voyage_family_bar_interactive
```

## Genus level, grouped by voyage and sample type

Using `ggnested` for class-grouped legend colors.

```{r bcc genus voyage and sample type}

library(ggnested)

BallastSeqR_genus_voyage <- Ballast_physeq %>%
  tax_glom(taxrank = "Genus") %>%
  transform_sample_counts(function(x) {x/sum(x)}) %>%
  psmelt() %>%
  group_by(Sample_type, Voyage, Kingdom, Phylum, Class, Order, Family, Genus) %>%
  dplyr::summarize(Mean = mean(Abundance, na.rm = TRUE), .groups = "keep") %>%
  filter(Mean > 0.01) %>%
  arrange(Class)

BallastSeqR_genus_voyage_filt <- data.frame(BallastSeqR_genus_voyage)

pal_genus <- c("#440154FF", "blue", "green", "yellow", "purple", "orange", "#FF6699")
names(pal_genus) <- unique(BallastSeqR_genus_voyage_filt$Class)

BallastSeqR_voyage_genus_bar_ggnested <- ggnested::ggnested(
  BallastSeqR_genus_voyage_filt,
  aes(x = Sample_type, y = Mean, main_group = Class, sub_group = Genus),
  main_palette = pal_genus) +
  geom_bar(stat = "identity", colour = "black", linewidth = 0.3, position = "fill") +
  facet_wrap(~ Voyage, scale = "free") +
  guides(fill = guide_legend(reverse = FALSE, keywidth = 1, keyheight = 1)) +
  ylab("Relative Abundance (Genus > 1%) \n") + xlab("Sample Type") +
  theme_minimal() + theme_ballast()

print(BallastSeqR_voyage_genus_bar_ggnested)

ggsave(filename = "R output/BallastSeqR_voyage_sample_genus_bar.svg",
       plot = BallastSeqR_voyage_genus_bar_ggnested, width = 10, height = 6, device = svg)
```

## Species level, grouped by voyage and sample type

```{r bcc species voyage and sample type}

#Examining BCC by sample type & voyage
BallastSeqR_species_voyage <- Ballast_physeq %>%
  tax_glom(taxrank = "Species") %>%                   #Agglomerate at species level
  transform_sample_counts(function(x) {x/sum(x)}) %>% #Transform to rel. abundance
  psmelt() %>%                                        #Melt to long format
  group_by(Sample_type, Voyage, 
           Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
  dplyr::summarize(Mean = 
                     mean(Abundance, na.rm=TRUE)) %>% #Calculate average
  arrange(Class)

```

```{r ggnested species}

#Convert to a dataframe
BallastSeqR_species_voyage_filt <- data.frame(BallastSeqR_species_voyage) 

#Custom color palette
pal <- c("#440154FF", "blue", "green", "yellow", "purple", "orange", "#FF6699",
         "#440154FF", "blue", "green", "yellow", "purple", "orange", "#FF6699",
         "#440154FF", "blue", "green", "yellow", "purple", "orange", "#FF6699",
         "#440154FF", "blue", "green", "yellow", "purple", "orange", "#FF6699",
          "#440154FF", "blue", "green", "yellow")
names(pal) <- unique(BallastSeqR_species_voyage_filt$Class)

#Use ggnested to create the bar plot
BallastSeqR_voyage_species_bar_ggnested <- ggnested::ggnested(BallastSeqR_species_voyage_filt, 
         aes(x = Sample_type, 
             y = Mean, 
             main_group = Class, 
             sub_group = Species),
          main_palette = pal) + 
  geom_bar(stat = "identity", colour = "black", linewidth = 0.3, position = "fill") +
  facet_wrap(~ Voyage, scale="free") +
  guides(fill = guide_legend(reverse = FALSE, keywidth = 1, keyheight = 1)) +
  ylab("Relative Abundance (Species > 1%) \n") + xlab("Sample Type") +
  theme_minimal()+theme(
    strip.text = element_text(size=12),
    panel.border = element_rect(fill=NA, colour = "black"), 
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(), 
    axis.title.x = element_text(size=12),
    axis.title.y = element_text(size=12),
    axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, size=10, 
                              colour="black"), 
    axis.text.y = element_text(size=10, colour="black"), 
    plot.title = element_text(hjust = 0.5), 
    axis.ticks.x = element_line(colour="#000000", linewidth=0.1), 
    axis.ticks.y = element_line(colour="#000000", linewidth=0.1),
    legend.text = element_text(size = 10)) +
  facet_wrap(~ Voyage, scale="free")

print(BallastSeqR_voyage_species_bar_ggnested)
```

### Create Figure 3: Faceted grouped bar plot

Top 10 species from the top 3 classes, shown as a faceted grouped bar plot with a common baseline. Each panel shows one species, with sample types on the x-axis and mean relative abundance (+/- SE) on the y-axis. This layout allows direct comparison of each taxon's abundance across treatments.

```{r fig3 data prep}

library("fantaxtic")

# Merge samples for identifying top taxa (used only for ranking)
Ballast_physeq_merged <- speedyseq::merge_samples2(Ballast_physeq, "Type_voyage",
                                                     fun_otu = mean)

top_nested <- nested_top_taxa(Ballast_physeq_merged,
                              top_tax_level = "Class",
                              nested_tax_level = "Species",
                              n_top_taxa = 3,
                              n_nested_taxa = 10)

# Extract class-species pairs from fantaxtic (preserves which species belongs to which class)
top_class_species <- tax_table(top_nested$ps_obj) %>%
  as.data.frame() %>%
  filter(Species != "Other", Class != "Other") %>%
  distinct(Class, Species)

top_class_names <- unique(top_class_species$Class)
top_species_names <- unique(top_class_species$Species)

# Clean species names (used for matching before gsub is applied to fig3_data)
top_species_names_clean <- gsub("_", " ", top_species_names)

# Now compute per-sample relative abundances at the species level (not merged)
fig3_data <- Ballast_physeq %>%
  tax_glom(taxrank = "Species") %>%
  transform_sample_counts(function(x) {x / sum(x)}) %>%
  psmelt() %>%
  # Join on class-species pairs to avoid cross-class contamination
  inner_join(top_class_species, by = c("Class", "Species")) %>%
  group_by(Sample_type, Voyage, Class, Species) %>%
  dplyr::summarize(
    Mean = mean(Abundance, na.rm = TRUE),
    SE = sd(Abundance, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Order Class as a factor so panels group by class
fig3_data$Class <- factor(fig3_data$Class, levels = top_class_names)

# Replace underscores with spaces in species names
fig3_data$Species <- gsub("_", " ", fig3_data$Species)

# Remove sample types that don't exist in Voyage 2
fig3_data <- fig3_data %>%
  filter(!(Voyage == "Voyage 2" & Sample_type %in% c("Ocean uptake", "BWT+BWE"))) %>%
  droplevels()
```

#### Calculate proportion of community represented by top taxa

```{r fig3 community proportion}

# Per-sample relative abundances at species level
fig3_all_species <- Ballast_physeq %>%
  tax_glom(taxrank = "Species") %>%
  transform_sample_counts(function(x) {x / sum(x)}) %>%
  psmelt()

# Clean species names to match
fig3_all_species$Species <- gsub("_", " ", fig3_all_species$Species)

# Total relative abundance accounted for by the top species per sample
fig3_top_proportion <- fig3_all_species %>%
  mutate(is_top = Species %in% gsub("_", " ", top_species_names) &
                  Class %in% top_class_names) %>%
  group_by(Sample, Sample_type, Voyage) %>%
  dplyr::summarize(
    top_taxa_abundance = sum(Abundance[is_top]),
    .groups = "drop"
  )

# Summary across all samples
cat("Proportion of community composition represented by top 10 species in top 3 classes:\n")
cat(sprintf("  Mean:   %.1f%%\n", mean(fig3_top_proportion$top_taxa_abundance) * 100))
cat(sprintf("  Median: %.1f%%\n", median(fig3_top_proportion$top_taxa_abundance) * 100))
cat(sprintf("  Range:  %.1f%% - %.1f%%\n",
            min(fig3_top_proportion$top_taxa_abundance) * 100,
            max(fig3_top_proportion$top_taxa_abundance) * 100))

# Breakdown by sample type
fig3_proportion_by_type <- fig3_top_proportion %>%
  group_by(Sample_type) %>%
  dplyr::summarize(
    Mean_pct = mean(top_taxa_abundance) * 100,
    SD_pct = sd(top_taxa_abundance) * 100,
    .groups = "drop"
  )

knitr::kable(fig3_proportion_by_type,
             col.names = c("Sample type", "Mean %", "SD %"),
             digits = 1) %>%
  kableExtra::kable_styling("striped", latex_options = "scale_down")
```

```{r fig3 color palette}

# Create class-based color palette
n_classes <- length(top_class_names)
class_base_colors <- RColorBrewer::brewer.pal(max(3, n_classes), "Dark2")[1:n_classes]
names(class_base_colors) <- top_class_names

# Map each species to a shade of its parent class color
species_colors <- fig3_data %>%
  distinct(Class, Species) %>%
  arrange(Class, Species) %>%
  group_by(Class) %>%
  mutate(
    n = n(),
    idx = row_number(),
    color = colorRampPalette(
      c(class_base_colors[as.character(Class[1])],
        colorspace::lighten(class_base_colors[as.character(Class[1])], 0.6))
    )(n[1])[idx]
  ) %>%
  ungroup()

species_pal <- setNames(species_colors$color, species_colors$Species)

# Create a combined Class_Species factor for faceting, ordered by Class then Species
fig3_data <- fig3_data %>%
  arrange(Class, Species) %>%
  mutate(Class_Species = factor(Species, levels = unique(Species)))
```

```{r fig3 grouped bar plot, fig.width=16, fig.height=14}

fig3 <- ggplot(fig3_data, aes(x = Sample_type, y = Mean, fill = Sample_type)) +
  geom_col(width = 0.7, colour = "black", linewidth = 0.2) +
  geom_errorbar(aes(ymin = pmax(Mean - SE, 0), ymax = Mean + SE),
                width = 0.25, linewidth = 0.3) +
  facet_nested(Class + Class_Species ~ Voyage, scales = "free_y",
               nest_line = element_line(colour = "black")) +
  scale_fill_manual(values = sample_type_colors, "Sample Type") +
  labs(x = "Sample type",
       y = "Mean relative abundance",
       caption = "Relative abundance calculated as proportion of total reads per sample.\nBars show mean across biological replicates; error bars show ± 1 SE.") +
  theme_minimal() +
  theme_ballast() +
  theme(
    strip.text.y = element_text(size = 9, angle = 0),
    strip.text.x = element_text(size = 11),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    plot.caption = element_text(size = 8, hjust = 0, colour = "grey40"),
    ggh4x.facet.nestline = element_line(linetype = "solid")
  )

print(fig3)
```

```{r fig3 save}

ggsave(filename = "R output/Fig3_grouped_bar_species.svg",
       plot = fig3, width = 12, height = 22, device = svg)

ggsave(filename = "R output/Fig3_grouped_bar_species.png",
       plot = fig3, width = 12, height = 22, dpi = 300)
```

##### Create this plot with patterns overlayed

```{r patterns and fantaxtic}

#Load ggpattern and magick packages so that patterns can be added over colors
library(ggpattern)
library(magick)

#This code prints a figure displaying possible magick patterns and their names that
#can be used in the following manual pattern scale function.
#From https://coolbutuseless.github.io/package/ggpattern/articles/pattern-magick.html

df1 <- data.frame(
  x    = rep(1:6, 9),
  y    = rep(1:9, each=6),
  name = gridpattern::names_magick,
  stringsAsFactors = FALSE
)


ggplot(df1) + 
  geom_tile_pattern(
    aes(x, y, pattern_type = I(name)),
    pattern       = 'magick',
    pattern_scale = 1.5,
    pattern_fill  = 'black', 
    width         = 0.9, 
    height        = 0.9
  ) + 
  geom_label(aes(x+0.4, y+0.4, label = name), hjust = 1, vjust = 1) + 
  theme_void() + 
  labs(
    title = "All the possible magick pattern names"
  ) +
  coord_fixed(1)
```

Create manual pattern scale function:

```{r pattern function}

#This assigns a pattern to each taxa and is used in place of 
#scale_pattern_type_manual; make sure all taxa from all data sets are included here
#so that this function can be used to create all graphs. The value 'gray100' leaves
#an assigned taxa blank.

scale_pattern_type_bacteria <- function(...){
  ggpattern:::manual_scale(
    'pattern_type',
    values = setNames(c('gray100', 'vertical', 
                        'bricks', 'gray100',
                        'gray100', 'horizontalsaw',
                        'hs_bdiagonal', 'verticalsaw',
                        'gray100', 'gray100',
                        'gray100', 'gray100',
                        'gray100', 
                        'gray100',
                        'gray100', 'gray100',
                        'gray100', 'gray100',
                        'gray100', 'gray100',
                        'gray100', 'gray100',
                        'gray100', 'hs_horizontal',
                        'gray100', 'gray100',
                        'gray100', 'gray100',
                        'gray100', 'gray100',
                        'gray100', 'gray100',
                        'gray100', 'gray100',
                        'gray100', 'gray100',
                        'gray100', 'gray100'),
                      
                      c("Oleispira Genus", "Pseudomonas Genus",
                        "Pseudomonas_anguilliseptica", "Aeromonas Genus", 
                        "Shewanella Genus", "Pseudoalteromonas Genus", 
                        "Vibrionaceae Family", "Rheinheimera Genus", 
                        "Comamonadaceae Family", "Polynucleobacter Genus",
                        "Clade_Ia Genus", "Clade_II Genus", 
                        "Caulobacter Genus", 
                        "Allorhizobium-Neorhizobium-Pararhizobium-Rhizobium Genus",
                        "Sphingorhabdus Genus", "Sphingomonadaceae Family", 
                        "Sulfitobacter Genus", "Roseovarius Genus", 
                        "Rhodobacteraceae Family", "Tistrella Genus", 
                        "uncultured_bacterium", "Mesoflavibacter_sp.", 
                        "Antarctic_bacterium", "Flavobacterium Genus", 
                        "Flavobacterium_jumunjinense", "Flavobacteriaceae Family",
                        "Fluviicola Genus", "Arcicella Genus", 
                        "Flectobacillus Genus", "Pseudarcicella Genus",
                        "Other", "Other Alphaproteobacteria",
                        "Other Bacteroidia", "Other Gammaproteobacteria")),
    
    ...
  )
}


#'verticalsaw', 'crosshatch', 'verticalrightshingle', 
 #                       'horizontal3', 'hs_diagcross', 'horizontalsaw',
  #                      'rightshingle', 'hexagons', 'circles', 'verticalbricks',
   #                     'smallfishscales', 'left45', 'right45', 
    #                    'verticalsaw',   'gray100',
     #                   'right30', 'hs_fdiagonal', 'hs_cross', 'vertical2'
```

Add patterns to plot:

```{r add patterns to fantaxtic plot}

type_voyage_order <- c("port_uptake_1", "BWT_1", "ocean_uptake_1",
                       "BWT_BWE_1", "port_uptake_2", "BWT_2")

#Plot
plot_pattern <- plot_nested_bar(ps_obj = top_nested$ps_obj,
                top_level = "Class",
                nested_level = "Species",
                sample_order = type_voyage_order) +
  facet_wrap(~ Voyage, scale="free") +
  labs(x = "Sample type", y = "Relative abundance") +
  scale_x_discrete(labels=c("Port uptake V1", "BWT V1", "Ocean uptake V1",
                          "BWT+BWE V1", "Port uptake V2", "BWT V2")) +
  geom_bar_pattern(position = "fill",
                   stat = "identity",
                   color = "black", 
                   pattern_fill = "black",
                   pattern_spacing = 0.05,
                   pattern = 'magick',
                   aes(pattern_type = Species)) +
  scale_pattern_type_bacteria(guide = "none") +
  
  theme_minimal()+theme(
    strip.text = element_text(size=12),
    panel.border = element_rect(fill=NA, colour = "black"), 
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(), 
    axis.title.x = element_text(size=12),
    axis.title.y = element_text(size=12),
    axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, size=10, 
                              colour="black"), 
    axis.text.y = element_text(size=10, colour="black"), 
    plot.title = element_text(hjust = 0.5), 
    axis.ticks.x = element_line(colour="#000000", linewidth=0.1), 
    axis.ticks.y = element_line(colour="#000000", linewidth=0.1),
    legend.text = element_text(size = 10))

print(plot_pattern)

```

Save as .svg:

```{r fantaxtic with patterns svg}

#Saving as an editable pdf
ggsave(filename="R output/BallastSeqR_fantaxtic_patterned.pdf", plot=plot_pattern, width=10, height=6, device=cairo_pdf)

```

#### Read abundance

Same plot as above, but here I'm using fantaxtic to create a plot displaying the number of reads for each of the top 10 species in the top 3 classes, rather than the relative abundance, as is shown above.

```{r total abundance fantaxtic class species}

#Order samples
order <- c("port_uptake_1", "BWT_1", "ocean_uptake_1",
                          "BWT_BWE_1", "port_uptake_2", "BWT_2")
order <- as.character(order)

#Get the top taxa
top_level <- "Class"
nested_level <- "Species"
sample_order <- order
top_asv <- nested_top_taxa(Ballast_physeq_merged,
                              top_tax_level = "Class",
                              nested_tax_level = "Species",
                              n_top_taxa = 3, 
                              n_nested_taxa = 10)


#Create names for NA taxa
ps_tmp <- top_asv$ps_obj %>%
  name_na_taxa()

#Add labels to taxa with the same names
ps_tmp <- ps_tmp %>%
  label_duplicate_taxa(tax_level = nested_level)

#Generate a palette basedon the phyloseq object
pal <- taxon_colours(ps_tmp,
                     tax_level = top_level)

#Convert physeq to df
psdf <- psmelt(ps_tmp)


#Move the merged labels to the appropriate positions in the plot:
#Top merged labels need to be at the top of the plot,
#nested merged labels at the bottom of each group
psdf <- move_label(psdf = psdf,
                   col_name = top_level,
                   label = "Other",
                   pos = 0)
psdf <- move_nested_labels(psdf,
                           top_level = top_level,
                           nested_level = nested_level,
                           top_merged_label = "Other",
                           nested_label = "Other",
                           pos = Inf)


# Reorder samples
if(!is.null(sample_order)){
  if(all(sample_order %in% unique(psdf$Sample))){
    psdf <- psdf %>%
      mutate(Sample = factor(Sample, levels = sample_order))
  } else {
    stop("Error: not all(sample_order %in% sample_names(ps_obj)).")
  }

}


#Generate a base plot
library(ggnested)
p <- ggnested(psdf,
              aes_string(main_group = top_level,
                         sub_group = nested_level,
                         x = "Sample",
                         y = "Abundance"),
              main_palette = pal) +
  facet_wrap(~ Voyage, scale="free") +
   scale_y_continuous(limits = c(0, 11000)) +
  labs(x = "Sample type", y = "Abundance") +
  scale_x_discrete(labels=c("Port uptake V1", "BWT V1", "Ocean uptake V1",
                          "BWT+BWE V1", "Port uptake V2", "BWT V2")) +
  theme_minimal()+theme(
    strip.text = element_text(size=12),
    panel.border = element_rect(fill=NA, colour = "black"), 
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(), 
    axis.title.x = element_text(size=12),
    axis.title.y = element_text(size=12),
    axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, size=10, 
                              colour="black"), 
    axis.text.y = element_text(size=10, colour="black"), 
    plot.title = element_text(hjust = 0.5), 
    axis.ticks.x = element_line(colour="#000000", linewidth=0.1), 
    axis.ticks.y = element_line(colour="#000000", linewidth=0.1),
    legend.text = element_text(size = 10))


#Add abundances
p <- p + geom_col()

#Use geom_col(position = position_fill()) if you want to plot the relative abundance

print(p)


#Make interactive plot
p_abundance <- ggplotly(p)

p_abundance

```

Save as svg:

```{r species fantaxtic reads svg}

ggsave(filename="R output/BallastSeqR_voyage_species_bar_interactive_fantaxtic_reads.svg", 
       plot=p, width=10, height=6, device=svg)
```

## Species level, grouped by voyage, sample type, and tank

### Create plot with fantaxtic

#### Relative abundance

```{r fantaxtic class species plot voyage sample type tank}

#Merge by sample_type and voyage; there is a separate category in the metadata that combines these categories together
Ballast_physeq_merged_group <- speedyseq::merge_samples2(Ballast_physeq, "Sample_group", fun_otu = mean)
SD <- speedyseq::merge_samples2(sample_data(Ballast_physeq), "Sample_group")
print(SD[, "Sample_group"])

#Now samples will be displayed by sample_group rather than individially when graphed using fantaxtic

#Load fantaxtic and create plot
library("fantaxtic")
                        
top_nested_group <- nested_top_taxa(Ballast_physeq_merged_group,
                              #grouping = "Type_voyage",
                              top_tax_level = "Class",
                              nested_tax_level = "Species",
                              n_top_taxa = 3, 
                              n_nested_taxa = 10)

#Order samples
order <- sample_data(top_nested_group$ps_obj) %>%
  data.frame() %>%
  arrange(Sample_group) %>%
  pull(Sample_group) %>%
  as.character()

#Order samples
order <- c("V1_UP_T5", "V1_UP_T6", "V1_BWT_T5", "V1_BWT_T6", "V1_Sea_T5",
           "V1_BWT_BWE_T5", "V2_UP_T2", "V2_UP_T6", "V2_BWT_T6")

#Plot
plot2 <- plot_nested_bar(ps_obj = top_nested_group$ps_obj,
                top_level = "Class",
                nested_level = "Species",
                sample_order = order) +
  facet_nested(~ Voyage + Sample_type, scale="free") +
  labs(x = "Sample group", y = "Relative abundance") +
  theme_minimal()+theme(
    strip.text = element_text(size=12),
    panel.border = element_rect(fill=NA, colour = "black"), 
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(), 
    axis.title.x = element_text(size=12),
    axis.title.y = element_text(size=12),
    axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, size=10, 
                              colour="black"), 
    axis.text.y = element_text(size=10, colour="black"), 
    plot.title = element_text(hjust = 0.5), 
    axis.ticks.x = element_line(colour="#000000", linewidth=0.1), 
    axis.ticks.y = element_line(colour="#000000", linewidth=0.1),
    legend.text = element_text(size = 10))

```

# Bacterial community composition dot plots

## Genus level, grouped by sample type and voyage

```{r genus dot plot}

#Calculate relative abundance at the genus level for each sample type and voyage
BallastSeqR_genus_voyage <- Ballast_physeq %>%
  tax_glom(taxrank = "Genus") %>%                     #Agglomerate at genus level
  transform_sample_counts(function(x) {x/sum(x)}) %>% #Transform to rel. abundance
  psmelt() %>%                                        #Melt to long format
  group_by(Sample_type, Voyage, 
           Kingdom, Phylum, Class, Order, Family, Genus) %>%
  dplyr::summarize(Mean = 
                     mean(Abundance, na.rm=TRUE),
                    .groups = "keep") %>%             #Calculate average
  filter(Mean > 0.01) %>%                             #Filter 
  arrange(Class)

#Create color palette
genus_col = colorRampPalette(brewer.pal(9,"Dark2"))(12)

#Create plot
genus_dot <- ggplot(BallastSeqR_genus_voyage, aes(x=Voyage, y=Genus, 
                                                  color=Class)) +  
  geom_point(aes(size=Mean), alpha = 1/2) +
  scale_color_manual(values = genus_col) +
  facet_nested( ~ Sample_type, scales = "free", 
                nest_line = element_line(colour = "black")) +
  theme_minimal()+
  theme(panel.border = element_rect(fill=NA, colour = "black"), 
        panel.grid.minor = element_blank(), 
        panel.grid.major = element_blank(), 
        axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, 
                                    size=8, colour="black"), 
        axis.text.y = element_text(size=8, colour="black"), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x = element_line(colour="#000000", linewidth = 0.1), 
        axis.ticks.y = element_line(colour="#000000", linewidth = 0.1),
        legend.text = element_text(size = 7),
        ggh4x.facet.nestline = element_line(linetype = "solid"))

print(genus_dot)
```

## Genus level, grouped by voyage and sample type

Same plot as above, but voyage is grouped first and sample type is subset under it.

```{r genus dot plot 2}

#Create plot
genus_dot_2 <- ggplot(BallastSeqR_genus_voyage, aes(x=Sample_type, y=Genus, 
                                                  color=Class)) +  
  geom_point(aes(size=Mean), alpha = 1/2) +
  scale_color_manual(values = genus_col) +
  facet_nested( ~ Voyage, scales = "free", 
                nest_line = element_line(colour = "black")) +
  theme_minimal()+
  theme(panel.border = element_rect(fill=NA, colour = "black"), 
        panel.grid.minor = element_blank(), 
        panel.grid.major = element_blank(), 
        axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, 
                                    size=8, colour="black"), 
        axis.text.y = element_text(size=8, colour="black"), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x = element_line(colour="#000000", linewidth = 0.1), 
        axis.ticks.y = element_line(colour="#000000", linewidth = 0.1),
        legend.text = element_text(size = 7),
        ggh4x.facet.nestline = element_line(linetype = "solid"))

print(genus_dot_2)
```

## Species level, grouped by sample type and voyage

Same as the first genus level plot, but displaying the abundance of species present in abundance \>1%.

```{r species dot plot}

#Calculate relative abundance at the genus level for each sample type and voyage
BallastSeqR_species_voyage <- Ballast_physeq %>%
  tax_glom(taxrank = "Species") %>%                   #Agglomerate at species level
  transform_sample_counts(function(x) {x/sum(x)}) %>% #Transform to rel. abundance
  psmelt() %>%                                        #Melt to long format
  group_by(Sample_type, Voyage, 
           Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
  dplyr::summarize(Mean = 
                     mean(Abundance, na.rm=TRUE),
                    .groups = "keep") %>%             #Calculate average
  filter(Mean > 0.01) %>%                             #Filter out <1% 
  arrange(Class)

#Create color palette
species_col = colorRampPalette(brewer.pal(9,"Dark2"))(12)

#Create plot
species_dot <- ggplot(BallastSeqR_species_voyage, aes(x=Voyage, y=Species, 
                                                  color=Class)) +  
  geom_point(aes(size=Mean), alpha = 1/2) +
  scale_color_manual(values = species_col) +
  facet_nested( ~ Sample_type, scales = "free", 
                nest_line = element_line(colour = "black")) +
  theme_minimal()+
  theme(panel.border = element_rect(fill=NA, colour = "black"), 
        panel.grid.minor = element_blank(), 
        panel.grid.major = element_blank(), 
        axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, 
                                    size=8, colour="black"), 
        axis.text.y = element_text(size=8, colour="black"), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x = element_line(colour="#000000", linewidth = 0.1), 
        axis.ticks.y = element_line(colour="#000000", linewidth = 0.1),
        legend.text = element_text(size = 7),
        ggh4x.facet.nestline = element_line(linetype = "solid"))

print(species_dot)

```

## Species level, grouped by voyage and sample type

Same as the second genus level plot, but displaying the abundance of species present in abundance \>1%.

```{r species dot plot 2}

#Create plot
species_dot_2 <- ggplot(BallastSeqR_species_voyage, aes(x=Sample_type, y=Species, 
                                                  color=Class)) +  
  geom_point(aes(size=Mean), alpha = 1/2) +
  scale_color_manual(values = species_col) +
  facet_nested( ~ Voyage, scales = "free", 
                nest_line = element_line(colour = "black")) +
  theme_minimal()+
  theme(panel.border = element_rect(fill=NA, colour = "black"), 
        panel.grid.minor = element_blank(), 
        panel.grid.major = element_blank(), 
        axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, 
                                    size=8, colour="black"), 
        axis.text.y = element_text(size=8, colour="black"), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x = element_line(colour="#000000", linewidth = 0.1), 
        axis.ticks.y = element_line(colour="#000000", linewidth = 0.1),
        legend.text = element_text(size = 7),
        ggh4x.facet.nestline = element_line(linetype = "solid"))

print(species_dot_2)


```

# Filter ASVs to determine how many were given species-level assignments

Recreate a fresh phyloseq object from the original .qza files so that the taxonomy is examined here without the relabeling (e.g. `tax_fix`) applied to `Ballast_physeq` earlier in the analysis.

```{r create new phyloseq object}

#Create a phyloseq object from the .qza files exported from qiime2 using
#the qiime2R package
Ballast_physeq2 <- qza_to_phyloseq(
  features="Qiime output/ballast-dada2-table.qza",
  tree="Qiime output/ballast-rooted-tree.qza",
  taxonomy="Qiime output/ballast-taxonomy.qza",
  metadata = "Qiime output/Metadata.tsv"
)

#Remove chloroplast, mitochondrial and archaeal sequences
Ballast_physeq2 <- subset_taxa(Ballast_physeq2, Kingdom != "d__Archaea")
Ballast_physeq2 <- subset_taxa(Ballast_physeq2, Order != "Chloroplast")
Ballast_physeq2 <- subset_taxa(Ballast_physeq2, Family != "Mitochondria")

#Remove blanks and positive controls
Ballast_physeq2 = subset_samples(Ballast_physeq2, Sample_type != "control")
Ballast_physeq2 = subset_samples(Ballast_physeq2, Sample_type != "negative")

#Remove samples with few reads
Ballast_physeq2 <- subset_samples(Ballast_physeq2,
                                  !(Sample_number %in% samples_to_exclude))

#Filter out low abundance ASVs
Ballast_physeq2 = prune_taxa(taxa_sums(Ballast_physeq2) > 2, Ballast_physeq2)

#Order factors
sample_data(Ballast_physeq2)$Sample_type <- factor(
  sample_data(Ballast_physeq2)$Sample_type,
  levels = sample_type_levels,
  labels = sample_type_labels)

sample_data(Ballast_physeq2)$Sample_number <- factor(
  sample_data(Ballast_physeq2)$Sample_number,
  levels = sample_numbers)

sample_data(Ballast_physeq2)$Voyage <- factor(
  sample_data(Ballast_physeq2)$Voyage,
  levels = c("1", "2"),
  labels = c("Voyage 1", "Voyage 2"))
```

```{r ASVs assigned to species}

#Use tax_mutate to edit the species level names
tax_table(Ballast_physeq2) <- tax_table(Ballast_physeq2) %>%
  tax_mutate(Species=gsub("_", " ", Species))

taxtab2 <- as.data.frame(tax_table(Ballast_physeq2))

#Create new dataframe that excludes any NA's in the species column
taxtabfilt <- taxtab2 %>% filter(!is.na(Species))

#Remove unassigned, uncultured and similar
taxtabfilt <- dplyr::filter(taxtabfilt, !grepl('unidentified|uncultured', Species))
taxtabfilt <- dplyr::filter(taxtabfilt, !grepl('metagenome|sp.', Species))

#Remove non-species 
taxtabfilt <- taxtabfilt %>%
  filter(!Species %in% 
           c("Antarctic bacterium", "bacterium enrichment", "microbial mat",
             "Rhodobacteraceae bacterium", "alpha proteobacterium", 
             "Actinobacteria bacterium", "Candidatus Planktophila")) 
```

```{r ASVs assigned to genus}

#Use tax_mutate to edit the species level names
tax_table(Ballast_physeq2) <- tax_table(Ballast_physeq2) %>%
  tax_mutate(Genus=gsub("_", " ", Genus))

taxtab3 <- as.data.frame(tax_table(Ballast_physeq2))

#Create new dataframe that excludes any NA's in the species column
taxtabfilt_genus <- taxtab3 %>% filter(!is.na(Genus))

#Remove unassigned, uncultured and similar
taxtabfilt_genus <- dplyr::filter(taxtabfilt_genus, 
                                  !grepl('unidentified|uncultured', Genus))
taxtabfilt_genus <- dplyr::filter(taxtabfilt_genus, 
                                  !grepl('clade|group', Genus))

#Remove non-genera 
taxtabfilt_genus <- taxtabfilt_genus %>%
  filter(!Genus %in% 
           c("env.OPS 17", "AKYH767",
             "Bacteroidetes vadinHA17", "Clade IV", "TRA3-20", "R7C24", 
             "JGI 0000069-P22", "DEV007", "UBA12409", "	67-14", "PeM15",
             "IMCC26256", "KD4-96", "S25-593", "T34", "Clade Ia", "Clade II", 
             "67-14", "C39", "Clade III", "GWD2-49-16", "MWH-Ta3", "OLB12", 
             "BSV13", "SH3-11", "SM1A02", "SH-PL14", "HTCC5015", "BIyi10",
             "Clostridium sensu stricto 1", "SUP05 cluster")) 
```

# Session Info

```{r info}

devtools::session_info()
```
