---
title: "Ballast water bacteria 16S rRNA analysis"
author: "Sarah Brown"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
    number_sections: true
  theme: bright
  highlight: default
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Packages and Import Data

Here, we load the required packages and then import the .qza files exported from QIIME2 using the qiime2R package. The final object is a phyloseq object called Ballast_physeq.

```{r packages, warning = FALSE, messages = FALSE, error = FALSE}

library(ggplot2)
library(vegan)
library(plyr)
library(dplyr)
library(scales)
library(grid)
library(reshape2)
library(phyloseq)
library(picante)
library(tidyr)
library(viridis)
library(qiime2R)
library(DESeq2)
library(patchwork)
library(RColorBrewer)
library(microViz)
library(speedyseq)
library(ComplexHeatmap)
library(plotly)
library(seecolor)
library(tibble)
library(tidyverse)
library(ggh4x)
library(kableExtra)
library(patchwork)

set.seed(13745)

#Create a phyloseq object from the .qza files exported from qiime2 using 
#the qiime2R package
Ballast_physeq <- qza_to_phyloseq(
  features="Qiime output/ballast-dada2-table.qza",
  tree="Qiime output/ballast-rooted-tree.qza",
  taxonomy="Qiime output/ballast-taxonomy.qza",
  metadata = "Qiime output/Metadata.tsv"
)
```

# Preparing the Data

Before performing any analyses, we need to examine the phyloseq object and remove chloroplast and mitochondrial sequences:

```{r prep}

#Check the rank names to make sure they are accurate
rank_names(Ballast_physeq)

#Correct output:[1] "Kingdom" "Phylum"  "Class"   "Order"   "Family"  
#                   "Genus"   "Species"

#Check sample variables
sample_variables(Ballast_physeq)

#Remove chloroplast sequences and any contaminant sequences
Ballast_physeq <- subset_taxa(Ballast_physeq, Kingdom != "d__Archaea")
Ballast_physeq <- subset_taxa(Ballast_physeq, Order != "Chloroplast")
Ballast_physeq <- subset_taxa(Ballast_physeq, Family != "Mitochondria")

#Check that contaminant sequences are removed (easiest to save as data frame and search)
taxtabl <- as.data.frame(tax_table(Ballast_physeq))

#At this point, blanks and positive controls should also be removed
Ballast_physeq = subset_samples(Ballast_physeq, Sample_type != "control")
Ballast_physeq = subset_samples(Ballast_physeq, Sample_type != "negative")
```

## Examine the total number of reads and ASV's

This part of the code examines the number of times each ASV occurs in the data set and the number of reads per sample; this will help determine which ASV's and samples need to be filtered out of the data set.

```{r asv, warning = FALSE}

#Create graph showing number of reads per ASV and number of reads per sample
readsumsdf = data.frame(nreads = sort(taxa_sums(Ballast_physeq), TRUE), 
                        sorted = 1:ntaxa(Ballast_physeq), type = "ASVs")
readsumsdf = rbind(readsumsdf, data.frame(nreads = sort(sample_sums(Ballast_physeq), TRUE), sorted = 1:nsamples(Ballast_physeq), type = "Samples"))
title = "Total number of reads"
p = ggplot(readsumsdf, aes(x = sorted, y = nreads)) + geom_bar(stat = "identity")
p + ggtitle(title) + scale_y_log10() + facet_wrap(~type, 1, scales = "free")
```

```{r ASV abundance}

#Create a dataframe that includes only the total number of times that each ASV occurs
readsumsdf_ASVs <- readsumsdf %>% 
    filter(type == "ASVs")  
  
#How many singletons are present?
length(which(readsumsdf_ASVs$nreads <= 0))
#Answer: 11
length(which(readsumsdf_ASVs$nreads == 1))
#Answer: 0

#How many doubletons are present?
length(which(readsumsdf_ASVs$nreads == 2))
#Answer: 29

#Create a bar plot showing the number of times each ASV occurs
ASV_occurance <- ggplot(data = readsumsdf_ASVs) +
    geom_bar(mapping = aes(x=nreads)) + 
    ylim(0, 20) +
    labs(x = "Number of reads", y = "Number of ASV's") +
    theme(text = element_text(size = 18), 
        axis.title = element_text(size = 15),
        panel.spacing = unit(1, "lines"), 
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5), 
        panel.background = element_blank())

print(ASV_occurance)
```

## Create table showing the number of reads per sample

```{r sample reads}

#Sum reads and sort by samples with least to greatest number of reads
samples_reads <- sort(sample_sums(Ballast_physeq))

#Create table 
knitr::kable(samples_reads) %>% 
    kableExtra::kable_styling("striped", 
                            latex_options="scale_down") %>% 
    kableExtra::scroll_box(width = "100%") 
 
 
```

## Create a rarefaction curve

```{r rarefaction, warning = FALSE, messages = FALSE}

#Rarefaction curve using vegan
#From: https://micca.readthedocs.io/en/latest/phyloseq.html
taxa_are_rows(Ballast_physeq)

mat <- t(otu_table(Ballast_physeq))
class(mat) <- "matrix"
class(mat)

mat <- as(t(otu_table(Ballast_physeq)), "matrix")
class(mat)

raremax <- min(rowSums(mat))

system.time(rarecurve(mat, step = 100, sample = raremax, col = "blue", label = FALSE))
```

## Filter samples

Based on the table above, there are some samples with very few to no reads, so we'll filter these out here. In this case, we keep all samples with more than 1,000 reads.

Removing these samples with low sequencing depths is important because samples with low reads typically are lower quality and have a greater probability of containing contaminant sequences ([Weiss et al., 2017](https://microbiomejournal.biomedcentral.com/articles/10.1186/s40168-017-0237-y)).

```{r filter}

#Remove samples with few reads
Ballast_physeq = subset_samples(Ballast_physeq, 
                            Sample_number != "104")
Ballast_physeq = subset_samples(Ballast_physeq, 
                            Sample_number != "105")
Ballast_physeq = subset_samples(Ballast_physeq, 
                            Sample_number != "107")
Ballast_physeq = subset_samples(Ballast_physeq, 
                            Sample_number != "108")
Ballast_physeq = subset_samples(Ballast_physeq, 
                            Sample_number != "109")
Ballast_physeq = subset_samples(Ballast_physeq, 
                            Sample_number != "106")

#Filter out low abundance ASVs
Ballast_physeq = prune_taxa(taxa_sums(Ballast_physeq) > 2, Ballast_physeq)

```

## Label and order variables

Now we can relabel the variables; if you type in Ballast_physeq you will see that sample_data is the matrix that holds the information that we want to change, so we need to include sample_data in our code here.

```{r label}

#Order factors
sample_data(Ballast_physeq)$Sample_type <- factor(sample_data(Ballast_physeq)$Sample_type, 
               levels = c("port_uptake", "BWT", "ocean_uptake", "BWT_BWE"),
               labels = c("Port uptake", "BWT", "Ocean uptake", "BWT+BWE"))

sample_data(Ballast_physeq)$Sample_number <- factor(sample_data(Ballast_physeq)$Sample_number, 
               levels = c("80", "81", "82", "83", "84", "85", "86", "87", "88", "89", "90", "91", "92", "93", "94", "95", "96", "97", "98", "100", "101", "102", "103", "110", "111", "112", "113", "114", "115"),
               labels = c("80", "81", "82", "83", "84", "85", "86", "87", "88", "89", "90", "91", "92", "93", "94", "95", "96", "97", "98", "100", "101", "102", "103", "110", "111", "112", "113", "114", "115"))

sample_data(Ballast_physeq)$Voyage <- factor(sample_data(Ballast_physeq)$Voyage, 
               levels = c("1", "2"),
               labels = c("Voyage 1", "Voyage 2"))
```

## Group rarefaction curves by Sample_type

Comparing the sequencing depth and species richness between different sample types. Using the ggrare function in the phyloseq.extended package which wraps vegan's rarecurve function so that it can be easily used with phyloseq objects.

Note: I load the phyloseq-extended package here instead of above in the 'Load packages' section because it seems to create conflicts with the phyloseq package and results in several warning messages.

```{r  rarefaction 2, warning = FALSE, messages = FALSE}

library(phyloseq.extended)

p <- ggrare(Ballast_physeq, step = 1000, color = "Sample_type", label = "Sample_number", se = FALSE) 
 
  
p <- p + facet_wrap(~Sample_type, scale="free") +
  scale_colour_manual(values = c("Port uptake" = "#E7298A", "Ocean uptake" = "#7570B3", "BWT" = "#1B9E77", "BWT+BWE" = "#D95F02"), "Sample Type") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none") +
  labs(x ="Number of Reads")


print(p)

#Save as svg
ggsave(filename="R output/rarecurve_sampletype.svg", plot=p, width=8, height=6, device=svg)
  
```

## Abundance transformation

We have to transform the count data to account for differences in library size between samples. This is often done using rarefaction, but rarefaction results in the elimination of valid data and can also increase the rate of false positives when testing for ASV's that are differentially abundant in different sample categories.

Additionally, here are some helpful references regarding library normalization strategies:

[Waste Not, Want Not: Why Rarefying Microbiome Data is Inadmissible](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1003531)

[Normalization and microbial differential abundance strategies depend on data characteristics](https://microbiomejournal.biomedcentral.com/articles/10.1186/s40168-017-0237-y)

Below, we'll use relative abundance as our variance-stabilizing method.

```{r abundance}

#Transform to relative abundance
Ballast_physeqRA = transform_sample_counts(Ballast_physeq, function(x){x / sum(x)})
```

Alternatively, you could also choose to log-transform the data:

```{r log transform}

Ballast_physeqlog = transform_sample_counts(Ballast_physeq, function(x) log(1 + x))
```

# Alpha Diversity

Note that while these data are not rarefied, they are normalized. We'll use the Shannon diversity index (based on richness AND evenness; examines how many different taxa are present and how evenly they're distributed within a sample) to analyze alpha diversity between variable types. This means it considers both the number of species and the inequality between species abundances.

## Calculate Shannon Diversity per Sample

Using the estimate_richness function in the phyloseq package to calculate Shannon diversity. The estimate_richness function can also take measures "Chao1" "ACE" "Simpson" and "Fisher".

Note that running this code results in an error message stating that the provided data does not have any singletons. See here for more information:

<https://github.com/benjjneb/dada2/issues/214>

<https://forum.qiime2.org/t/singletons-and-diversity-richness-indices/2971/9>

```{r shannon, warning = FALSE, messages = FALSE}

#Change the - in the sample names to _ to prevent the estimate_richness function from changing the - to a . during the calculation of Observed richness
sample_names(Ballast_physeq) <- paste(gsub("-", "_", sample_names(Ballast_physeq)), sep = "_")

#Calculating richness - Shannon diversity and number of observations (Observed) in a new dataframe
richness <- data.frame(estimate_richness(Ballast_physeq, measures = c("Shannon", "Observed")))

richness <- setNames(cbind(rownames(richness), richness, row.names = NULL), 
                     c("sample-id", "Observed", "Shannon"))

#Add the sample metadata to the dataframe
s <- data.frame(sample_data(Ballast_physeq))
s <- setNames(cbind(rownames(s), s, row.names = NULL), 
              c("sample-id", "Sample_type", "Voyage", "Tank", 
                "Sample_number", "Sample_site", "Sample_date"))

alphadiv <- merge(s, richness, by = "sample-id")

#Order factors
alphadiv$Sample_type <- factor(alphadiv$Sample_type, 
                      levels = c("Port uptake", "BWT", "Ocean uptake", "BWT+BWE"),
                      labels = c("Port uptake", "BWT", "Ocean uptake", "BWT+BWE")) 

#Shows the calculated indices
knitr::kable(head(alphadiv)) %>% 
  kableExtra::kable_styling("striped", latex_options="scale_down") %>% 
  kableExtra::scroll_box(width = "100%")
```

## Create Alpha Diversity Boxplots

### Shannon diversity: sample type

```{r boxplot}

#Plot
ggplot(data=alphadiv, aes(x=Sample_type, y=Shannon), alpha=0.1) + 
  geom_boxplot(aes(fill=Sample_type)) +
  scale_fill_manual(values = c("Port uptake" = "#E7298A", 
                                 "BWT" = "#1B9E77", 
                                 "Ocean uptake" = "#7570B3",
                                 "BWT+BWE" = "#D95F02"), "Sample Type") +
  geom_point(position=position_dodge(width=0.75),aes(group=Sample_type)) +
  scale_y_continuous(limits = c(0, 4.5)) +
  theme(legend.position="right",
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5), 
        panel.background = element_blank(), 
        axis.title.x = element_blank())
```

### Shannon diversity: sample type and voyage

```{r boxplot2}

#Plot
ggplot(data=alphadiv, aes(x=Voyage, y=Shannon), alpha=0.1) + 
  geom_boxplot(aes(fill=Sample_type)) +
  scale_fill_manual(values = c("Port uptake" = "#E7298A", 
                                 "Ocean uptake" = "#7570B3", 
                                 "BWT" = "#1B9E77", 
                                 "BWT+BWE" = "#D95F02"), "Sample Type") +
  geom_point(position=position_dodge(width=0.75),aes(group=Sample_type)) +
  facet_wrap(~Sample_type, scale="free") +
  theme(legend.position="right",
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5), 
        panel.background = element_blank(),
        axis.title.x = element_blank())
```

### Shannon diversity: voyage and sample type

```{r boxplot3}

#Plot
ggplot(data=alphadiv, aes(x=Sample_type, y=Shannon), alpha=0.1) + 
  geom_boxplot(aes(fill=Sample_type)) +
  scale_fill_manual(values = c("Port uptake" = "#E7298A", 
                                 "Ocean uptake" = "#7570B3", 
                                 "BWT" = "#1B9E77", 
                                 "BWT+BWE" = "#D95F02"), "Sample Type") +
  geom_point(position=position_dodge(width=0.75),aes(group=Voyage)) +
  facet_wrap(~Voyage, scale="free") +
  scale_y_continuous(limits = c(0, 4.5)) +
  theme(legend.position="right",
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5), 
        panel.background = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1))
```

### Richness: sample type

```{r boxplot richness 1}

#Plot
ggplot(data=alphadiv, aes(x=Sample_type, y=Observed), alpha=0.1) + 
  geom_boxplot(aes(fill=Sample_type)) +
  scale_fill_manual(values = c("Port uptake" = "#E7298A", 
                                 "BWT" = "#1B9E77", 
                                 "Ocean uptake" = "#7570B3",
                                 "BWT+BWE" = "#D95F02"), "Sample Type") +
  geom_point(position=position_dodge(width=0.75),aes(group=Sample_type)) +
  scale_y_continuous(limits = c(0, 100)) +
  theme(legend.position="right",
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5), 
        panel.background = element_blank(), 
        axis.title.x = element_blank())
```

### Richness: voyage and sample type

```{r boxplot richness 2}

#Plot
ggplot(data=alphadiv, aes(x=Sample_type, y=Observed), alpha=0.1) + 
  geom_boxplot(aes(fill=Sample_type)) +
  scale_fill_manual(values = c("Port uptake" = "#E7298A", 
                                 "Ocean uptake" = "#7570B3", 
                                 "BWT" = "#1B9E77", 
                                 "BWT+BWE" = "#D95F02"), "Sample Type") +
  geom_point(position=position_dodge(width=0.75),aes(group=Voyage)) +
  facet_wrap(~Voyage, scale="free") +
  scale_y_continuous(limits = c(0, 100)) +
  theme(legend.position="right",
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5), 
        panel.background = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1))
```

### Reads: voyage and sample type

```{r boxplot reads}

#Sum reads and sort by samples with least to greatest number of reads
samples_reads <- sort(sample_sums(Ballast_physeq))

samples_reads_df <- data.frame(sample_sums(Ballast_physeq))

#Make column out of the row names
samples_reads_df <- tibble::rownames_to_column(samples_reads_df, "sample-id")

#Add the sample metadata to the dataframe
s <- data.frame(sample_data(Ballast_physeq))
s <- setNames(cbind(rownames(s), s, row.names = NULL), 
              c("sample-id", "Sample_type", "Voyage", "Tank", 
                "Sample_number", "Sample_site", "Sample_date"))

reads_df <- merge(s, samples_reads_df, by = "sample-id")

#Order factors
reads_df$Sample_type <- factor(reads_df$Sample_type, 
                      levels = c("Port uptake", "BWT", "Ocean uptake", "BWT+BWE"),
                      labels = c("Port uptake", "BWT", "Ocean uptake", "BWT+BWE"))


#Plot
reads_box <- ggplot(data=reads_df, aes(x=Sample_type, y=sample_sums.Ballast_physeq.), 
       alpha=0.1) + 
  labs(y = "Reads") +
  geom_boxplot(aes(fill=Sample_type)) +
  scale_fill_manual(values = c("Port uptake" = "#E7298A", 
                                 "Ocean uptake" = "#7570B3", 
                                 "BWT" = "#1B9E77", 
                                 "BWT+BWE" = "#D95F02"), "Sample Type") +
  geom_point(position=position_dodge(width=0.75),aes(group=Voyage)) +
  facet_wrap(~Voyage, scale="free") +
  #scale_y_continuous(limits = c(0, 100)) +
  theme(legend.position="right",
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5), 
        panel.background = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1))
```

### Combine plots for Fig. 2

```{r fig 1 combo}

#Observed without x-axis labels
richness_box <- ggplot(data=alphadiv, aes(x=Sample_type, y=Observed), alpha=0.1) + 
  geom_boxplot(aes(fill=Sample_type)) +
  scale_fill_manual(values = c("Port uptake" = "#E7298A", 
                                 "Ocean uptake" = "#7570B3", 
                                 "BWT" = "#1B9E77", 
                                 "BWT+BWE" = "#D95F02"), "Sample Type") +
  geom_point(position=position_dodge(width=0.75),aes(group=Voyage)) +
  facet_wrap(~Voyage, scale="free") +
  scale_y_continuous(limits = c(0, 100)) +
  theme(legend.position="right",
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5), 
        panel.background = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

#Shannon without x-axis labels
shannon_box <- ggplot(data=alphadiv, aes(x=Sample_type, y=Shannon), alpha=0.1) + 
  geom_boxplot(aes(fill=Sample_type)) +
  scale_fill_manual(values = c("Port uptake" = "#E7298A", 
                                 "Ocean uptake" = "#7570B3", 
                                 "BWT" = "#1B9E77", 
                                 "BWT+BWE" = "#D95F02"), "Sample Type") +
  geom_point(position=position_dodge(width=0.75),aes(group=Voyage)) +
  facet_wrap(~Voyage, scale="free") +
  scale_y_continuous(limits = c(0, 4.5)) +
  theme(legend.position="right",
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5), 
        panel.background = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks = element_blank())

#Combine with the sample coverage plot using patchwork
boxplot_combo <- richness_box / shannon_box / reads_box

#Save as an png file
ggsave(filename="R output/Fig2_combined_boxplot.png", 
       plot=boxplot_combo, width=5, height=8, device=png)
```

## Alpha Diversity statistics

Perform Wilcoxon test.

Use section 7.2 <https://microbiome.github.io/course_2021_radboud/alpha-diversity.html>

Some information from the tutorial linked above: "To further investigate if patient status could explain the variation of Shannon index, let's do a Wilcoxon test. This is a non-parametric test that doesn't make specific assumptions about the distribution, unlike popular parametric tests, such as the t test, which assumes normally distributed observations. Wilcoxon test can be used to estimate whether the differences between two groups is statistically significant."

### Subset alpha diversity dataframe

```{r alpha stats}

#Subset the alphadiversity dataframe by voyage
v1.alpha <- subset(alphadiv, Voyage == "Voyage 1")
v2.alpha <- subset(alphadiv, Voyage == "Voyage 2")

```

### Shannon diversity: sample type

Voyages are grouped together and differences in sample type are tested using a pairwise Wilcoxon test.

```{r wilcoxon shannon sample type}

sample.shan.wil <- pairwise.wilcox.test(alphadiv$Shannon, alphadiv$Sample_type, p.adjust.method="fdr", paired=FALSE) 
sample.shan.wil
```

```{r wilcoxon shannon sample type table}

#Convert output to dataframe
sample.shan.wil <- as.data.frame(sample.shan.wil$p.value)

#Create table of p-values
knitr::kable(sample.shan.wil) %>% 
  kableExtra::kable_styling("striped", 
                            latex_options="scale_down") %>% 
  kableExtra::scroll_box(width = "100%")
```

### Shannon diversity: voyage and sample type

Testing for differences in sample types in the two voyages separately using a pairwise Wilcoxon test.

```{r wilcoxon shannon voyage sample type}

#Voyage 1
v1.sample.shan.wil <- pairwise.wilcox.test(v1.alpha$Shannon, v1.alpha$Sample_type, p.adjust.method="fdr", paired=FALSE) 
v1.sample.shan.wil

#Voyage 2
v2.sample.shan.wil <- pairwise.wilcox.test(v2.alpha$Shannon, v2.alpha$Sample_type, p.adjust.method="fdr", paired=FALSE) 
v2.sample.shan.wil
```

```{r wilcoxon shannon voyage sample type table v1}

#Convert output to dataframe
v1.sample.shan.wil.df <- as.data.frame(v1.sample.shan.wil$p.value)

#Create table of p-values
knitr::kable(v1.sample.shan.wil.df) %>% 
  kableExtra::kable_styling("striped", 
                            latex_options="scale_down") %>% 
  kableExtra::scroll_box(width = "100%")

```

```{r wilcoxon shannon voyage sample type table v2}

#Convert output to dataframe
v2.sample.shan.wil.df <- as.data.frame(v2.sample.shan.wil$p.value)

#Create table of p-values
knitr::kable(v2.sample.shan.wil.df) %>% 
  kableExtra::kable_styling("striped", 
                            latex_options="scale_down") %>% 
  kableExtra::scroll_box(width = "100%")
```

### Richness: sample type

Voyages are grouped together and differences in sample type are tested using a pairwise Wilcoxon test.

```{r wilcoxon richness sample type}

sample.richness.wil <- pairwise.wilcox.test(alphadiv$Observed, alphadiv$Sample_type, p.adjust.method="fdr", paired=FALSE) 
sample.richness.wil
```

```{r wilcoxon richness sample type table}

#Convert output to dataframe
sample.richness.wil <- as.data.frame(sample.richness.wil$p.value)

#Create table of p-values
knitr::kable(sample.richness.wil) %>% 
  kableExtra::kable_styling("striped", 
                            latex_options="scale_down") %>% 
  kableExtra::scroll_box(width = "100%")
```

### Richness: voyage and sample type

Testing for differences in sample types in the two voyages separately using a pairwise Wilcoxon test.

```{r wilcoxon richness voyage sample type}

#Voyage 1
v1.sample.richness.wil <- pairwise.wilcox.test(v1.alpha$Observed, v1.alpha$Sample_type, p.adjust.method="fdr", paired=FALSE) 
v1.sample.richness.wil

#Voyage 2
v2.sample.richness.wil <- pairwise.wilcox.test(v2.alpha$Observed, v2.alpha$Sample_type, p.adjust.method="fdr", paired=FALSE) 
v2.sample.richness.wil
```

```{r wilcoxon richness voyage sample type table v1}

#Convert output to dataframe
v1.sample.richness.wil.df <- as.data.frame(v1.sample.richness.wil$p.value)

#Create table of p-values
knitr::kable(v1.sample.richness.wil.df) %>% 
  kableExtra::kable_styling("striped", 
                            latex_options="scale_down") %>% 
  kableExtra::scroll_box(width = "100%")

```

```{r wilcoxon richness voyage sample type table v2}

#Convert output to dataframe
v2.sample.richness.wil.df <- as.data.frame(v2.sample.richness.wil$p.value)

#Create table of p-values
knitr::kable(v2.sample.richness.wil.df) %>% 
  kableExtra::kable_styling("striped", 
                            latex_options="scale_down") %>% 
  kableExtra::scroll_box(width = "100%")

```

# Ordination plots

## Find plot color hex codes

```{r colors}

#For points
brewer.pal(4, "Dark2")

#For lines
brewer.pal(3, "Set1")

```

## Create NMDS with the Bray-Curtis dissimilarity matrix

Here, we use the relative abundance transformed data to create an NMDS plot.

Also, pay attention to the printout of the stress here - less than 0.05 means that the two dimensional plot is an excellent representation of the data (i.e., the NMDS ordination is a 'good fit' for the data), \<0.1 and 0.2 are also okay, but \> or = to 0.3 means that the NMDS ordination does not fit the data well.

```{r NMDS Bray}

all.nmds.source.ord <- ordinate(
  physeq = Ballast_physeqRA, 
  method = "NMDS", 
  distance = "bray"
)

#Plot, color coding by Sample_type
all.nmds.sample.type <- plot_ordination(
  physeq = Ballast_physeqRA,
  ordination = all.nmds.source.ord) + 
  scale_colour_manual(values = c("Port uptake" = "#E7298A", "Ocean uptake" = "#7570B3", "BWT" = "#1B9E77", "BWT+BWE" = "#D95F02"), "Sample Type") +
  scale_shape_manual(values = c("Voyage 1" = 16, "Voyage 2" = 17), name = "Voyage") +
  geom_point(mapping = aes(colour = factor(Sample_type), shape = factor(Voyage), size = 5)) +
  guides(size=FALSE) +
  guides(shape = guide_legend(override.aes = list(size = 3))) +
  theme(plot.title = element_text(size = 18),
        text = element_text(size = 18), 
        axis.title = element_text(size = 15),
        panel.spacing = unit(1, "lines"), 
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5), 
        panel.background = element_blank(), 
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15),
        legend.justification = c("right", "top")) 

print(all.nmds.sample.type)

```

## Create PCoA with the Bray-Curtis dissimilarity matrix

### PCoA with labels and arrows

```{r PCoA Bray labels}

all.pcoa.source.ord <- ordinate(
  physeq = Ballast_physeqRA, 
  method = "PCoA", 
  distance = "bray"
)

#Plot, color coding by Sample_type
all.pcoa.sample.type <- plot_ordination(
  physeq = Ballast_physeqRA,
  ordination = all.pcoa.source.ord) + 
  scale_colour_manual(values = c("Port uptake" = "#E7298A", "Ocean uptake" = "#7570B3", "BWT" = "#1B9E77", "BWT+BWE" = "#D95F02"), "Sample Type") +
  scale_shape_manual(values = c("Voyage 1" = 16, "Voyage 2" = 17), name = "Voyage") +
  geom_point(mapping = aes(colour = factor(Sample_type), shape = factor(Voyage), size = 5)) +
  geom_text(aes(label = Sample_number, colour = factor(Sample_type)), nudge_x=0.05, nudge_y=0.05, check_overlap = TRUE) +
  guides(size=FALSE) +
  guides(shape = guide_legend(override.aes = list(size = 3))) +
  theme(plot.title = element_text(size = 18),
        text = element_text(size = 18), 
        axis.title = element_text(size = 15),
        panel.spacing = unit(1, "lines"), 
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5), 
        panel.background = element_blank(), 
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15),
        legend.justification = c("right", "top")) +
  geom_segment(
  aes(x = -0.45, y = -0.30, xend = 0.30, yend = -0.30),
  arrow = arrow(length = unit(0.03, "npc"))) +
   geom_segment(
  aes(x = -0.08, y = 0.03, xend = 0.29, yend = -0.08),
  arrow = arrow(length = unit(0.03, "npc")), colour = "#E41A1C") +
   geom_segment(
  aes(x = -0.14, y = 0.35, xend = 0.04, yend = 0.34),
  arrow = arrow(length = unit(0.03, "npc"))) +
  geom_curve(
  aes(x = -0.14, y = 0.43, xend = 0.25, yend = 0.21),
  arrow = arrow(length = unit(0.03, "npc")), curvature = -0.3) +
  geom_segment(
  aes(x = 0.06, y = 0.34, xend = 0.29, yend = -0.08),
  arrow = arrow(length = unit(0.03, "npc")), colour = "#E41A1C") 
  

print(all.pcoa.sample.type)

```

### PCoA without labels and arrows

```{r PCoA Bray no labels}

all.pcoa.source.ord <- ordinate(
  physeq = Ballast_physeqRA, 
  method = "PCoA", 
  distance = "bray"
)

#Plot, color coding by Sample_type
all.pcoa.sample.type.no.label <- plot_ordination(
  physeq = Ballast_physeqRA,
  ordination = all.pcoa.source.ord) + 
  scale_colour_manual(values = c("Port uptake" = "#E7298A", "Ocean uptake" = "#7570B3", "BWT" = "#1B9E77", "BWT+BWE" = "#D95F02"), "Sample Type") +
  scale_shape_manual(values = c("Voyage 1" = 16, "Voyage 2" = 17), name = "Voyage") +
  geom_point(mapping = aes(colour = factor(Sample_type), shape = factor(Voyage), size = 5)) +
  guides(size=FALSE) +
  guides(shape = guide_legend(override.aes = list(size = 3))) +
  theme(plot.title = element_text(size = 18),
        text = element_text(size = 18), 
        axis.title = element_text(size = 15),
        panel.spacing = unit(1, "lines"), 
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5), 
        panel.background = element_blank(), 
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15),
        legend.justification = c("right", "top"))  
  

print(all.pcoa.sample.type.no.label)
```

# PERMANOVA Analysis

Using the relative abundance-transformed phyloseq object here (Ballast_physeqRA). The adonis PERMANOVA function requires the input to be an abundance table; the output of this function will tell you if environmental factors are influencing differences in bacterial community composition. Useful sources with information on how to interpret PERMANOVA results:

<https://forum.qiime2.org/t/adonis-vs-anosim-vs-permanova/9744>

<https://forum.qiime2.org/t/adonis-correct-interpretation-of-interaction-among-factors/10327>

<https://sites.google.com/site/mb3gustame/hypothesis-tests/manova/npmanova>

## PERMANOVA analysis

```{r Bray PERMANOVA}

#Run PERMANOVA with Bray-Curtis dissimilarity
permanova_bray <- phyloseq::distance(Ballast_physeqRA, method = "bray")
permanova_bray_df <- data.frame(sample_data(Ballast_physeqRA))
all <- adonis2(permanova_bray ~ Voyage*Sample_type, data = permanova_bray_df)
all


```

```{r PERMANOVA table}

#Convert output to dataframe
all.pvalue <- as.data.frame(all)

#Create table of p-values
knitr::kable(all.pvalue) %>% 
  kableExtra::kable_styling("striped", 
                            latex_options="scale_down") %>% 
  kableExtra::scroll_box(width = "100%")

```

## Pairwise PERMANOVA

Examine whether BWT from different tank was statistically the same

```{r pairwise PERMANOVA}

library(pairwiseAdonis)

#Run pairwise PERMANOVA with bray-curtis dissimilarity
Ballast_physeq_df <- data.frame(sample_data(Ballast_physeq))

Ballast_physeq_pairwise <- phyloseq::distance(Ballast_physeq, method = "bray")

Ballast.physeq.pairwise <- pairwise.adonis(Ballast_physeq_pairwise, Ballast_physeq_df$Sample_type, p.adjust.m='BH')

Ballast.physeq.pairwise
```

# Preliminary taxonomy fix

Some of the taxonomic levels in the phyloseq object have NA's or short values such as "g\_" which are uninformative when graphing bacterial taxonomy at those levels. The tax_fix function from the microViz package fills in these unknown values with useful information from higher taxonomic levels.

This should be run prior to running the following code for generating community composition barcharts, heatmaps, or any other analysis involving the visualization of bacterial community composition.

```{r tax fix}

#First, view the tax_table
knitr::kable(head(tax_table(Ballast_physeq))) %>% 
  kableExtra::kable_styling("striped") %>% 
  kableExtra::scroll_box(width = "100%")

#Now fix the labels
Ballast_physeq <- tax_fix(Ballast_physeq)

#View the relabeled tax_table
knitr::kable(head(tax_table(Ballast_physeq))) %>% 
  kableExtra::kable_styling("striped") %>% 
  kableExtra::scroll_box(width = "100%")
```

# Bacterial community composition bar charts

## Find color hex codes

I've included here a list of hex codes for colors that can be used to identify taxa in plots of bacterial community composition; these hex codes are visualized with the seecolor package.

```{r bcc hex, messages = FALSE}

bcc_hex <- print_color(c("black", "#440154FF", "#450659FF","#460B5EFF","#472D7AFF","#3B518BFF", "#3A548CFF", "#38598CFF", "#365C8DFF", "#34608DFF", "#33638DFF", "#31678EFF", "#306A8EFF","#2E6E8EFF", "#2D718EFF", "#2B748EFF","#2A778EFF", "#297B8EFF","#1F958BFF", "#1F988BFF", "#1E9C89FF", "#1F9F88FF","#1FA287FF", "#21A585FF", "#23A983FF","#25AC82FF", "#29AF7FFF", "#2DB27DFF","#32B67AFF", "#37B878FF", "#3CBC74FF","#57C766FF", "#5EC962FF","#A2DA37FF","#DAE319FF", "#E4E419FF",  "#ECE51BFF", "#F5E61FFF","darkorchid1", "darkorchid2", "darkorchid3","#A21C9AFF", "#A62098FF", "#AB2394FF", "#AE2892FF", "#B22B8FFF", "#B6308BFF", "#BA3388FF", "#BE3885FF", "#C13B82FF", "#C53F7EFF","#C8437BFF", "#CC4678FF", "#CE4B75FF", "#D14E72FF", "#D5536FFF", "#D7566CFF", "#DA5B69FF", "#DD5E66FF", "#E06363FF","#FF6699","#F68D45FF", "#FCA537FF","#F6E726FF", "#F4ED27FF", "#00489C", "#CCCCCC", "#999999", "#A1C299","#300018"), type = "r")

print(bcc_hex)
```

## Family level, individual samples

```{r bcc sample id}

#Examining BCC in each individual sample first
#BallastSeqR_family <- Ballast_physeq %>%
#  tax_glom(taxrank = "Family") %>%                    #Agglomerate at family level
#  transform_sample_counts(function(x) {x/sum(x)}) %>% #Transform to rel. abundance
#  psmelt() %>%                                        # Melt to long format
#  group_by(Sample, 
#           Kingdom, Phylum, Class, Order, Family) %>%
#  filter(Abundance > 0.01) %>%                        #Filter 
#  arrange(Class)

#Creating a new column that combines both the Class and Family level information
#so that Family level identifiers that are not unique to one Class (such as Ambiguous_taxa) aren't merged into a single large category when graphed
#BallastSeqR_family$Class_family <- paste(BallastSeqR_family$Class, BallastSeqR_family$Family, sep="_")

#Creating plot  
#BallastSeqR_family_bar <- ggplot(BallastSeqR_family, aes(x = Sample, y = Abundance, fill = Class_family)) + 
#  geom_bar(stat="identity", colour = "black", size=0.3, position = "fill") +
# scale_fill_manual(values = c("black", 
                               
#                               "#440154FF", "#450659FF","#460B5EFF",
                               
 #                              "#472D7AFF", 
                               
#                               "#3B518BFF", "#3A548CFF", "#38598CFF", "#365C8DFF", 
#                               "#34608DFF", "#33638DFF", "#31678EFF", "#306A8EFF",
#                               "#2E6E8EFF", "#2D718EFF", "#2B748EFF",
#                               "#2A778EFF", "#297B8EFF",
                               
#                               "#1F958BFF", "#1F988BFF", "#1E9C89FF", "#1F9F88FF",
#                               "#1FA287FF", "#21A585FF", "#23A983FF",
#                               "#25AC82FF", "#29AF7FFF", "#2DB27DFF",
#                               "#32B67AFF", "#37B878FF", "#3CBC74FF",
                               
#                               "#57C766FF", "#5EC962FF",
                               
#                               "#A2DA37FF",
                               
#                               "#DAE319FF", "#E4E419FF",  "#ECE51BFF", "#F5E61FFF"#,
                               
#                               "darkorchid1", "darkorchid2", "darkorchid3", 
                               
#                               "#A21C9AFF", "#A62098FF", "#AB2394FF", "#AE2892FF", "#B22B8FFF", "#B6308BFF", "#BA3388FF", "#BE3885FF", "#C13B82FF", "#C53F7EFF",
#                               "#C8437BFF", "#CC4678FF", "#CE4B75FF", "#D14E72FF", #"#D5536FFF", "#D7566CFF", "#DA5B69FF", "#DD5E66FF", "#E06363FF",
                               
#                               "#FF6699",
                               
#                               "#F68D45FF", "#FCA537FF",
                               
#                               "#F6E726FF", "#F4ED27FF", 
                               
                               
#                               "#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018",
                               
#                               "#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018",
#                               "#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018",
#                               "#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018")) +
  
#  guides(fill = guide_legend(reverse = FALSE, keywidth = 1, keyheight = 1)) +
#  ylab("Relative Abundance (Family > 1%) \n") + xlab("Sample ID") +
#  theme_minimal()+theme(panel.border = element_rect(fill=NA, colour = "black"), panel.grid.minor = element_blank(), 
#                        panel.grid.major = element_blank(), axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, size=8, colour="black"), 
#                        axis.text.y = element_text(size=8, colour="black"), plot#.title = element_text(hjust = 0.5), 
#                        axis.ticks.x = element_line(colour="#000000", size=0.1), axis.ticks.y = element_line(colour="#000000", size=0.1),
#                        legend.text = element_text(size = 7))  

#print(BallastSeqR_family_bar)

#Make interactive plot
#BallastSeqR_family_bar_interactive <- ggplotly(BallastSeqR_family_bar)

#BallastSeqR_family_bar_interactive

#Save html of interactive plot on knit
#htmlwidgets::saveWidget(BallastSeqR_family_bar_interactive, "BallastSeqR_family_bar_interactive.html")

#htmltools::tags$iframe(
#  src=file.path(getwd(), "BallastSeqR_family_bar_interactive.html"),
#  width="100%",
#  height="600",
#  scrolling="no",
#  seamless="seamless",
#  frameBorder="0"
#)
```

## Family level, grouped by sample type

```{r bcc sample type}

#Examining BCC by sample type
#BallastSeqR_family_sample <- Ballast_physeq %>%
#  tax_glom(taxrank = "Family") %>%                    #Agglomerate at family level
#  transform_sample_counts(function(x) {x/sum(x)}) %>% #Transform to rel. abundance
#  psmelt() %>%                                        #Melt to long format
#  group_by(Sample_type, 
#           Kingdom, Phylum, Class, Order, Family) %>%
#  dplyr::summarize(Mean = 
#                     mean(Abundance, na.rm=TRUE),
#                    .groups = "keep") %>%             #Calculate average
#  filter(Mean > 0.01) %>%                             #Filter 
#  arrange(Class)

#Creating a new column that combines both the Class and Family level information
#so that Family level identifiers that are not unique to one Class (such as Ambiguous_taxa) aren't merged into a single large category when graphed
#BallastSeqR_family_sample$Class_family <- paste(BallastSeqR_family_sample$Class, BallastSeqR_family_sample$Family, sep="_")

#Creating plot  
#BallastSeqR_sample_family_bar <- ggplot(BallastSeqR_family_sample, aes(x = Sample_type, y = Mean, fill = Class_family)) + 
#  geom_bar(stat="identity", colour = "black", size=0.3, position = "fill") +
#  scale_fill_manual(values = c("black", "#440154FF", "#450659FF","#460B5EFF","#472D7AFF", "#3B518BFF", "#3A548CFF", "#38598CFF", "#365C8DFF","#34608DFF", "#33638DFF", "#31678EFF", "#306A8EFF","#2E6E8EFF", "#2D718EFF", "#2B748EFF","#2A778EFF", "#297B8EFF","#1F958BFF", "#1F988BFF", "#1E9C89FF", "#1F9F88FF",
#"#1FA287FF", "#21A585FF", "#23A983FF","#25AC82FF", "#29AF7FFF", "#2DB27DFF","#32B67AFF", "#37B878FF", "#3CBC74FF","#57C766FF", "#5EC962FF","#A2DA37FF",
#"#DAE319FF", "#E4E419FF",  "#ECE51BFF", "#F5E61FFF","darkorchid1", "darkorchid2", "darkorchid3", "#A21C9AFF", "#A62098FF", "#AB2394FF", "#AE2892FF", "#B22B8FFF", "#B6308BFF", "#BA3388FF", "#BE3885FF", "#C13B82FF", "#C53F7EFF","#C8437BFF", "#CC4678FF", "#CE4B75FF", "#D14E72FF", "#D5536FFF", "#D7566CFF", "#DA5B69FF", "#DD5E66FF", "#E06363FF","#FF6699","#F68D45FF", "#FCA537FF","#F6E726FF", "#F4ED27FF", "#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018","#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018","#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018","#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018")) +
  
#  guides(fill = guide_legend(reverse = FALSE, keywidth = 1, keyheight = 1)) +
#  ylab("Relative Abundance (Family > 1%) \n") + xlab("Sample ID") +
#  theme_minimal()+theme(panel.border = element_rect(fill=NA, colour = "black"), panel.grid.minor = element_blank(), 
#                        panel.grid.major = element_blank(), axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, size=8, colour="black"), 
#                        axis.text.y = element_text(size=8, colour="black"), plot.title = element_text(hjust = 0.5), 
#                        axis.ticks.x = element_line(colour="#000000", size=0.1), axis.ticks.y = element_line(colour="#000000", size=0.1),
#                        legend.text = element_text(size = 7))  

#print(BallastSeqR_sample_family_bar)

#Make interactive plot
#BallastSeqR_sample_family_bar_interactive <- ggplotly(BallastSeqR_sample_family_bar)

#BallastSeqR_sample_family_bar_interactive

#Save html of interactive plot on knit
#htmlwidgets::saveWidget(BallastSeqR_sample_family_bar_interactive, "BallastSeqR_sample_family_bar_interactive.html")

#htmltools::tags$iframe(
#  src=file.path(getwd(), "BallastSeqR_sample_family_bar_interactive.html"),
#  width="100%",
#  height="600",
#  scrolling="no",
#  seamless="seamless",
#  frameBorder="0"
#)
```

## Family level, grouped by sample type and voyage

Creating the same bar plot as above, but with subsets within sample type showing the different voyages, since similar sample types that were collected during different voyages cluster separately on the NMDS and PCoA plots.

```{r bcc sample type and voyage}

#Examining BCC by sample type & voyage
#BallastSeqR_family_voyage <- Ballast_physeq %>%
#  tax_glom(taxrank = "Family") %>%                    #Agglomerate at family level
#  transform_sample_counts(function(x) {x/sum(x)}) %>% #Transform to rel. abundance
#  psmelt() %>%                                        # Melt to long format
#  group_by(Sample_type, Voyage, 
#           Kingdom, Phylum, Class, Order, Family) %>%
#  dplyr::summarize(Mean = 
#                     mean(Abundance, na.rm=TRUE))# %>% #Calculate average
#  filter(Mean > 0.01) %>%                             #Filter 
#  arrange(Class)

#Creating a new column that combines both the Class and Family level information
#so that Family level identifiers that are not unique to one Class (such as Ambiguous_taxa) aren't merged into a single large category when graphed
#BallastSeqR_family_voyage$Class_family <- paste(BallastSeqR_family_voyage$Class, BallastSeqR_family_voyage$Family, sep="_")

#Creating plot  
#BallastSeqR_voyage_family_bar <- ggplot(BallastSeqR_family_voyage, aes(x = Voyage, y = Mean, fill = Class_family)) + 
#  geom_bar(stat="identity", colour = "black", size=0.3, position = "fill") +
#  scale_fill_manual(values = c("black", "#440154FF", "#450659FF","#460B5EFF","#472D7AFF", "#3B518BFF", "#3A548CFF", "#38598CFF", "#365C8DFF","#34608DFF", "#33638DFF", "#31678EFF", "#306A8EFF","#2E6E8EFF", "#2D718EFF", "#2B748EFF","#2A778EFF", "#297B8EFF","#1F958BFF", "#1F988BFF", "#1E9C89FF", "#1F9F88FF",
#"#1FA287FF", "#21A585FF", "#23A983FF","#25AC82FF", "#29AF7FFF", "#2DB27DFF","#32B67AFF", "#37B878FF", "#3CBC74FF","#57C766FF", "#5EC962FF","#A2DA37FF",
#"#DAE319FF", "#E4E419FF",  "#ECE51BFF", "#F5E61FFF","darkorchid1", "darkorchid2", "darkorchid3", "#A21C9AFF", "#A62098FF", "#AB2394FF", "#AE2892FF", "#B22B8FFF", "#B6308BFF", "#BA3388FF", "#BE3885FF", "#C13B82FF", "#C53F7EFF","#C8437BFF", "#CC4678FF", "#CE4B75FF", "#D14E72FF", "#D5536FFF", "#D7566CFF", "#DA5B69FF", "#DD5E66FF", "#E06363FF","#FF6699","#F68D45FF", "#FCA537FF","#F6E726FF", "#F4ED27FF", "#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018","#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018","#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018","#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018")) +
  
#  guides(fill = guide_legend(reverse = FALSE, keywidth = 1, keyheight = 1)) +
#  ylab("Relative Abundance (Family > 1%) \n") + xlab("Voyage") +
#  theme_minimal()+theme(panel.border = element_rect(fill=NA, colour = "black"), 
#  panel.grid.minor = element_blank(),
#  panel.grid.major = element_blank(), 
#  axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, size=8, 
#                              colour="black"), 
#  axis.text.y = element_text(size=8, colour="black"), 
#  plot.title = element_text(hjust = 0.5), 
#  axis.ticks.x = element_line(colour="#000000", size=0.1), 
#  axis.ticks.y = element_line(colour="#000000", size=0.1),
#  legend.text = element_text(size = 7)) +
#  facet_grid(. ~ Sample_type, margins = TRUE, scale="free")

#print(BallastSeqR_voyage_family_bar)

#Make interactive plot
#p <- ggplotly(BallastSeqR_voyage_family_bar)

#p
```

```{r keep interactive plot}

#htmlwidgets::saveWidget(p, "BallastSeqR_voyage_family_bar_interactive.html")

#htmltools::tags$iframe(
#  src=file.path(getwd(), "BallastSeqR_voyage_family_bar_interactive.html"),
#  width="100%",
#  height="600",
#  scrolling="no",
#  seamless="seamless",
#  frameBorder="0"
#)
```

## Genus level, grouped by voyage and sample type

```{r bcc genus voyage and sample type}

#Examining BCC by sample type & voyage
#BallastSeqR_genus_voyage <- Ballast_physeq %>%
#  tax_glom(taxrank = "Genus") %>%                     #Agglomerate at genus level
#  transform_sample_counts(function(x) {x/sum(x)}) %>% #Transform to rel. abundance
#  psmelt() %>%                                        #Melt to long format
#  group_by(Sample_type, Voyage, 
#           Kingdom, Phylum, Class, Order, Family, Genus) %>%
#  dplyr::summarize(Mean = 
#                     mean(Abundance, na.rm=TRUE)) %>% #Calculate average
#  filter(Mean > 0.01) %>%                             #Filter 
#  arrange(Class)

#Creating a new column that combines both the Class and Family level information
#so that Family level identifiers that are not unique to one Class (such as Ambiguous_taxa) aren't merged into a single large category when graphed
#BallastSeqR_genus_voyage$Class_genus <- paste(BallastSeqR_genus_voyage$Class, BallastSeqR_genus_voyage$Genus, sep="_")

#Creating plot  
#BallastSeqR_voyage_genus_bar <- ggplot(BallastSeqR_genus_voyage, 
#  aes(x = Sample_type, y = Mean, fill = Class_genus)) +
#  geom_bar(stat="identity", colour = "black", size=0.3, position = "fill") +
#  scale_fill_manual(
#  breaks=BallastSeqR_genus_voyage$Class_genus,
#  labels=BallastSeqR_genus_voyage$Genus, 
#  values = c("black", "#440154FF", "#450659FF","#460B5EFF","#472D7AFF", "#3B518BFF", "#3A548CFF", "#38598CFF", "#365C8DFF","#34608DFF", "#33638DFF", "#31678EFF", "#306A8EFF","#2E6E8EFF", "#2D718EFF", "#2B748EFF","#2A778EFF", "#297B8EFF","#1F958BFF", "#1F988BFF", "#1E9C89FF", "#1F9F88FF",
#"#1FA287FF", "#21A585FF", "#23A983FF","#25AC82FF", "#29AF7FFF", "#2DB27DFF","#32B67AFF", "#37B878FF", "#3CBC74FF","#57C766FF", "#5EC962FF","#A2DA37FF",
#"#DAE319FF", "#E4E419FF",  "#ECE51BFF", "#F5E61FFF","darkorchid1", "darkorchid2", "darkorchid3", "#A21C9AFF", "#A62098FF", "#AB2394FF", "#AE2892FF", "#B22B8FFF", "#B6308BFF", "#BA3388FF", "#BE3885FF", "#C13B82FF", "#C53F7EFF","#C8437BFF", "#CC4678FF", "#CE4B75FF", "#D14E72FF", "#D5536FFF", "#D7566CFF", "#DA5B69FF", "#DD5E66FF", "#E06363FF","#FF6699","#F68D45FF", "#FCA537FF","#F6E726FF", "#F4ED27FF", "#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018","#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018","#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018","#00489C", "#CCCCCC", "#999999", "#A1C299", "#300018")) +
  
#  guides(fill = guide_legend(reverse = FALSE, keywidth = 1, keyheight = 1)) +
#  ylab("Relative Abundance (Genus > 1%) \n") + xlab("Sample Type") +
#  theme_minimal()+theme(panel.border = element_rect(fill=NA, colour = "black"), 
#  panel.grid.minor = element_blank(),
#  panel.grid.major = element_blank(), 
#  axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, size=8, 
#                              colour="black"), 
#  axis.text.y = element_text(size=8, colour="black"), 
#  plot.title = element_text(hjust = 0.5), 
#  axis.ticks.x = element_line(colour="#000000", size=0.1), 
#  axis.ticks.y = element_line(colour="#000000", size=0.1),
#  legend.text = element_text(size = 7)) +
#  facet_wrap(~ Voyage, scale="free")

#print(BallastSeqR_voyage_genus_bar)

#Make interactive plot
#p <- ggplotly(BallastSeqR_voyage_genus_bar)

#p
```

```{r keep genus level interactive plot}

#htmlwidgets::saveWidget(p, "R output/BallastSeqR_voyage_genus_bar_interactive.html")

#htmltools::tags$iframe(
#  src=file.path(getwd(), "R output/BallastSeqR_voyage_genus_bar_interactive.html"),
#  width="100%",
#  height="600",
#  scrolling="no",
#  seamless="seamless",
#  frameBorder="0"
#)
```

### Re-creating the above plot using ggnested

The ggnested package enables grouping of categories in the legend, which allows for genera to be grouped under class labels. The package also color-codes by main_group (in this case Class).

```{r ggnested genus}

#require("ggnested")

#Convert to a dataframe
#BallastSeqR_genus_voyage_filt <- data.frame(BallastSeqR_genus_voyage) 

#Custom color palette
#pal <- c("#440154FF", "blue", "green", "yellow", "purple", "orange", "#FF6699")
#names(pal) <- unique(BallastSeqR_genus_voyage_filt$Class)

#Use ggnested to create the bar plot
#BallastSeqR_voyage_genus_bar_ggnested <- ggnested::ggnested(BallastSeqR_genus_voyage_filt, 
#         aes(x = Sample_type, 
#             y = Mean, 
#             main_group = Class, 
#             sub_group = Genus),
#          main_palette = pal) + 
#  geom_bar(stat="identity", colour = "black", size=0.3, position = "fill") +
#  facet_wrap(~ Voyage, scale="free") +
#  guides(fill = guide_legend(reverse = FALSE, keywidth = 1, keyheight = 1)) +
#  ylab("Relative Abundance (Genus > 1%) \n") + xlab("Sample Type") +
#  theme_minimal()+theme(
#    strip.text = element_text(size=12),
#    panel.border = element_rect(fill=NA, colour = "black"), 
#    panel.grid.minor = element_blank(),
#    panel.grid.major = element_blank(), 
#    axis.title.x = element_text(size=12),
#    axis.title.y = element_text(size=12),
#    axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, size=10, 
#                              colour="black"), 
#    axis.text.y = element_text(size=10, colour="black"), 
#    plot.title = element_text(hjust = 0.5), 
#    axis.ticks.x = element_line(colour="#000000", size=0.1), 
#    axis.ticks.y = element_line(colour="#000000", size=0.1),
#    legend.text = element_text(size = 10)) +
#  facet_wrap(~ Voyage, scale="free")

#print(BallastSeqR_voyage_genus_bar_ggnested)

```

Save as svg:

```{r ggnested plot svg}

#ggsave(filename="R output/BallastSeqR_voyage_sample_genus_bar.svg", 
#       plot=BallastSeqR_voyage_genus_bar_ggnested, width=10, height=6, device=svg)
```

## Species level, grouped by voyage and sample type

```{r bcc species voyage and sample type}

#Examining BCC by sample type & voyage
BallastSeqR_species_voyage <- Ballast_physeq %>%
  tax_glom(taxrank = "Species") %>%                   #Agglomerate at species level
  transform_sample_counts(function(x) {x/sum(x)}) %>% #Transform to rel. abundance
  psmelt() %>%                                        #Melt to long format
  group_by(Sample_type, Voyage, 
           Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
  dplyr::summarize(Mean = 
                     mean(Abundance, na.rm=TRUE)) %>% #Calculate average
  arrange(Class)

```

```{r ggnested species}

#Convert to a dataframe
BallastSeqR_species_voyage_filt <- data.frame(BallastSeqR_species_voyage) 

#Custom color palette
pal <- c("#440154FF", "blue", "green", "yellow", "purple", "orange", "#FF6699",
         "#440154FF", "blue", "green", "yellow", "purple", "orange", "#FF6699",
         "#440154FF", "blue", "green", "yellow", "purple", "orange", "#FF6699",
         "#440154FF", "blue", "green", "yellow", "purple", "orange", "#FF6699",
          "#440154FF", "blue", "green", "yellow")
names(pal) <- unique(BallastSeqR_species_voyage_filt$Class)

#Use ggnested to create the bar plot
BallastSeqR_voyage_species_bar_ggnested <- ggnested::ggnested(BallastSeqR_species_voyage_filt, 
         aes(x = Sample_type, 
             y = Mean, 
             main_group = Class, 
             sub_group = Species),
          main_palette = pal) + 
  geom_bar(stat="identity", colour = "black", size=0.3, position = "fill") +
  facet_wrap(~ Voyage, scale="free") +
  guides(fill = guide_legend(reverse = FALSE, keywidth = 1, keyheight = 1)) +
  ylab("Relative Abundance (Species > 1%) \n") + xlab("Sample Type") +
  theme_minimal()+theme(
    strip.text = element_text(size=12),
    panel.border = element_rect(fill=NA, colour = "black"), 
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(), 
    axis.title.x = element_text(size=12),
    axis.title.y = element_text(size=12),
    axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, size=10, 
                              colour="black"), 
    axis.text.y = element_text(size=10, colour="black"), 
    plot.title = element_text(hjust = 0.5), 
    axis.ticks.x = element_line(colour="#000000", size=0.1), 
    axis.ticks.y = element_line(colour="#000000", size=0.1),
    legend.text = element_text(size = 10)) +
  facet_wrap(~ Voyage, scale="free")

print(BallastSeqR_voyage_species_bar_ggnested)
```

### Create plot with fantaxtic

#### Relative abundance

```{r fantaxtic class species plot}

#Merge by sample_type and voyage; there is a separate category in the metadata that combines these categories together
Ballast_physeq_merged <- speedyseq::merge_samples2(Ballast_physeq, "Type_voyage", fun_otu = mean)
SD <- speedyseq::merge_samples2(sample_data(Ballast_physeq), "Type_voyage")
print(SD[, "Type_voyage"])

#Now samples will be displayed by sample_type rather than individially when graphed using fantaxtic

#Label and order variables
sample_data(Ballast_physeq_merged)$Type_voyage <- factor(sample_data(Ballast_physeq_merged)$Type_voyage, 
               levels = c("port_uptake_1", "BWT_1", "ocean_uptake_1",
                          "BWT_BWE_1", "port_uptake_2", "BWT_2"),
               labels = c("Port uptake V1", "BWT V1", "Ocean uptake V1",
                          "BWT+BWE V1", "Port uptake V2", "BWT V2"))

#Load fantaxtic and create plot
library("fantaxtic")
                        
top_nested <- nested_top_taxa(Ballast_physeq_merged,
                              #grouping = "Type_voyage",
                              top_tax_level = "Class",
                              nested_tax_level = "Species",
                              n_top_taxa = 3, 
                              n_nested_taxa = 10)

#Order samples
order <- c("port_uptake_1", "BWT_1", "ocean_uptake_1",
                          "BWT_BWE_1", "port_uptake_2", "BWT_2")
order <- as.character(order)

#Plot
plot <- plot_nested_bar(ps_obj = top_nested$ps_obj,
                top_level = "Class",
                nested_level = "Species",
                sample_order = order) +
  facet_wrap(~ Voyage, scale="free") +
  labs(x = "Sample type", y = "Relative abundance") +
  scale_x_discrete(labels=c("Port uptake V1", "BWT V1", "Ocean uptake V1",
                          "BWT+BWE V1", "Port uptake V2", "BWT V2")) +
  theme_minimal()+theme(
    strip.text = element_text(size=12),
    panel.border = element_rect(fill=NA, colour = "black"), 
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(), 
    axis.title.x = element_text(size=12),
    axis.title.y = element_text(size=12),
    axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, size=10, 
                              colour="black"), 
    axis.text.y = element_text(size=10, colour="black"), 
    plot.title = element_text(hjust = 0.5), 
    axis.ticks.x = element_line(colour="#000000", size=0.1), 
    axis.ticks.y = element_line(colour="#000000", size=0.1),
    legend.text = element_text(size = 10))

print(plot)

#Make interactive plot
p <- ggplotly(plot)

p


```

Save as svg:

```{r species fantaxtic svg}

ggsave(filename="R output/BallastSeqR_voyage_species_bar_interactive_fantaxtic.svg", 
       plot=plot, width=10, height=6, device=svg)
```

##### Create this plot with patterns overlayed

```{r patterns and fantaxtic}

#Load ggpattern and magick packages so that patterns can be added over colors
library(ggpattern)
library(magick)

#This code prints a figure displaying possible magick patterns and their names that
#can be used in the following manual pattern scale function.
#From https://coolbutuseless.github.io/package/ggpattern/articles/pattern-magick.html

df1 <- data.frame(
  x    = rep(1:6, 9),
  y    = rep(1:9, each=6),
  name = gridpattern::names_magick,
  stringsAsFactors = FALSE
)


ggplot(df1) + 
  geom_tile_pattern(
    aes(x, y, pattern_type = I(name)),
    pattern       = 'magick',
    pattern_scale = 1.5,
    pattern_fill  = 'black', 
    width         = 0.9, 
    height        = 0.9
  ) + 
  geom_label(aes(x+0.4, y+0.4, label = name), hjust = 1, vjust = 1) + 
  theme_void() + 
  labs(
    title = "All the possible magick pattern names"
  ) +
  coord_fixed(1)
```

Create manual pattern scale function:

```{r pattern function}

#This assigns a pattern to each taxa and is used in place of 
#scale_pattern_type_manual; make sure all taxa from all data sets are included here
#so that this function can be used to create all graphs. The value 'gray100' leaves
#an assigned taxa blank.

scale_pattern_type_bacteria <- function(...){
  ggpattern:::manual_scale(
    'pattern_type',
    values = setNames(c('gray100', 'vertical', 
                        'bricks', 'gray100',
                        'gray100', 'horizontalsaw',
                        'hs_bdiagonal', 'verticalsaw',
                        'gray100', 'gray100',
                        'gray100', 'gray100',
                        'gray100', 
                        'gray100',
                        'gray100', 'gray100',
                        'gray100', 'gray100',
                        'gray100', 'gray100',
                        'gray100', 'gray100',
                        'gray100', 'hs_horizontal',
                        'gray100', 'gray100',
                        'gray100', 'gray100',
                        'gray100', 'gray100',
                        'gray100', 'gray100',
                        'gray100', 'gray100',
                        'gray100', 'gray100',
                        'gray100', 'gray100'),
                      
                      c("Oleispira Genus", "Pseudomonas Genus",
                        "Pseudomonas_anguilliseptica", "Aeromonas Genus", 
                        "Shewanella Genus", "Pseudoalteromonas Genus", 
                        "Vibrionaceae Family", "Rheinheimera Genus", 
                        "Comamonadaceae Family", "Polynucleobacter Genus",
                        "Clade_Ia Genus", "Clade_II Genus", 
                        "Caulobacter Genus", 
                        "Allorhizobium-Neorhizobium-Pararhizobium-Rhizobium Genus",
                        "Sphingorhabdus Genus", "Sphingomonadaceae Family", 
                        "Sulfitobacter Genus", "Roseovarius Genus", 
                        "Rhodobacteraceae Family", "Tistrella Genus", 
                        "uncultured_bacterium", "Mesoflavibacter_sp.", 
                        "Antarctic_bacterium", "Flavobacterium Genus", 
                        "Flavobacterium_jumunjinense", "Flavobacteriaceae Family",
                        "Fluviicola Genus", "Arcicella Genus", 
                        "Flectobacillus Genus", "Pseudarcicella Genus",
                        "Other", "Other Alphaproteobacteria",
                        "Other Bacteroidia", "Other Gammaproteobacteria")),
    
    ...
  )
}


#'verticalsaw', 'crosshatch', 'verticalrightshingle', 
 #                       'horizontal3', 'hs_diagcross', 'horizontalsaw',
  #                      'rightshingle', 'hexagons', 'circles', 'verticalbricks',
   #                     'smallfishscales', 'left45', 'right45', 
    #                    'verticalsaw',   'gray100',
     #                   'right30', 'hs_fdiagonal', 'hs_cross', 'vertical2'
```

Add patterns to plot:

```{r add patterns to fantaxtic plot}

#Plot
plot_pattern <- plot_nested_bar(ps_obj = top_nested$ps_obj,
                top_level = "Class",
                nested_level = "Species",
                sample_order = order) +
  facet_wrap(~ Voyage, scale="free") +
  labs(x = "Sample type", y = "Relative abundance") +
  scale_x_discrete(labels=c("Port uptake V1", "BWT V1", "Ocean uptake V1",
                          "BWT+BWE V1", "Port uptake V2", "BWT V2")) +
  geom_bar_pattern(position = "fill",
                   stat = "identity",
                   color = "black", 
                   pattern_fill = "black",
                   pattern_spacing = 0.05,
                   pattern = 'magick',
                   aes(pattern_type = Species)) +
  scale_pattern_type_bacteria(guide = "none") +
  
  theme_minimal()+theme(
    strip.text = element_text(size=12),
    panel.border = element_rect(fill=NA, colour = "black"), 
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(), 
    axis.title.x = element_text(size=12),
    axis.title.y = element_text(size=12),
    axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, size=10, 
                              colour="black"), 
    axis.text.y = element_text(size=10, colour="black"), 
    plot.title = element_text(hjust = 0.5), 
    axis.ticks.x = element_line(colour="#000000", size=0.1), 
    axis.ticks.y = element_line(colour="#000000", size=0.1),
    legend.text = element_text(size = 10))

print(plot_pattern)

```

Save as .svg:

```{r fantaxtic with patterns svg}

#Saving as an editable pdf
ggsave(filename="R output/BallastSeqR_fantaxtic_patterned.pdf", plot=plot_pattern, width=10, height=6, device=cairo_pdf)

```

#### Read abundance

Same plot as above, but here I'm using fantaxtic to create a plot displaying the number of reads for each of the top 10 species in the top 3 classes, rather than the relative abundance, as is shown above.

```{r total abundance fantaxtic class species}

#Order samples
order <- c("port_uptake_1", "BWT_1", "ocean_uptake_1",
                          "BWT_BWE_1", "port_uptake_2", "BWT_2")
order <- as.character(order)

#Get the top taxa
top_level <- "Class"
nested_level <- "Species"
sample_order <- order
top_asv <- nested_top_taxa(Ballast_physeq_merged,
                              top_tax_level = "Class",
                              nested_tax_level = "Species",
                              n_top_taxa = 3, 
                              n_nested_taxa = 10)


#Create names for NA taxa
ps_tmp <- top_asv$ps_obj %>%
  name_na_taxa()

#Add labels to taxa with the same names
ps_tmp <- ps_tmp %>%
  label_duplicate_taxa(tax_level = nested_level)

#Generate a palette basedon the phyloseq object
pal <- taxon_colours(ps_tmp,
                     tax_level = top_level)

#Convert physeq to df
psdf <- psmelt(ps_tmp)


#Move the merged labels to the appropriate positions in the plot:
#Top merged labels need to be at the top of the plot,
#nested merged labels at the bottom of each group
psdf <- move_label(psdf = psdf,
                   col_name = top_level,
                   label = "Other",
                   pos = 0)
psdf <- move_nested_labels(psdf,
                           top_level = top_level,
                           nested_level = nested_level,
                           top_merged_label = "Other",
                           nested_label = "Other",
                           pos = Inf)


# Reorder samples
if(!is.null(sample_order)){
  if(all(sample_order %in% unique(psdf$Sample))){
    psdf <- psdf %>%
      mutate(Sample = factor(Sample, levels = sample_order))
  } else {
    stop("Error: not all(sample_order %in% sample_names(ps_obj)).")
  }

}


#Generate a base plot
library(ggnested)
p <- ggnested(psdf,
              aes_string(main_group = top_level,
                         sub_group = nested_level,
                         x = "Sample",
                         y = "Abundance"),
              main_palette = pal) +
  facet_wrap(~ Voyage, scale="free") +
   scale_y_continuous(limits = c(0, 11000)) +
  labs(x = "Sample type", y = "Abundance") +
  scale_x_discrete(labels=c("Port uptake V1", "BWT V1", "Ocean uptake V1",
                          "BWT+BWE V1", "Port uptake V2", "BWT V2")) +
  theme_minimal()+theme(
    strip.text = element_text(size=12),
    panel.border = element_rect(fill=NA, colour = "black"), 
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(), 
    axis.title.x = element_text(size=12),
    axis.title.y = element_text(size=12),
    axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, size=10, 
                              colour="black"), 
    axis.text.y = element_text(size=10, colour="black"), 
    plot.title = element_text(hjust = 0.5), 
    axis.ticks.x = element_line(colour="#000000", size=0.1), 
    axis.ticks.y = element_line(colour="#000000", size=0.1),
    legend.text = element_text(size = 10))


#Add abundances
p <- p + geom_col()

#Use geom_col(position = position_fill()) if you want to plot the relative abundance

print(p)


#Make interactive plot
p_abundance <- ggplotly(p)

p_abundance

```

Save as svg:

```{r species fantaxtic reads svg}

ggsave(filename="R output/BallastSeqR_voyage_species_bar_interactive_fantaxtic_reads.svg", 
       plot=p, width=10, height=6, device=svg)
```

## Species level, grouped by voyage, sample type, and tank

### Create plot with fantaxtic

#### Relative abundance

```{r fantaxtic class species plot voyage sample type tank}

#Merge by sample_type and voyage; there is a separate category in the metadata that combines these categories together
Ballast_physeq_merged_group <- speedyseq::merge_samples2(Ballast_physeq, "Sample_group", fun_otu = mean)
SD <- speedyseq::merge_samples2(sample_data(Ballast_physeq), "Sample_group")
print(SD[, "Sample_group"])

#Now samples will be displayed by sample_group rather than individially when graphed using fantaxtic

#Load fantaxtic and create plot
library("fantaxtic")
                        
top_nested_group <- nested_top_taxa(Ballast_physeq_merged_group,
                              #grouping = "Type_voyage",
                              top_tax_level = "Class",
                              nested_tax_level = "Species",
                              n_top_taxa = 3, 
                              n_nested_taxa = 10)

#Order samples
order <- sample_data(top_nested_group$ps_obj) %>%
  data.frame() %>%
  arrange(Sample_group) %>%
  pull(Sample_group) %>%
  as.character()

#Order samples
order <- c("V1_UP_T5", "V1_UP_T6", "V1_BWT_T5", "V1_BWT_T6", "V1_Sea_T5",
           "V1_BWT_BWE_T5", "V2_UP_T2", "V2_UP_T6", "V2_BWT_T6")

#Plot
plot2 <- plot_nested_bar(ps_obj = top_nested_group$ps_obj,
                top_level = "Class",
                nested_level = "Species",
                sample_order = order) +
  facet_nested(~ Voyage + Sample_type, scale="free") +
  labs(x = "Sample group", y = "Relative abundance") +
  theme_minimal()+theme(
    strip.text = element_text(size=12),
    panel.border = element_rect(fill=NA, colour = "black"), 
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(), 
    axis.title.x = element_text(size=12),
    axis.title.y = element_text(size=12),
    axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, size=10, 
                              colour="black"), 
    axis.text.y = element_text(size=10, colour="black"), 
    plot.title = element_text(hjust = 0.5), 
    axis.ticks.x = element_line(colour="#000000", size=0.1), 
    axis.ticks.y = element_line(colour="#000000", size=0.1),
    legend.text = element_text(size = 10))

```

# Bacterial community composition dot plots

## Genus level, grouped by sample type and voyage

```{r genus dot plot}

#Calculate relative abundance at the genus level for each sample type and voyage
BallastSeqR_genus_voyage <- Ballast_physeq %>%
  tax_glom(taxrank = "Genus") %>%                     #Agglomerate at genus level
  transform_sample_counts(function(x) {x/sum(x)}) %>% #Transform to rel. abundance
  psmelt() %>%                                        #Melt to long format
  group_by(Sample_type, Voyage, 
           Kingdom, Phylum, Class, Order, Family, Genus) %>%
  dplyr::summarize(Mean = 
                     mean(Abundance, na.rm=TRUE),
                    .groups = "keep") %>%             #Calculate average
  filter(Mean > 0.01) %>%                             #Filter 
  arrange(Class)

#Create color palette
genus_col = colorRampPalette(brewer.pal(9,"Dark2"))(12)

#Create plot
genus_dot <- ggplot(BallastSeqR_genus_voyage, aes(x=Voyage, y=Genus, 
                                                  color=Class)) +  
  geom_point(aes(size=Mean), alpha = 1/2) +
  scale_color_manual(values = genus_col) +
  facet_nested( ~ Sample_type, scales = "free", 
                nest_line = element_line(colour = "black")) +
  theme_minimal()+
  theme(panel.border = element_rect(fill=NA, colour = "black"), 
        panel.grid.minor = element_blank(), 
        panel.grid.major = element_blank(), 
        axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, 
                                    size=8, colour="black"), 
        axis.text.y = element_text(size=8, colour="black"), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x = element_line(colour="#000000", linewidth = 0.1), 
        axis.ticks.y = element_line(colour="#000000", linewidth = 0.1),
        legend.text = element_text(size = 7),
        ggh4x.facet.nestline = element_line(linetype = "solid"))

print(genus_dot)
```

## Genus level, grouped by voyage and sample type

Same plot as above, but voyage is grouped first and sample type is subset under it.

```{r genus dot plot 2}

#Create plot
genus_dot_2 <- ggplot(BallastSeqR_genus_voyage, aes(x=Sample_type, y=Genus, 
                                                  color=Class)) +  
  geom_point(aes(size=Mean), alpha = 1/2) +
  scale_color_manual(values = genus_col) +
  facet_nested( ~ Voyage, scales = "free", 
                nest_line = element_line(colour = "black")) +
  theme_minimal()+
  theme(panel.border = element_rect(fill=NA, colour = "black"), 
        panel.grid.minor = element_blank(), 
        panel.grid.major = element_blank(), 
        axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, 
                                    size=8, colour="black"), 
        axis.text.y = element_text(size=8, colour="black"), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x = element_line(colour="#000000", linewidth = 0.1), 
        axis.ticks.y = element_line(colour="#000000", linewidth = 0.1),
        legend.text = element_text(size = 7),
        ggh4x.facet.nestline = element_line(linetype = "solid"))

print(genus_dot_2)
```

## Species level, grouped by sample type and voyage

Same as the first genus level plot, but displaying the abundance of species present in abundance \>1%.

```{r species dot plot}

#Calculate relative abundance at the genus level for each sample type and voyage
BallastSeqR_species_voyage <- Ballast_physeq %>%
  tax_glom(taxrank = "Species") %>%                   #Agglomerate at species level
  transform_sample_counts(function(x) {x/sum(x)}) %>% #Transform to rel. abundance
  psmelt() %>%                                        #Melt to long format
  group_by(Sample_type, Voyage, 
           Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
  dplyr::summarize(Mean = 
                     mean(Abundance, na.rm=TRUE),
                    .groups = "keep") %>%             #Calculate average
  filter(Mean > 0.01) %>%                             #Filter out <1% 
  arrange(Class)

#Create color palette
species_col = colorRampPalette(brewer.pal(9,"Dark2"))(12)

#Create plot
species_dot <- ggplot(BallastSeqR_species_voyage, aes(x=Voyage, y=Species, 
                                                  color=Class)) +  
  geom_point(aes(size=Mean), alpha = 1/2) +
  scale_color_manual(values = species_col) +
  facet_nested( ~ Sample_type, scales = "free", 
                nest_line = element_line(colour = "black")) +
  theme_minimal()+
  theme(panel.border = element_rect(fill=NA, colour = "black"), 
        panel.grid.minor = element_blank(), 
        panel.grid.major = element_blank(), 
        axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, 
                                    size=8, colour="black"), 
        axis.text.y = element_text(size=8, colour="black"), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x = element_line(colour="#000000", linewidth = 0.1), 
        axis.ticks.y = element_line(colour="#000000", linewidth = 0.1),
        legend.text = element_text(size = 7),
        ggh4x.facet.nestline = element_line(linetype = "solid"))

print(species_dot)

```

## Species level, grouped by voyage and sample type

Same as the second genus level plot, but displaying the abundance of species present in abundance \>1%.

```{r species dot plot 2}

#Create plot
species_dot_2 <- ggplot(BallastSeqR_species_voyage, aes(x=Sample_type, y=Species, 
                                                  color=Class)) +  
  geom_point(aes(size=Mean), alpha = 1/2) +
  scale_color_manual(values = species_col) +
  facet_nested( ~ Voyage, scales = "free", 
                nest_line = element_line(colour = "black")) +
  theme_minimal()+
  theme(panel.border = element_rect(fill=NA, colour = "black"), 
        panel.grid.minor = element_blank(), 
        panel.grid.major = element_blank(), 
        axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, 
                                    size=8, colour="black"), 
        axis.text.y = element_text(size=8, colour="black"), 
        plot.title = element_text(hjust = 0.5), 
        axis.ticks.x = element_line(colour="#000000", linewidth = 0.1), 
        axis.ticks.y = element_line(colour="#000000", linewidth = 0.1),
        legend.text = element_text(size = 7),
        ggh4x.facet.nestline = element_line(linetype = "solid"))

print(species_dot_2)


```

# DESeq2 analysis on MetaCyc data

## Create the DESeq object

```{r MetaCyc DESeq object}

#Load metadata as a tibble
picrust_metadata <- read_delim("PICRUSt2-analysis/PICRUSt-Metadata.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE) 

#Load metacyc pathway abundance as a matrix
ballast_metacyc_abundance <- read_delim("PICRUSt2-analysis/MetaCyc_version/pathways_out/path_abun_unstrat_edited.tsv", delim = "\t")

#Check to see if metadata and pathway abundance samples are in the same order
head(ballast_metacyc_abundance)

head(picrust_metadata)

#In the picrust_metadata, make the sample-id column the rownames
picrust_metadata <- column_to_rownames(picrust_metadata, var = "sample-id")

#In the pathway abundance matrix, make the pathway column the rownames
ballast_metacyc_abundance <- column_to_rownames(ballast_metacyc_abundance, var = "pathway")

#The order of the rownames in the metadata file and the column names in the pathway abundance file don't match; they need to be in the same order before we can make this a DESeq object, because DESeq will assume that the sample order is consistent between the two
all(rownames(picrust_metadata) == colnames(ballast_metacyc_abundance))
ballast_metacyc_abundance <- ballast_metacyc_abundance[, rownames(picrust_metadata)]
all(rownames(picrust_metadata) == colnames(ballast_metacyc_abundance))

#Make Voyage & Sample_type factors
picrust_metadata$Voyage <- factor(picrust_metadata$Voyage, 
               levels = c("1", "2"),
               labels = c("Voyage 1", "Voyage 2"))

picrust_metadata$Sample_type <- factor(picrust_metadata$Sample_type, 
               levels = c("port_uptake", "ocean_uptake", "BWT", "BWT_BWE"),
               labels = c("Port uptake", "Ocean uptake", "BWT", "BWT+BWE"))

#Create a DESeq object from the MetaCyc count data; using both Voyage and Sample_type in the design formula in order to get a clearer picture of the differences in function that can be attributed to the different sample types
ballast_deseq <- DESeqDataSetFromMatrix(countData = round(ballast_metacyc_abundance),
                              colData = picrust_metadata, 
                              design = ~ Voyage + Sample_type)
ballast_deseq


#Make sure that the control is the reference level so that the default log2fold changes are calculated as treatment vs. control instead of vice versa
ballast_deseq$Sample_type <- relevel(ballast_deseq$Sample_type, ref = "Port uptake")

#May add this part in later if I find that I have technical replicates: Collapse technical replicates (i.e., multiple sequencing runs of the same library) using the collapseReplicates function?

```

## Load pathway descriptions

Load dataframes containing pathway descriptions; this will be used to add descriptions to the pathways identified by DESeq2 later. The path_abun_unstrat_descrip file, which contains the lowest level pathway description, was exported from PICRUSt2. The files in the upperlevel_descrip folder contain higher level descriptions of the pathways and were downloaded from <https://github.com/Jiung-Wen/picrust_mapping/tree/master>

The upper level descriptions include the top levels of the MetaCyc pathways, while the second level file contains the sub-classes of this top level.

```{r load pathway descriptions}

#Load pathways descriptions as a matrix
ballast_metacyc_descriptions <- read_delim("PICRUSt2-analysis/MetaCyc_version/pathways_out/path_abun_unstrat_descrip.tsv", delim = "\t")

#Keep just the pathway and descritpions columns in this dataframe
ballast_metacyc_descriptions <- dplyr::select(ballast_metacyc_descriptions, c(pathway, description))

#Load upper level MetaCyc pathway descriptions; add column names that match the descriptions dataframe above
ballast_metacyc_second_level <- read_delim("PICRUSt2-analysis/MetaCyc_version/pathways_out/upperlevel_descrip/metacyc_sec.tsv", delim = "\t", col_names = c("pathway", "second_level"))

ballast_metacyc_top_level <- read_delim("PICRUSt2-analysis/MetaCyc_version/pathways_out/upperlevel_descrip/metacyc_top.tsv", delim = "\t", col_names = c("pathway", "top_level"))

#Combine upper level descriptions with the description pathways dataframe
ballast_metacyc_descriptions2 <- ballast_metacyc_descriptions %>%
  left_join(ballast_metacyc_second_level, by = "pathway")  

ballast_metacyc_all_levels <- ballast_metacyc_descriptions2 %>%
  left_join(ballast_metacyc_top_level, by = "pathway")  
```

## Create PCA of all MetaCyc results in DESeq2

### PCA with labels and arrows

```{r MetaCyc PCA labels}

#Transforming the DESeq data using a regularized log transformation
rld <- rlog(ballast_deseq, blind=TRUE)

#Plot the PCA (Principle Component Analysis, not to be confused with PCoA, Principle Coordinate Analysis)
plotPCA(rld, intgroup=c("Voyage", "Sample_type", "Sample_number"))

#Customize the PCA plot using the ggplot function
pcaData <- plotPCA(rld, intgroup=c("Voyage", "Sample_type", "Sample_number"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(PC1, PC2, color=Sample_type, shape=Voyage)) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() +
  geom_point(mapping = aes(colour = factor(Sample_type), shape = factor(Voyage), size = 3)) +
  geom_text(aes(label = Sample_number, colour = factor(Sample_type)), nudge_x=0.25, nudge_y=0.25, check_overlap = TRUE) +
  scale_colour_manual(values = c("Port uptake" = "#E7298A", "Ocean uptake" = "#7570B3", "BWT" = "#1B9E77", "BWT+BWE" = "#D95F02"), "Sample Type") +
  scale_shape_manual(values = c("Voyage 1" = 16, "Voyage 2" = 17), name = "Voyage") +
  guides(size=FALSE) +
  guides(shape = guide_legend(override.aes = list(size = 3))) +
  theme(plot.title = element_text(size = 18),
        text = element_text(size = 18), 
        axis.title = element_text(size = 15),
        panel.spacing = unit(1, "lines"), 
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5), 
        panel.background = element_blank(), 
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15),
        legend.justification = c("right", "top")) +
  geom_segment(
  aes(x = -2.1, y = -2.4, xend = 1.5, yend = -1.0),
  arrow = arrow(length = unit(0.03, "npc"))) +
   geom_segment(
  aes(x = -1.7, y = 1.9, xend = 0.4, yend = 1.4),
  arrow = arrow(length = unit(0.03, "npc")), colour = "#E41A1C") +
   geom_segment(
  aes(x = -0.8, y = 0.40, xend = 3.2, yend = -0.5),
  arrow = arrow(length = unit(0.03, "npc"))) +
  geom_curve(
  aes(x = -1.5, y = 3.45, xend = 1.5, yend = 1.6),
  arrow = arrow(length = unit(0.03, "npc")), curvature = -0.3) +
  geom_segment(
  aes(x = 3.3, y = -0.4, xend = 0.90, yend = 1.2),
  arrow = arrow(length = unit(0.03, "npc")), colour = "#E41A1C") 
```

### PCA without labels and arrows

```{r MetaCyc PCA no labels}

#Transforming the DESeq data using a regularized log transformation
rld <- rlog(ballast_deseq, blind=TRUE)

#Plot the PCA (Principle Component Analysis, not to be confused with PCoA, Principle Coordinate Analysis)
plotPCA(rld, intgroup=c("Voyage", "Sample_type"))

#Customize the PCA plot using the ggplot function
pcaData <- plotPCA(rld, intgroup=c("Voyage", "Sample_type"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(PC1, PC2, color=Sample_type, shape=Voyage)) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() +
  geom_point(mapping = aes(colour = factor(Sample_type), shape = factor(Voyage), size = 3)) +
  scale_colour_manual(values = c("Port uptake" = "#E7298A", "Ocean uptake" = "#7570B3", "BWT" = "#1B9E77", "BWT+BWE" = "#D95F02"), "Sample Type") +
  scale_shape_manual(values = c("Voyage 1" = 16, "Voyage 2" = 17), name = "Voyage") +
  guides(size=FALSE) +
  guides(shape = guide_legend(override.aes = list(size = 3))) +
  theme(plot.title = element_text(size = 18),
        text = element_text(size = 18), 
        axis.title = element_text(size = 15),
        panel.spacing = unit(1, "lines"), 
        panel.border = element_rect(colour = "black", fill = NA, size = 0.5), 
        panel.background = element_blank(), 
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15),
        legend.justification = c("right", "top"))
```

## BWT vs port uptake

### Based on lowest level of functional annotation

```{r DESeq2 MetaCyc}

#Run DESeq
ballast_deseq <- DESeq(ballast_deseq, test="Wald")

#Show results
res_ballast_deseq <- results(ballast_deseq)
res_ballast_deseq

#Check which comparisons are contained in the deseq object:
resultsNames(ballast_deseq)

#Just want to compare Port_uptake with BWT; creating a results table with only this information
res_ballast_deseq_BWT <- results(ballast_deseq, contrast=c("Sample_type","BWT","Port uptake"), alpha=0.05)

#Show results summary
summary(res_ballast_deseq_BWT)

#Log fold change shrinkage on the effluent_TWW_vs_CON comparison
resLFC <- lfcShrink(ballast_deseq, coef="Sample_type_BWT_vs_Port.uptake", type="apeglm")
resLFC

#"It is more useful visualize the MA-plot for the shrunken log2 fold changes, 
#which remove the noise associated with log2 fold changes from low count genes 
#without requiring arbitrary filtering thresholds."
#https://www.bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#why-un-normalized-counts
plotMA(resLFC)

#Convert results to dataframe
df_ballast_deseq_BWT <- as.data.frame(res_ballast_deseq_BWT)

#Filter based on adj p-value, removing non-significant values (p>0.05). This also removes p-values that are NA; NA p-values exist because when p-values do not pass the filter threshold (p-values less than 0.1) they are set to NA. More information in vignette("DESeq2") under "Note on p-values set to NA"
df_ballast_deseq_BWT_filt <- filter(df_ballast_deseq_BWT, padj <= 0.05)

#Order by adjusted p-value
df_ballast_deseq_BWT_filt = df_ballast_deseq_BWT_filt[order(df_ballast_deseq_BWT_filt$padj, na.last=NA), ]

#Add pathway descriptions to the DESeq results using the left_join function in dplyr; rownames in the DESeq output need to be a column labeled pathway in order to match that of the descriptions dataframe
df_ballast_deseq_BWT_filt <- rownames_to_column(df_ballast_deseq_BWT_filt, var = "pathway")

df_ballast_deseq_BWT_filt_descrip <- df_ballast_deseq_BWT_filt %>%
  left_join(ballast_metacyc_all_levels, by = "pathway")

```

#### Create dot plot of results - log2fold change, colored by top_level

```{r deseq dot plot colors}

#Create color scheme
fold_change_col = colorRampPalette(brewer.pal(9,"Dark2"))(12)
```

```{r log2fold change}

#Plot
log2foldplot <- ggplot(df_ballast_deseq_BWT_filt_descrip, 
                       aes(x=log2FoldChange, y=description, color=top_level)) + geom_point(size=6) +
  scale_color_manual(values = fold_change_col) +
  geom_vline(xintercept = 0, linetype="dashed", color = "black", size=1) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) 

print(log2foldplot)

```

```{r save log2fold change dot plot}

#Save as an svg file
ggsave(filename="R output/Ballast_l2fplot.svg", plot=log2foldplot, width=12, height=20, device=svg)

```

#### Create dot plot of results - log2fold change, colored by second_level, faceted by second_level

```{r deseq dot plot extra colors}

#Create color scheme
fold_change_col = colorRampPalette(brewer.pal(9,"Dark2"))(12)

plot_col_extra <- c("black", "#440154FF", "#450659FF","#460B5EFF","#472D7AFF","#3B518BFF", "#3A548CFF", "#38598CFF", "#365C8DFF", "#34608DFF", "#33638DFF", "#31678EFF", "#306A8EFF","#2E6E8EFF", "#2D718EFF", "#2B748EFF","#2A778EFF", "#297B8EFF","#1F958BFF", "#1F988BFF", "#1E9C89FF", "#1F9F88FF","#1FA287FF", "#21A585FF", "#23A983FF","#25AC82FF", "#29AF7FFF", "#2DB27DFF","#32B67AFF", "#37B878FF", "#3CBC74FF","#57C766FF", "#5EC962FF","#A2DA37FF","#DAE319FF", "#E4E419FF",  "#ECE51BFF", "#F5E61FFF","darkorchid1", "darkorchid2", "darkorchid3","#A21C9AFF", "#A62098FF", "#AB2394FF", "#AE2892FF", "#B22B8FFF", "#B6308BFF", "#BA3388FF", "#BE3885FF", "#C13B82FF", "#C53F7EFF","#C8437BFF", "#CC4678FF", "#CE4B75FF", "#D14E72FF", "#D5536FFF", "#D7566CFF", "#DA5B69FF", "#DD5E66FF", "#E06363FF","#FF6699","#F68D45FF", "#FCA537FF","#F6E726FF", "#F4ED27FF", "#00489C", "#CCCCCC", "#999999", "#A1C299","#300018")
```

```{r log2fold change 2}

#Plot
log2foldplot <- ggplot(df_ballast_deseq_BWT_filt_descrip, 
                       aes(x=log2FoldChange, y=description, color=second_level)) + geom_point(size=6) +
  scale_color_manual(values = plot_col_extra) +
  geom_vline(xintercept = 0, linetype="dashed", color = "black", size=1) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
facet_wrap(~top_level, scales = "free") 


print(log2foldplot)
```

#### Create dot plot of results - log2fold change, separate graphs for each top_level

##### Subset data by top level

```{r log2fold subset top level}

df_ballast_deseq_BWT_filt_Biosynthesis <- df_ballast_deseq_BWT_filt_descrip %>%
  subset(top_level == "Biosynthesis")

df_ballast_deseq_BWT_filt_Degradation <- df_ballast_deseq_BWT_filt_descrip %>%
  subset(top_level == "Degradation/Utilization/Assimilation")

df_ballast_deseq_BWT_filt_Energy <- df_ballast_deseq_BWT_filt_descrip %>%
  subset(top_level == "Generation of Precursor Metabolite and Energy")

df_ballast_deseq_BWT_filt_Glycan <- df_ballast_deseq_BWT_filt_descrip %>%
  subset(top_level == "Glycan Pathways")	

df_ballast_deseq_BWT_filt_Macro <- df_ballast_deseq_BWT_filt_descrip %>%
  subset(top_level == "Macromolecule Modification")

df_ballast_deseq_BWT_filt_Super <- df_ballast_deseq_BWT_filt_descrip %>%
  subset(top_level == "Superpathways")

```

##### Plot Biosynthesis pathways

```{r plot biosynthesis log2fold}

#Plot Biosynthesis log2fold changes
log2foldplot_Biosynthesis <- ggplot(df_ballast_deseq_BWT_filt_Biosynthesis, 
                       aes(x=log2FoldChange, y=description, color=second_level)) + geom_point(size=6) +
  scale_color_manual(values = fold_change_col) +
  geom_vline(xintercept = 0, linetype="dashed", color = "black", size=1) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())  


print(log2foldplot_Biosynthesis)
```

```{r biosynthesis log2fold svg}

ggsave(filename="R output/Metacyc_pathways_log2fold_Biosynthesis.svg", 
       plot=log2foldplot_Biosynthesis, width=20, height=10, device=svg)
```

##### Plot Degradation/Utilization/Assimilation pathways

```{r plot degradation log2fold}

#Plot Biosynthesis log2fold changes
log2foldplot_Degradation <- ggplot(df_ballast_deseq_BWT_filt_Degradation, 
                       aes(x=log2FoldChange, y=description, color=second_level)) + geom_point(size=6) +
  scale_color_manual(values = fold_change_col) +
  geom_vline(xintercept = 0, linetype="dashed", color = "black", size=1) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())  


print(log2foldplot_Degradation)
```

```{r degradation log2fold svg}

ggsave(filename="R output/Metacyc_pathways_log2fold_Degradation.svg", 
       plot=log2foldplot_Degradation, width=15, height=8, device=svg)
```

##### Plot Generation of Precursor Metabolite and Energy pathways

```{r plot metabolite and energy log2fold}

#Plot Biosynthesis log2fold changes
log2foldplot_Energy <- ggplot(df_ballast_deseq_BWT_filt_Energy, 
                       aes(x=log2FoldChange, y=description, color=second_level)) + geom_point(size=6) +
  scale_color_manual(values = fold_change_col) +
  geom_vline(xintercept = 0, linetype="dashed", color = "black", size=1) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())  


print(log2foldplot_Energy)
```

```{r energy log2fold svg}

ggsave(filename="R output/Metacyc_pathways_log2fold_Energy.svg", 
       plot=log2foldplot_Energy, width=10, height=5, device=svg)
```

##### Plot Glycan pathways

```{r plot glycan log2fold}

#Plot Biosynthesis log2fold changes
log2foldplot_Glycan <- ggplot(df_ballast_deseq_BWT_filt_Glycan, 
                       aes(x=log2FoldChange, y=description, color=second_level)) + geom_point(size=6) +
  scale_color_manual(values = fold_change_col) +
  geom_vline(xintercept = 0, linetype="dashed", color = "black", size=1) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())  


print(log2foldplot_Glycan)
```

```{r glycan log2fold svg}

ggsave(filename="R output/Metacyc_pathways_log2fold_Glycan.svg", 
       plot=log2foldplot_Glycan, width=10, height=5, device=svg)
```

##### Plot Macromolecule Modification pathways

```{r plot macro log2fold}

#Plot Macromolecule Medification log2fold changes
log2foldplot_Macro <- ggplot(df_ballast_deseq_BWT_filt_Macro, 
                       aes(x=log2FoldChange, y=description, color=second_level)) + geom_point(size=6) +
  scale_color_manual(values = fold_change_col) +
  geom_vline(xintercept = 0, linetype="dashed", color = "black", size=1) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())  


print(log2foldplot_Macro)
```

```{r macro log2fold svg}

ggsave(filename="R output/Metacyc_pathways_log2fold_Macro.svg", 
       plot=log2foldplot_Macro, width=10, height=5, device=svg)
```

##### Plot Superpathways

```{r plot superpathways log2fold}

#Plot Superpathway log2fold changes
log2foldplot_Super <- ggplot(df_ballast_deseq_BWT_filt_Super, 
                       aes(x=log2FoldChange, y=description, color=second_level)) + geom_point(size=6) +
  scale_color_manual(values = fold_change_col) +
  geom_vline(xintercept = 0, linetype="dashed", color = "black", size=1) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())  


print(log2foldplot_Super)
```

```{r super log2fold svg}

ggsave(filename="R output/Metacyc_pathways_log2fold_Super.svg", 
       plot=log2foldplot_Super, width=10, height=5, device=svg)
```

#### Bar plot of results - normalized counts

```{r find normalized counts}

#Find normalized counts
port_BWT_deseq_norm_counts <- as.data.frame(counts(ballast_deseq, 
                                                   normalize = TRUE))

#Make the rownames a column
sig_norm_counts <- tibble::rownames_to_column(port_BWT_deseq_norm_counts, var = "pathway")

#Gathering the normalized counts into 2 columns
sig_norm_counts_a <- gather(sig_norm_counts, key = "samplename",
                                value = "normalized_counts", -pathway)

#Create a column with the filtered deseq dataframe rownames as well
pathway_all <- data.frame(df_ballast_deseq_BWT_filt)

#Add normalized counts to the deseq dataframe
sig_norm_counts_filt <- pathway_all %>%
  left_join(sig_norm_counts_a, by = "pathway") %>%
  filter(pathway == pathway) 

#Create dataframe out of the phyloseq metadata so that we can ID these samples
meta <- data.frame(sample_data(Ballast_physeq)) %>%
  tibble::rownames_to_column(var = "samplename")

#Merge metadata with sig_norm_counts_filt
sig_norm_counts_filt <- sig_norm_counts_filt %>%
  left_join(meta, by = "samplename")

#Filter: Remove deseq results columns that aren't relevant for the following graph,
#and filter the Sample_type column so that the dataset includes only the 
#samples contrasted in deseq
sig_norm_counts_filt <- sig_norm_counts_filt %>% 
  dplyr::select(-baseMean, -lfcSE, -stat, -pvalue, -padj) %>%
  filter(Sample_type %in% c("Port uptake", "BWT"))

#Add 1 to account for 0's and then log transform the count data
sig_norm_counts_filt$log_normalized_counts <- log(sig_norm_counts_filt$normalized_counts+1)
 
#Add pathway descriptions
sig_norm_counts_filt <- sig_norm_counts_filt %>%
  left_join(ballast_metacyc_all_levels, by = "pathway")

#Remove pathways that have a negative log2FoldChange
sig_norm_counts_filt_pos <- filter(sig_norm_counts_filt, log2FoldChange > 0)


```

Examine pathways that increased in abundance in BWT samples:

```{r normalized counts bar}

#Barplot version
BWT_positive_barplot <- ggplot(sig_norm_counts_filt_pos, aes(x = description, y = normalized_counts, fill = Sample_type)) +
   stat_summary(geom = "bar", fun = mean, position = "dodge", colour = "black") +
   stat_summary(geom = "errorbar", fun.data = mean_se, position = "dodge") +
   xlab("description") +
   ylab("Normalized counts") +
   theme_bw() +
   theme(axis.text.x = element_text(angle = 45, hjust = 1),
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank()) 
   
 #facet_grid(. ~ Class, scales = "free") 
 
 print(BWT_positive_barplot)
```

#### Dot plot of results - average normalized counts

##### Grouped by port uptake and BWT

```{r counts dot plot}

#Calculate average
sig_norm_counts_filt_pos_avg <- sig_norm_counts_filt_pos %>% 
  group_by(pathway, Sample_type, description, second_level, top_level) %>%
    dplyr::summarize(Mean = mean(normalized_counts, na.rm=TRUE)) 
    
#Make discrete categories to match with the color palette
sig_norm_counts_filt_pos_avg$Mean_factor <- cut(sig_norm_counts_filt_pos_avg$Mean, breaks = c(20, 1000, 2000, 3000, 4000, 5000, 6000, 7000, 8000, 9000, 10000), labels = c("0-1000", "1000-2000", "2000-3000", "3000-4000", "4000-5000", "5000-6000", "6000-7000", "7000-8000", "8000-9000", "9000-10000"))

#Order the new variable
sig_norm_counts_filt_pos_avg$Mean_factor <- 
  factor(sig_norm_counts_filt_pos_avg$Mean_factor, 
               levels = c("9000-10000", "8000-9000", "7000-8000", "6000-7000",
                          "5000-6000", "4000-5000", "3000-4000", "2000-3000", 
                          "1000-2000", "0-1000"))

#Plot
pathway_avg_count_plot <- ggplot(sig_norm_counts_filt_pos_avg, 
                       aes(x=Sample_type, y=description, color=Mean_factor)) + 
  geom_point(size=6) +
  scale_color_brewer(palette = "RdYlBu") + #color-blind friendly palette
  theme_bw() +
  theme(panel.grid.minor = element_blank()) 

print(pathway_avg_count_plot)
```

##### Grouped by sample type and voyage

```{r counts dot plot voyage}

#Calculate average
sig_norm_counts_filt_pos_avg_voyage <- sig_norm_counts_filt_pos %>% 
  group_by(pathway, Sample_type, Voyage, description, second_level, top_level) %>%
    dplyr::summarize(Mean = mean(normalized_counts, na.rm=TRUE)) %>%
    arrange(second_level)
    
#Make discrete categories to match with the color palette
sig_norm_counts_filt_pos_avg_voyage$Mean_factor <- cut(sig_norm_counts_filt_pos_avg_voyage$Mean, breaks = c(-1, 1000, 2000, 3000, 4000, 5000, 6000, 7000, 8000, 9000, 10000), labels = c("0-1000", "1000-2000", "2000-3000", "3000-4000", "4000-5000", "5000-6000", "6000-7000", "7000-8000", "8000-9000", "9000-10000"))

#Order the new variable
sig_norm_counts_filt_pos_avg_voyage$Mean_factor <- 
  factor(sig_norm_counts_filt_pos_avg_voyage$Mean_factor, 
               levels = c("9000-10000", "8000-9000", "7000-8000", "6000-7000",
                          "5000-6000", "4000-5000", "3000-4000", "2000-3000", 
                          "1000-2000", "0-1000"))

#Order factors
sig_norm_counts_filt_pos_avg_voyage$second_level <- 
  factor(sig_norm_counts_filt_pos_avg_voyage$second_level, 
                      levels = c("Amine and Polyamine Biosynthesis", 
                                 "Amine and Polyamine Degradation", 
                                 "Amino Acid Biosynthesis", 
                                 "Amino Acid Degradation", 
                                 "Aromatic Compound Biosynthesis", 
                                 "Aromatic Compound Degradation", 
                                 "Carbohydrate Biosynthesis", 
                                 "Carbohydrate Degradation", "Cofactor, Prosthetic Group, Electron Carrier, and Vitamin Biosynthesis", "Degradation/Utilization/Assimilation - Other", "Fatty Acid and Lipid Biosynthesis", "Fatty Acid and Lipid Degradation", "Fermentation", "Inorganic Nutrient Metabolism", "Metabolic Regulator Biosynthesis", "Nucleoside and Nucleotide Degradation", "Secondary Metabolite Biosynthesis", "Secondary Metabolite Degradation", "TCA cycle", "glyoxylate cycle")) 

#Plot
pathway_avg_count_voyage_plot <- ggplot(sig_norm_counts_filt_pos_avg_voyage, 
                       aes(x=Sample_type, y = description,
                           color=Mean_factor)) + 
  geom_point(size=4) +
  scale_color_brewer(palette = "RdYlBu") + #color-blind friendly palette
  facet_nested(second_level ~ Voyage, scales = "free_y", space = "free") +
  labs(x = "Sample type", y = "Pathway", color = "Normalized abundance") +
  theme_minimal()+theme(
    strip.text = element_text(size=12),
    panel.border = element_rect(fill=NA, colour = "black"), 
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(), 
    axis.title.x = element_text(size=12),
    axis.title.y = element_text(size=12),
    axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, size=10, 
                              colour="black"), 
    axis.text.y = element_text(size=10, colour="black"), 
    plot.title = element_text(hjust = 0.5), 
    axis.ticks.x = element_line(colour="#000000", size=0.1), 
    axis.ticks.y = element_line(colour="#000000", size=0.1),
    legend.text = element_text(size = 10), 
    strip.text.y = element_text(angle = 0))


```

##### Grouped by sample type and voyage - \>1 log2foldchange only

```{r counts dot plot voyage greater than 1}

#Remove pathways that have a log2FoldChange <1
sig_norm_counts_filt_1 <- filter(sig_norm_counts_filt, log2FoldChange >= 1)

#Calculate average
sig_norm_counts_filt_1_avg <- sig_norm_counts_filt_1 %>% 
  group_by(pathway, Sample_type, Voyage, description, second_level, top_level) %>%
    dplyr::summarize(Mean = mean(normalized_counts, na.rm=TRUE)) %>%
  arrange(second_level)

#Make discrete categories to match with the color palette
sig_norm_counts_filt_1_avg$Mean_factor <- cut(sig_norm_counts_filt_1_avg$Mean, breaks = c(-1, 1000, 2000, 3000, 4000, 5000, 6000), labels = c("0-1000", "1000-2000", "2000-3000", "3000-4000", "4000-5000", "5000-6000"))

#Order the new variable
sig_norm_counts_filt_1_avg$Mean_factor <- 
  factor(sig_norm_counts_filt_1_avg$Mean_factor, 
               levels = c("5000-6000", "4000-5000", "3000-4000", "2000-3000", 
                          "1000-2000", "0-1000"))

#Order factors
sig_norm_counts_filt_1_avg$second_level <- 
  factor(sig_norm_counts_filt_1_avg$second_level, 
                      levels = c("Amine and Polyamine Degradation",  
                                 "Amino Acid Degradation",  
                                 "Aromatic Compound Degradation", 
                                 "Carbohydrate Degradation", 
                                 "Cofactor, Prosthetic Group, Electron Carrier, and Vitamin Biosynthesis", "Degradation/Utilization/Assimilation - Other", "Fermentation", "Inorganic Nutrient Metabolism", "Secondary Metabolite Degradation", "TCA cycle")) 

#Plot
pathway_avg_1_voyage_plot <- ggplot(sig_norm_counts_filt_1_avg, 
                       aes(x=Sample_type, y = description,
                           color=Mean_factor)) + 
  geom_point(aes(size=Mean)) +
  scale_color_brewer(palette = "RdYlBu") + #color-blind friendly palette
  facet_nested(second_level ~ Voyage, scales = "free_y", space = "free") +
  labs(x = "Sample type", y = "Pathway", color = "Normalized abundance") +
  theme_minimal()+theme(
    strip.text = element_text(size=12),
    panel.border = element_rect(fill=NA, colour = "black"), 
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(), 
    axis.title.x = element_text(size=12),
    axis.title.y = element_text(size=12),
    axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, size=10, 
                              colour="black"), 
    axis.text.y = element_text(size=10, colour="black"), 
    plot.title = element_text(hjust=0.5), 
    axis.ticks.x = element_line(colour="#000000", size=0.1), 
    axis.ticks.y = element_line(colour="#000000", size=0.1),
    legend.text = element_text(size=10), 
    strip.text.y = element_text(angle=0, size=10))

```

### Based on second level of functional annotation

Need to group by second level annotation

```{r metacyc deseq object second level}


#Group by second_level
ballast_deseq_BWT_second <- df_ballast_deseq_BWT_filt_descrip %>% 
  group_by(top_level, second_level) %>%
  dplyr::summarize(log2FoldChange = mean(log2FoldChange, na.rm=TRUE)) %>%
  arrange(top_level)



#Plot
log2foldplot <- ggplot(ballast_deseq_BWT_second, 
                       aes(x=log2FoldChange, y=second_level, color=top_level)) + geom_point(size=6) +
  scale_color_manual(values = fold_change_col) +
  geom_vline(xintercept = 0, linetype="dashed", color = "black", size=1) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) 

print(log2foldplot)
```

## BWT vs BWT+BWE

```{r DeSeq2 MetaCyc BWT vs BWT BWE}

# #Run DESeq
# ballast_deseq <- DESeq(ballast_deseq, test="Wald")
# 
# #Show results
# res_ballast_deseq <- results(ballast_deseq)
# res_ballast_deseq
# 
# #Check which comparisons are contained in the deseq object:
# resultsNames(ballast_deseq)
# 
# #Just want to compare Port_uptake with BWT; creating a results table with only this information
# res_ballast_deseq_BWT <- results(ballast_deseq, contrast=c("Sample_type","BWT","Port_uptake"), alpha=0.05)
# 
# #Show results summary
# summary(res_ballast_deseq_BWT)
# 
# #Log fold change shrinkage on the effluent_TWW_vs_CON comparison
# resLFC <- lfcShrink(ballast_deseq, coef="Sample_type_BWT_vs_Port_uptake", type="apeglm")
# resLFC
# 
# #"It is more useful visualize the MA-plot for the shrunken log2 fold changes, 
# #which remove the noise associated with log2 fold changes from low count genes 
# #without requiring arbitrary filtering thresholds."
# #https://www.bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#why-un-normalized-counts
# plotMA(resLFC)
# 
# #Convert results to dataframe
# df_ballast_deseq_BWT <- as.data.frame(res_ballast_deseq_BWT)
# 
# #Filter based on adj p-value, removing non-significant values (p>0.05). This also removes p-values that are NA; NA p-values exist because when p-values do not pass the filter threshold (p-values less than 0.1) they are set to NA. More information in vignette("DESeq2") under "Note on p-values set to NA"
# df_ballast_deseq_BWT_filt <- filter(df_ballast_deseq_BWT, padj <= 0.05)
# 
# #Order by adjusted p-value
# df_ballast_deseq_BWT_filt = df_ballast_deseq_BWT_filt[order(df_ballast_deseq_BWT_filt$padj, na.last=NA), ]
# 
# #Add pathway descriptions to the DESeq results using the left_join function in dplyr; rownames in the DESeq output need to be a column labeled pathway in order to match that of the descriptions dataframe
# df_ballast_deseq_BWT_filt <- rownames_to_column(df_ballast_deseq_BWT_filt, var = "pathway")
# 
# df_ballast_deseq_BWT_filt_descrip <- df_ballast_deseq_BWT_filt %>%
#   left_join(ballast_metacyc_all_levels, by = "pathway")
```

# DESeq2 analysis on KEGG data (TEST)

## Load ggpicrust2 and dependent packages

```{r picrust packages, warning=FALSE, error=FALSE, message=FALSE}

library("ggpicrust2")
library("phyloseq")
library("ALDEx2")
library("SummarizedExperiment")
library("Biobase")
library("devtools")
library("ComplexHeatmap")
library("BiocGenerics")
library("BiocManager")
library("metagenomeSeq")
library("Maaslin2")
library("edgeR")
library("lefser")
library("limma")
library("KEGGREST")
library("DESeq2")
library("aplot")
library("dplyr")
library("ggplot2")
library("grid")
library("MicrobiomeStat")
library("readr")
library("stats")
library("tibble")
library("tidyr")
library("ggprism")
library("cowplot")
library("ggforce")
library("ggplotify")
library("magrittr")
library("utils")


```

## Convert KO abundance to KEGG pathway abundance

```{r ko to kegg conversion}

# #Load packages
# library(ggpicrust2) 
# library(readr) 
# 
# #Run ko2kegg_abundance function on input file  
# kegg_abundance <- ko2kegg_abundance("PICRUSt2-analysis/KO_version/pred_metagenome_unstrat.tsv") 
# 
# #Read in metadata  
# metadata <- read_delim("PICRUSt2-analysis/KO_version/Metadata.tsv", delim = "\t")  
# 
# #Run differential abundance analysis in ggpicrust2  
# daa_results_df <- pathway_daa(abundance = kegg_abundance, metadata = metadata,                                group = "Sample_type", daa_method = "DESeq2") 
# 
# #Filter results to include port uptake vs BWT only  
# daa_results_df_BWT <- daa_results_df %>%   
#   filter(group1 == "port_uptake") 
# 
# daa_results_df_BWT <- daa_results_df_BWT %>%     
#   filter(group2 == "BWT") 
# 
# #Filter based on adj p-value, removing non-significant values (p>0.05). This also removes p-values that are NA; NA p-values exist because when p-values do not pass the filter threshold (p-values less than 0.1) they are set to NA. More information in vignette("DESeq2") under "Note on p-values set to NA"  
# daa_results_df_BWT_filt_final <- filter(daa_results_df_BWT, p_adjust <= 0.05)  
# 
# #Break the results into chunks that can be easily processed by the pathway_annotation function; only a few querys can be run at a time 
# daa_results_df_BWT_filt_final_A <- daa_results_df_BWT_filt_final %>%                                            slice_head(n=50) 
# 
# #Annotate as kegg pathways # 
# daa_annotated_results_df_BWT <- pathway_annotation(pathway = "KO",
#                                 daa_results_df = daa_results_df_BWT_filt_final_A,
#                                 ko_to_kegg = FALSE,
#                                 file = NULL) 
# 
# #Can delete this later - quick look at what these KOs are 
# #Read in the function descriptions  
# ballast_KO_descrip <- read_delim("PICRUSt2-analysis/KO_version/pred_metagenome_unstrat_descrip.tsv", delim = "\t") 
# 
# result1 <- pathway_annotation(file = "PICRUSt2-analysis/KO_version/pred_metagenome_unstrat_descrip.tsv", pathway = "KO", daa_results_df = NULL,  ko_to_kegg = FALSE)
```

# MBPD

## Import pathogen data

```{r pathogen, warning = FALSE, messages = FALSE, error = FALSE}

#Create a phyloseq object from the .qza files exported from qiime2 using 
#the qiime2R package
Ballast_pathogen_physeq <- qza_to_phyloseq(
  features="Qiime output/MBPD/ballast-dada2-table.qza",
  tree="Qiime output/MBPD/ballast-rooted-tree.qza",
  taxonomy="Qiime output/MBPD/ballast-pathogen-taxonomy.qza",
  metadata = "Qiime output/Metadata.tsv")
```

## Filter samples

Based on the table above, there are some samples with very few to no reads, so we'll filter these out here. In this case, we keep all samples with more than 1,000 reads.

Removing these samples with low sequencing depths is important because samples with low reads typically are lower quality and have a greater probability of containing contaminant sequences ([Weiss et al., 2017](https://microbiomejournal.biomedcentral.com/articles/10.1186/s40168-017-0237-y)).

```{r filter pathogen}

#Remove samples with few reads
Ballast_pathogen_physeq = subset_samples(Ballast_pathogen_physeq, 
                                         Sample_number != "104")
Ballast_pathogen_physeq = subset_samples(Ballast_pathogen_physeq, 
                            Sample_number != "105")
Ballast_pathogen_physeq = subset_samples(Ballast_pathogen_physeq, 
                            Sample_number != "107")
Ballast_pathogen_physeq = subset_samples(Ballast_pathogen_physeq, 
                            Sample_number != "108")
Ballast_pathogen_physeq = subset_samples(Ballast_pathogen_physeq, 
                            Sample_number != "109")
Ballast_pathogen_physeq = subset_samples(Ballast_pathogen_physeq, 
                            Sample_number != "106")

#Filter out low abundance ASVs
Ballast_pathogen_physeq = prune_taxa(taxa_sums(Ballast_pathogen_physeq) > 2, Ballast_pathogen_physeq)
```

## Label and order variables

Now we can relabel the variables; if you type in Ballast_physeq you will see that sample_data is the matrix that holds the information that we want to change, so we need to include sample_data in our code here.

```{r label and order pathogen}

#Order factors
sample_data(Ballast_pathogen_physeq)$Sample_type <- factor(sample_data(Ballast_pathogen_physeq)$Sample_type, 
               levels = c("port_uptake", "BWT", "ocean_uptake", "BWT_BWE"),
               labels = c("Port uptake", "BWT", "Ocean uptake", "BWT+BWE"))

sample_data(Ballast_pathogen_physeq)$Sample_number <- factor(sample_data(Ballast_pathogen_physeq)$Sample_number, 
               levels = c("80", "81", "82", "83", "84", "85", "86", "87", "88", "89", "90", "91", "92", "93", "94", "95", "96", "97", "98", "100", "101", "102", "103", "110", "111", "112", "113", "114", "115"),
               labels = c("80", "81", "82", "83", "84", "85", "86", "87", "88", "89", "90", "91", "92", "93", "94", "95", "96", "97", "98", "100", "101", "102", "103", "110", "111", "112", "113", "114", "115"))

sample_data(Ballast_pathogen_physeq)$Voyage <- factor(sample_data(Ballast_pathogen_physeq)$Voyage, 
               levels = c("1", "2"),
               labels = c("Voyage 1", "Voyage 2"))
```

## Bacterial community composition bar charts

### Species level, grouped by voyage and sample type

```{r pathogen species}

#Examining BCC by sample type & voyage
BallastSeqR_pathogen_species_voyage <- Ballast_pathogen_physeq %>%
  tax_glom(taxrank = "Species") %>%                   #Agglomerate at species level
  transform_sample_counts(function(x) {x/sum(x)}) %>% #Transform to rel. abundance
  psmelt() %>%                                        #Melt to long format
  group_by(Sample_type, Voyage, 
           Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
  dplyr::summarize(Mean = 
                     mean(Abundance, na.rm=TRUE)) %>% #Calculate average
  arrange(Class)
```

#### Plot with ggnested

```{r pathogen species plotted}

require("ggnested")

#Convert to a dataframe
BallastSeqR_pathogen_species_voyage_filt <- data.frame(BallastSeqR_pathogen_species_voyage) 

#Custom color palette
pal <- c("#440154FF", "blue", "green", "yellow", "purple", "orange", "#FF6699", "black", "red", "yellow")
names(pal) <- unique(BallastSeqR_pathogen_species_voyage_filt$Class)

#Use ggnested to create the bar plot
BallastSeqR_pathogen_voyage_species_bar_ggnested <- ggnested::ggnested(BallastSeqR_pathogen_species_voyage_filt, 
         aes(x = Sample_type, 
             y = Mean, 
             main_group = Class, 
             sub_group = Species),
          main_palette = pal) + 
  geom_bar(stat="identity", colour = "black", size=0.3, position = "fill") +
  facet_wrap(~ Voyage, scale="free") +
  guides(fill = guide_legend(reverse = FALSE, keywidth = 1, keyheight = 1)) +
  ylab("Relative Abundance (Species > 1%) \n") + xlab("Sample Type") +
  theme_minimal()+theme(
    strip.text = element_text(size=12),
    panel.border = element_rect(fill=NA, colour = "black"), 
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(), 
    axis.title.x = element_text(size=12),
    axis.title.y = element_text(size=12),
    axis.text.x  = element_text(angle = 90, vjust = 0.5, hjust=1, size=10, 
                              colour="black"), 
    axis.text.y = element_text(size=10, colour="black"), 
    plot.title = element_text(hjust = 0.5), 
    axis.ticks.x = element_line(colour="#000000", size=0.1), 
    axis.ticks.y = element_line(colour="#000000", size=0.1),
    legend.text = element_text(size = 10)) +
  facet_wrap(~ Voyage, scale="free")

print(BallastSeqR_pathogen_voyage_species_bar_ggnested)
```

# Filter ASVs to determine how many were given species-level assignments

```{r create new phyloseq object}

#Create a phyloseq object from the .qza files exported from qiime2 using 
#the qiime2R package
Ballast_physeq2 <- qza_to_phyloseq(
  features="Qiime output/ballast-dada2-table.qza",
  tree="Qiime output/ballast-rooted-tree.qza",
  taxonomy="Qiime output/ballast-taxonomy.qza",
  metadata = "Qiime output/Metadata.tsv"
)

#Check the rank names to make sure they are accurate
rank_names(Ballast_physeq2)

#Correct output:[1] "Kingdom" "Phylum"  "Class"   "Order"   "Family"  
#                   "Genus"   "Species"

#Check sample variables
sample_variables(Ballast_physeq2)

#Remove chloroplast sequences and any contaminant sequences
Ballast_physeq2 <- subset_taxa(Ballast_physeq2, Kingdom != "d__Archaea")
Ballast_physeq2 <- subset_taxa(Ballast_physeq2, Order != "Chloroplast")
Ballast_physeq2 <- subset_taxa(Ballast_physeq2, Family != "Mitochondria")

#Check that contaminant sequences are removed (easiest to save as data frame and search)
taxtabl <- as.data.frame(tax_table(Ballast_physeq2))

#At this point, blanks and positive controls should also be removed
Ballast_physeq2 = subset_samples(Ballast_physeq2, Sample_type != "control")
Ballast_physeq2 = subset_samples(Ballast_physeq2, Sample_type != "negative")

#Remove samples with few reads
Ballast_physeq2 = subset_samples(Ballast_physeq2, 
                            Sample_number != "104")
Ballast_physeq2 = subset_samples(Ballast_physeq2, 
                            Sample_number != "105")
Ballast_physeq2 = subset_samples(Ballast_physeq2, 
                            Sample_number != "107")
Ballast_physeq2 = subset_samples(Ballast_physeq2, 
                            Sample_number != "108")
Ballast_physeq2 = subset_samples(Ballast_physeq2, 
                            Sample_number != "109")
Ballast_physeq2 = subset_samples(Ballast_physeq2, 
                            Sample_number != "106")

#Filter out low abundance ASVs
Ballast_physeq2 = prune_taxa(taxa_sums(Ballast_physeq2) > 2, Ballast_physeq2)

#Order factors
sample_data(Ballast_physeq2)$Sample_type <- factor(sample_data(Ballast_physeq2)$Sample_type, 
               levels = c("port_uptake", "BWT", "ocean_uptake", "BWT_BWE"),
               labels = c("Port uptake", "BWT", "Ocean uptake", "BWT+BWE"))

sample_data(Ballast_physeq2)$Sample_number <- factor(sample_data(Ballast_physeq2)$Sample_number, 
               levels = c("80", "81", "82", "83", "84", "85", "86", "87", "88", "89", "90", "91", "92", "93", "94", "95", "96", "97", "98", "100", "101", "102", "103", "110", "111", "112", "113", "114", "115"),
               labels = c("80", "81", "82", "83", "84", "85", "86", "87", "88", "89", "90", "91", "92", "93", "94", "95", "96", "97", "98", "100", "101", "102", "103", "110", "111", "112", "113", "114", "115"))

sample_data(Ballast_physeq2)$Voyage <- factor(sample_data(Ballast_physeq2)$Voyage, 
               levels = c("1", "2"),
               labels = c("Voyage 1", "Voyage 2"))
```

```{r ASVs assigned to species}

#Use tax_mutate to edit the species level names
tax_table(Ballast_physeq2) <- tax_table(Ballast_physeq2) %>%
  tax_mutate(Species=gsub("_", " ", Species))

taxtab2 <- as.data.frame(tax_table(Ballast_physeq2))

#Create new dataframe that excludes any NA's in the species column
taxtabfilt <- taxtab2 %>% filter(!is.na(Species))

#Remove unassigned, uncultured and similar
taxtabfilt <- dplyr::filter(taxtabfilt, !grepl('unidentified|uncultured', Species))
taxtabfilt <- dplyr::filter(taxtabfilt, !grepl('metagenome|sp.', Species))

#Remove non-species 
taxtabfilt <- taxtabfilt %>%
  filter(!Species %in% 
           c("Antarctic bacterium", "bacterium enrichment", "microbial mat",
             "Rhodobacteraceae bacterium", "alpha proteobacterium", 
             "Actinobacteria bacterium", "Candidatus Planktophila")) 
```

```{r ASVs assigned to genus}

#Use tax_mutate to edit the species level names
tax_table(Ballast_physeq2) <- tax_table(Ballast_physeq2) %>%
  tax_mutate(Genus=gsub("_", " ", Genus))

taxtab3 <- as.data.frame(tax_table(Ballast_physeq2))

#Create new dataframe that excludes any NA's in the species column
taxtabfilt_genus <- taxtab3 %>% filter(!is.na(Genus))

#Remove unassigned, uncultured and similar
taxtabfilt_genus <- dplyr::filter(taxtabfilt_genus, 
                                  !grepl('unidentified|uncultured', Genus))
taxtabfilt_genus <- dplyr::filter(taxtabfilt_genus, 
                                  !grepl('clade|group', Genus))

#Remove non-genera 
taxtabfilt_genus <- taxtabfilt_genus %>%
  filter(!Genus %in% 
           c("env.OPS 17", "AKYH767",
             "Bacteroidetes vadinHA17", "Clade IV", "TRA3-20", "R7C24", 
             "JGI 0000069-P22", "DEV007", "UBA12409", "	67-14", "PeM15",
             "IMCC26256", "KD4-96", "S25-593", "T34", "Clade Ia", "Clade II", 
             "67-14", "C39", "Clade III", "GWD2-49-16", "MWH-Ta3", "OLB12", 
             "BSV13", "SH3-11", "SM1A02", "SH-PL14", "HTCC5015", "BIyi10",
             "Clostridium sensu stricto 1", "SUP05 cluster")) 
```

# Session Info

```{r info}

devtools::session_info()
```
